---
title: "Hands-on 1B | Thematic Mapping and GeoVisualisation with R"
subtitle: "In this exercise, we will learn to create thematic maps and perform geovisualization in R using the tmap package, including data preparation, classification, color schemes, and advanced mapping techniques."
date: "Aug 29, 2025"
date-modified: "last-modified" # auto render date
author: Linh Nguyen
execute: # how YML page work
  echo: true # will show code
  eval: true # will run code every time and show output
  freeze: auto # controls rendering - if false then no render if no changes
  message: false
  warning: false # dont show warning msg
format:
  html:
    code-link: true
    toc: true
number-sections: true
number-offset: 1
editor: visual
---

## Overview

In general, thematic mapping involves the use of map symbols to visualize selected properties of geographic features that are not naturally visible, such as population, temperature, crime rate, and property prices.

Geovisualisation, on the other hand, is the process of using visual representations and cartographic techniques to explore, analyze, and communicate geospatial data. It combines elements of cartography, computer science, and information visualization to enhance spatial understanding and knowledge discovery.

## The Data

The following data sources will be used in this exercise:

| Dataset | Source | Description | Format |
|-----------------|-----------------|---------------------|-----------------|
| **Master Plan 2019 Subzone Boundary (Web) (**`MasterPlan2019SubzoneBoundaryNoSeaKML.kml`**)** | [data.gov.sg](https://www.data.gov.sg/) | Geographical boundary of Singapore at the planning subzone level. The data is based on URA Master Plan 2019. | KML |
| **Singapore Residents by Planning Area / Subzone, Age Group, Sex and Type of Dwelling, June 2024 (**`respopagesex2024.csv`**)** | [Department of Statistics, Singapore](https://www.singstat.gov.sg/) | Aspatial data containingPA and SZ fields, which can be used as unique identifiers to georeference to Master Plan 2019 Subzone Boundary | CSV |

## Installing and Loading the R Packages

The following R packages will be used in this exercise:

| Package | Purpose | Use Case in Exercise |
|------------------|-----------------------------|-------------------------|
| **tmap** | Thematic mapping for visualizing geospatial data. | Creating and customizing thematic maps. |
| **sf** | Handling geospatial data in simple features format. | Importing, managing, and processing geospatial data. |
| **readr** | Importing delimited text files (part of `tidyverse`). | Loading CSV data files. |
| **tidyr** | Tidying data into a usable format (part of `tidyverse`). | Reshaping and organizing data. |
| **dplyr** | Data wrangling and transformation tasks (part of `tidyverse`). | Manipulating and transforming datasets. |

To install and load these packages in R, use the following code:

```{r}
pacman::p_load(sf, tmap, tidyverse, rvest)
```

## Import Geospatial Data into R

The code block below uses st_read() function of sf package to import the `MP14_SUBZONE_WEB_PL` shapefile into R as a polygon feature data frame.

```{r}
mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
```

### Tidying data

Function to extract values from the HTML description

```{r}
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}
```

```{r}
mpsz <- mpsz %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())
```

Show the first 10 features:

```{r}
mpsz
```

### Importing Attribute Data into R

Next, *respopagsex2024.csv* file is imported into RStudio and save the file into an R dataframe called *popdata*.

The task will be performed by using *read_csv()* function of **readr** package as shown in the code block below.

```{r, echo=TRUE, message=FALSE}
popdata <- read_csv("data/aspatial/respopagesextod2024.csv")
```

### Data Preparation

Before a thematic map can be prepared, you are required to prepare a data table with year 2020 values. The data table should include the variables PA, SZ, YOUNG, ECONOMY ACTIVE, AGED, TOTAL, DEPENDENCY.

YOUNG: age group 0 to 4 until age groyup 20 to 24, ECONOMY ACTIVE: age group 25-29 until age group 60-64, AGED: age group 65 and above, TOTAL: all age group, and DEPENDENCY: the ratio between young and aged against economy active group

#### Data wrangling

The following data wrangling and transformation functions will be used:

-   *pivot_wider()* of **tidyr** package, and
-   *mutate()*, *filter()*, *group_by()* and *select()* of **dplyr** package

```{r}
popdata2024 <- popdata %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup()%>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

#### Joining the attribute data and geospatial data

Before we can perform the georelational join, one extra step is required to convert the values in PA and SZ fields to uppercase. This is because the values of PA and SZ fields are made up of upper- and lowercase. On the other, hand the SUBZONE_N and PLN_AREA_N are in uppercase.

```{r}
popdata2024 <- popdata2024 %>%
  mutate_at(.vars = vars(PA, SZ), 
          .funs = list(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

Next, the left_join() function from the dplyr package is used to merge the geographic data and attribute table, using the planning subzone names (e.g. SUBZONE_N and SZ) as the common identifier.

```{r}
mpsz_pop2024 <- left_join(mpsz, popdata2024,
                          by = c("SUBZONE_N" = "SZ"))
```

::: callout-note
**Key Takeaway:**

The left_join() function from the dplyr package is applied to ensure that the resulting output remains a simple features data frame, with mpsz as the left table.
:::

```{r}
dir.create("data/rds/", recursive = TRUE)
write_rds(mpsz_pop2024, "data/rds/mpszpop2024.rds")
```

## Choropleth Mapping Geospatial Data Using *tmap*

Choropleth mapping involves the symbolisation of enumeration units, such as countries, provinces, states, counties or census units, using area patterns or graduated colors. For example, a social scientist may need to use a choropleth map to portray the spatial distribution of aged population of Singapore by Master Plan 2014 Subzone Boundary.

There are two methods for creating thematic maps using tmap:

-   Using `qtm()` for quick thematic map plotting.
-   Using individual tmap elements for creating highly customizable maps.

### Plotting a choropleth map quickly by using *qtm()*

The easiest and quickest to draw a choropleth map using **tmap** is using *qtm()*. It is concise and provides a good default visualisation in many cases.

The following code draws a standard choropleth map:

```{r}
tmap_mode("plot")
qtm(mpsz_pop2024,
    fill = "DEPENDENCY")
```

::: callout-note
**Key Takeaway:**

-   tmap_mode("plot") is used to generate a static map. To enable interactive mode, use the "view" option.
-   The fill argument is used to map the attribute to be visualized (in this case, DEPENDENCY).
:::

### Creating a choropleth map by using *tmap*'s elements

While `qtm()` is useful for quickly and easily drawing choropleth maps, its drawback is the limited control over the aesthetics of individual layers. To create a high-quality, cartographic choropleth map like the one shown below, it's recommended to use the individual elements provided by *tmap*.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

In the following sub-section, more *tmap* functions will be used to plot these elements.

#### Drawing a base map

The basic building block of tmap is tm_shape() followed by one or more layer elemments such as **tm_polygons()**, **tm_symbols()**, **tm_lines()**, **tm_raster()** and **tm_text()**.

In the code block below, *tm_shape()* is used to define the input data (i.e *mpsz_pop2024*) and *tm_polygons()* is used to draw the planning subzone polygons

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons()
```

#### Drawing a choropleth map using *tm_polygons()*

To draw a choropleth map showing the geographical distribution of a selected variable by planning subzone, we just need to assign the target variable such as *Dependency* to *tm_polygons()*.

```{r}
tm_shape(mpsz_pop2024)+
  tm_polygons("DEPENDENCY")
```

::: callout-note
Key takeaways from *tm_polygons()*:

-   The default method for interval binning when creating a choropleth map is "pretty." A detailed explanation of the data classification methods available in **tmap** will be covered in sub-section 4.3.
-   The default color palette used is `YlOrRd` from ColorBrewer. More information on color schemes will be explored in sub-section 4.4.
-   Missing values are automatically represented with a grey shade by default.
:::

#### Drawing a choropleth map using *tm_fill()* and *tm_border()*

Actually, *tm_polygons()* is a wraper of *tm_fill()* and *tm_border()*. *tm_fill()* shades the polygons by using the default colour scheme and *tm_borders()* adds the borders of the shapefile onto the choropleth map.

The code block below draws a choropleth map by using *tm_fill()* alone.

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY")
```

Notice that the planning subzones are shared according to the respective dependecy values

To add the boundary of the planning subzones, tm_borders will be used as shown in the code block below.

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY") +
  tm_borders(lwd = 0.1,  alpha = 1)
```

Note that light-gray border lines have been added on the choropleth map.

The *fill_alpha* option specifies a transparency value ranging from 0 (completely transparent) to 1 (not transparent). By default, the col's alpha value is used (usually 1).

Beside *fill_alpha* argument, there are three other arguments for *tm_borders()*, they are:

-   *col* = border colour,
-   *lwd* = border line width. The default is 1, and
-   *lty* = border line type. The default is "solid".

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY") +
  tm_borders(col = "grey60",
             lwd = 0.1,
             lty = "dashed")
```

### Data classification methods of **tmap**

Most choropleth maps employ some methods of data classification. The point of classification is to take a large number of observations and group them into data ranges or classes.

**tmap** provides a total ten data classification methods, namely: *fixed*, *sd*, *equal*, *pretty* (default), *quantile*, *kmeans*, *hclust*, *bclust*, *fisher*, and *jenks*.

To define a data classification method, the *style* argument of *tm_fill()* or *tm_polygons()* will be used.

#### Plotting choropleth maps with built-in classification methods

The code block below shows a quantile data classification that used 5 classes.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

In the code block below, *equal* data classification method is used.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

::: callout-note
**Observation:** The distribution of quantile data classification method are more evenly distributed then equal data classification method.
:::

::: callout-warning
**Warning: Maps Lie!**
:::

**Exercise: Using what you had learned, prepare choropleth maps by using different classification methods supported by tmap and compare their differences.**

The tabs below shows the choropleth maps drawn using different classifications methods with the same number of classes (n=5).

::: panel-tabset
#### quantile

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5)
```

#### equal

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5)
```

#### pretty

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "pretty") +
  tm_borders(alpha = 0.5)
```

#### jenks

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```

#### sd

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "sd") +
  tm_borders(alpha = 0.5)
```

#### hclust

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "hclust") +
  tm_borders(alpha = 0.5)
```

#### fisher

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "fisher") +
  tm_borders(alpha = 0.5)
```

#### bclust

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "bclust") +
  tm_borders(alpha = 0.5)
```

#### kmeans

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "kmeans") +
  tm_borders(alpha = 0.5)
```
:::

::: callout-note
**Observation:** Different data classification methods produce varying color patterns. Methods like *pretty*, *equal*, and *sd* are more sensitive to outliers, often resulting in fewer distinct areas on the map. In contrast, methods like *quantile*, *jenks*, and *kmeans* create more evenly distributed colors. Therefore, choosing the right classification method based on data distribution is crucial to avoid misleading representations in choropleth maps.
:::

**Exercise: Preparing choropleth maps by using similar classification method but with different numbers of classes (i.e. 2, 6, 10, 20). Compare the output maps, what observation can you draw?**

The tabs below show choropleth maps with `kmeans` classification method. The numbers of classes used are (2, 6, 10, 20).

::: panel-tabset
#### n=2

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY",
          n = 2,
          style = "kmeans") +
  tm_borders(fill_alpha = 0.5)
```

#### n=6

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY",
          n = 6,
          style = "kmeans") +
  tm_borders(fill_alpha = 0.5)
```

#### n=10

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY",
          n = 10,
          style = "kmeans") +
  tm_borders(fill_alpha = 0.5)
```

#### n=20

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY",
          n = 20,
          style = "kmeans") +
  tm_borders(fill_alpha = 0.5)
```
:::

::: callout-note
**Observation:** Using different class numbers within the same classification method affects the map's coloring. With n=2, outliers dominate, resulting in only one distinct area. Increasing the class count distributes the data more evenly across the map.
:::

#### Plotting choropleth map with custom break

For all the built-in styles, the category breaks are computed internally. In order to override these defaults, the breakpoints can be set explicitly by means of the *breaks* argument to the *tm_fill()*. It is important to note that, in **tmap** the breaks include a minimum and maximum. As a result, in order to end up with n categories, n+1 elements must be specified in the *breaks* option (the values must be in increasing order).

Before we get started, it is always a good practice to get some descriptive statistics on the variable before setting the break points. Code block below will be used to compute and display the descriptive statistics of **DEPENDENCY** field.

```{r}
summary(mpsz_pop2024$DEPENDENCY)
```

Based on the results, we choose breakpoints at 0.60, 0.70, 0.80, and 0.90, with the minimum at 0 and maximum at 1. Our breaks vector is c(0, 0.60, 0.70, 0.80, 0.90, 1.00).

Now, we will plot the choropleth map with the following code:

```{r}
tm_shape(mpsz_pop2024)+
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00))) +
  tm_borders(fill_alpha = 0.5)
```

### Colour Scheme

**tmap** supports colour ramps either user-defined or a set of predefined colour ramps from the **RColorBrewer** package.

#### Using ColourBrewer palette

To apply a specific color palette, assign it to the palette argument in `tm_fill()` as shown below:

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

Notice that the choropleth map is shaded in green.

To reverse the colour shading, add a "-" prefix.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

Notice that the colour scheme has been reversed.

#### Cartographic Furniture

Beside map style, **tmap** also also provides arguments to draw other map furniture such as compass, scale bar and grid lines. In the following example, *tm_compass()*, *tm_scale_bar()* and *tm_grid()* are used to add compass, scale bar and grid lines onto the choropleth map.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = c("left", "bottom"))
```

### Map Layouts

Map layout refers to the combination of all map elements into a cohesive map. Map elements include among others the objects to be mapped, the title, the scale bar, the compass, margins and aspects ratios. Colour settings and data classification methods covered in the previous section relate to the palette and break-points are used to affect how the map looks.

#### Map Legend

In **tmap**, there are several legend options for adjusting the placement, format, and style of the legend.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_pos_auto_in() +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = c("left", "bottom"))
```

#### Map style

**tmap** allows extensive customization of layout settings, accessible through the `tmap_style()` function. The code below demonstrates the use of the *classic* style.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "-brewer.greens")) + 
  tm_borders(fill_alpha = 0.5) + 
  tmap_style("classic")
```

To reset the default style, refer to the code chunk below.

```{r}
tmap_style("white")
```

### Drawing Small Multiple Choropleth Maps

**Small multiple maps**, also referred to as **facet maps**, are composed of many maps arrange side-by-side, and sometimes stacked vertically. Small multiple maps enable the visualisation of how spatial relationships change with respect to another variable, such as time.

In **tmap**, small multiple maps can be created through three methods:

-   Assigning multiple values to an aesthetic argument.
-   Defining a grouping variable with `tm_facets()`.
-   Creating multiple stand-alone maps with `tmap_arrange()`.

#### By assigning multiple Values to at least one aesthetic argument

In this example, small multiple choropleth maps are created by defining ***ncols*** in **tm_fill()**

```{r}
tm_shape(mpsz_pop2024) + 
  tm_polygons(
    fill = c("YOUNG", "AGED"),
    fill.legend = 
      tm_legend(position = tm_pos_in(
        "right", "bottom")),
    fill.scale = tm_scale_intervals(
      style = "equal", 
      n = 5,
      values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tmap_style("classic")
```

#### By arrange multiples choropleth maps in a grid layout

In this example, multiple choropleth maps are created and **tmap_arrnage()** is used to arrnage them in a grid layout.

```{r}
youngmap <- tm_shape(mpsz_pop2024)+ 
  tm_polygons(fill = "YOUNG",
              fill.legend = tm_legend(
                position = tm_pos_in(
                  "right", "bottom"),
                  item.height = 0.8),
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of young population")
                
agedmap <- tm_shape(mpsz_pop2024)+ 
  tm_polygons(fill = "AGED",
              fill.legend = tm_legend(
                position = tm_pos_in(
                  "right", "bottom"),
                item.height = 0.8),
              fill.scale = tm_scale_intervals(
              style = "quantile", 
              values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of aged population")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

#### By defining a group-by variable in *tm_facets()*

This example generates multiple small maps by using `tm_facets()`.

```{r}
tm_shape(mpsz_pop2024) +
  tm_fill(fill = "DEPENDENCY",
          fill.scale = tm_scale_intervals(
            style = "quantile",
            values = "brewer.blues")) + 
  tm_facets(by = "REGION_N",
            nrow = 2, 
            ncols = 3,
            free.coords=TRUE, 
            drop.units=TRUE) +
  tm_layout(legend.show = TRUE,
            title.position = c("center", "center"), 
            title.size = 20) +
  tm_borders(fill_alpha = 0.5)
```

## Mappping Spatial Object Meeting a Selection Criterion

Instead of creating small multiple choropleth map, you can also use **filter()** of **dplyr** package to select geographical area of interest and plot a choropleth map focus only on the selected region.

```{r}
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                values = "brewer.greens"),
              fill.legend = tm_legend()) +
  tm_borders(fill_alpha = 0.5)
```

## Complementing Thematic Map with Statistical Chart

Maps and statistical charts complement each other by visually representing different aspects of the same data, offering a more comprehensive understanding. Maps excel at showing spatial relationships and geographical patterns, while charts effectively display numerical data, trends, and comparisons. Combining both allows for a more insightful and engaging data narrative, especially when dealing with spatial data that also has quantifiable characteristics.

With tmap, statistical chart and be incorporate into the map visualisation by using **fill.chat** argument of map layers and legend chart feature as shown in the code chunk below.

```{r}
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                values = "brewer.greens"),
              fill.legend = tm_legend(),
              fill.chart = tm_chart_box()) +
  tm_borders() +
  tm_layout(asp = 0.8)
```

In the code chunk below, We improve the visual representation further by highlighting and lebaling the outliers on the choropleth map.

```{r}
mpsz_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")

stats <- boxplot.stats(mpsz_selected$DEPENDENCY)

outlier_vals <- stats$out

outlier_sf <- mpsz_selected[mpsz_selected$DEPENDENCY %in% outlier_vals, ]

tm_shape(mpsz_selected) +
  tm_polygons(fill = "DEPENDENCY",
          fill.scale = tm_scale_intervals(
            style = "quantile", 
            values = "brewer.blues"),
          fill.legend = tm_legend(),
          fill.chart = tm_chart_box()) +
  tm_borders(fill_alpha = 0.5) +
tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_text("SUBZONE_N", col = "red", size = 0.7) +
  tm_layout(asp = 0.8)
```

## Creating Interactive Map

Interactive maps let users actively explore and interact with the data they display. Unlike static maps, you can zoom in and out, pan across areas, click on locations for more information, and even work with data overlays or visualizations—making the experience more dynamic and informative. One of the great things about tmap is that it lets you switch easily between static and interactive maps using **tmap_mode()**, so you can choose the view that best suits your analysis.

By modifying the code chunk in sub-section 2.6, the code chunks below build an interactive map using

```{r}
region_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")
region_bbox <- st_bbox(region_selected)

stats <- boxplot.stats(region_selected$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- region_selected[region_selected$DEPENDENCY %in% outlier_vals, ]

tmap_mode("view")
tm_shape(region_selected, 
         bbox = region_bbox) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          popup.vars = c(
            "Name" = "SUBZONE_N", 
            "Dependency" = "DEPENDENCY")) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2)
```

```{r}
tmap_mode("plot")
```

The interactive map above is far from satisfactory. While we want to encourage users to engage and explore the interactive by zooming in and out of the study area freely. But, users might lost in the cyberspace with too much freedom to zoom-in and zoom-out.

To address this issue, **set_zoom_limits** argument can be used to limit the map extend users can zooming in and out of the map areas as shown below.

```{r}
region_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")
region_bbox <- st_bbox(region_selected)

stats <- boxplot.stats(region_selected$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- region_selected[region_selected$DEPENDENCY %in% outlier_vals, ]

tmap_mode("view")

tm_shape(region_selected, 
         bbox = region_bbox) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          popup.vars = c(
            "Name" = "SUBZONE_N", 
            "Dependency" = "DEPENDENCY")) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_view(set_zoom_limits = c(12,14))
```

```{r}
tmap_mode("plot")
```
## References

### All about **tmap** package

-   [tmap: Thematic Maps in R](https://www.jstatsoft.org/article/view/v084i06)
-   [tmap](https://cran.r-project.org/web/packages/tmap/index.html)
-   [tmap: get started!](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html)
-   [tmap: changes in version 2.0](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-changes-v2.html)
-   [tmap: creating thematic maps in a flexible way (useR!2015)](http://von-tijn.nl/tijn/research/presentations/tmap_user2015.pdf)
-   [Exploring and presenting maps with tmap (useR!2017)](http://von-tijn.nl/tijn/research/presentations/tmap_user2017.pdf)

### Geospatial Data Wrangling

-   [sf: Simple Features for R](https://cran.r-project.org/web/packages/sf/index.html)
-   [Simple Features for R: StandardizedSupport for Spatial Vector Data](https://journal.r-project.org/archive/2018/RJ-2018-009/RJ-2018-009.pdf)
-   [Reading, Writing and Converting Simple Features](https://cran.r-project.org/web/packages/sf/vignettes/sf2.html)

### Data Wrangling

-   [dplyr](https://dplyr.tidyverse.org/)
-   [Tidy data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)
-   [tidyr: Easily Tidy Data with 'spread()' and 'gather()' Functions](https://cran.r-project.org/web/packages/tidyr/tidyr.pdf)