---
title: "Hands-on Exercise 6 | Geographical Segmentation with Spatially Constrained Clustering Techniques"
subtitle: "In this hands-on exercise, you will gain hands-on experience on how to delineate homogeneous region by using geographically referenced multivariate data."
date: "Oct 3, 2025"
date-modified: "last-modified"
author: Linh Nguyen
execute:
  echo: true
  eval: true
  freeze: auto
  message: false
  warning: false
format:
  html:
    code-link: true
    toc: true
number-sections: true
number-offset: 1
editor: visual
---

## Overview

Homogeneous regions can be delineated using geographically referenced multivariate data.

There are two major analysis: - hierarchical cluster analysis - spatially constrained cluster analysis

### Analytical Question

We are looking to delineate **Shan State, Myanmmar**, into homogeneous regions by using multiple Information and Communication technology (ICT) measures, namely: Radio, Television, Land line phone, Mobile phone, Computer, and Internet at home.

### The Data

| Dataset | Description | Source | Format |
|-----------------|---------------------|------------------|-----------------|
| Myanmar Township Boundary Data | Township boundary information of Myanmar | [Myanmar Information Management Unit (MIMU)](https://myanmar.unfpa.org/en/publications/2014-population-and-housing-census-myanmar-data-sheet) | ESRI Shapefile |
| Shan-ICT | This is an extract of The 2014 Myanmar Population and Housing Census Myanmar at the township level | [The 2014 Myanmar Population and Housing Census Myanmar](https://myanmar.unfpa.org/en/publications/2014-population-and-housing-census-myanmar-data-sheet) | csv |

## Installing and loading R packages

We use the following packages - `sf`, `rgadal`, `spdep`: For spatial data handling - `tidyverse`: For attribute data handling - `tmap`: For choropleth mapping - `coorplot`, `ggpubr`, `heatmaply`: For multivariate data visualisation and analysis - `cluster`, `clustergeo`: For cluster analysis

```{r}
pacman::p_load(spdep, tmap, sf, ClustGeo, 
               ggpubr, cluster, factoextra, NbClust,
               heatmaply, corrplot, psych, tidyverse, GGally)
```

## Data Import and Preparation

### Importing geospatial data into R environment

The Myanmar Township Boundary GIS data is in ESRI shapefile format and imported using `st_read()` function.

```{r}
shan_sf <- st_read(dsn = "data/geospatial", 
                   layer = "myanmar_township_boundaries") %>%
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)")) %>%
  select(c(2:7))
```

The imported township boundary object is *shan_sf*, and saved in a **simple feature dataframe format**.

```{r}
shan_sf
```

We can use `glimpse()` to reveal the datatype of the fields.

```{r}
glimpse(shan_sf)
```

### Importing aspatial data into R environment

```{r}
ict <- read_csv ("data/aspatial/Shan-ICT.csv")
```

The imported InfoComm variables are extracted from **The 2014 Myanmar Population and Housing Census Myanmar**. The attribute data set is called ict, and is saved in R's tibble dataframe format.

```{r}
summary(ict)
```

There are a total 11 fields and 55 observations.

### Derive new variables using dplyr package

The unit of measurement is based on number of households, however this will be biased by the underlying total number of households.

In order to overcome this problem, we derive the penetration rate of each ICT variable instead.

```{r}
ict_derived <- ict %>%
  mutate(`RADIO_PR` = `Radio`/`Total households`*1000) %>%
  mutate(`TV_PR` = `Television`/`Total households`*1000) %>%
  mutate(`LLPHONE_PR` = `Land line phone`/`Total households`*1000) %>%
  mutate(`MPHONE_PR` = `Mobile phone`/`Total households`*1000) %>%
  mutate(`COMPUTER_PR` = `Computer`/`Total households`*1000) %>%
  mutate(`INTERNET_PR` = `Internet at home`/`Total households`*1000) %>%
  rename(`DT_PCODE` =`District Pcode`,`DT`=`District Name`,
         `TS_PCODE`=`Township Pcode`, `TS`=`Township Name`,
         `TT_HOUSEHOLDS`=`Total households`,
         `RADIO`=`Radio`, `TV`=`Television`, 
         `LLPHONE`=`Land line phone`, `MPHONE`=`Mobile phone`,
         `COMPUTER`=`Computer`, `INTERNET`=`Internet at home`) 
```

```{r}
summary(ict_derived)
```

## Exploratory Data Analysis (EDA)

### EDA using statistical graphics

We can plot the distribution of the variables using histogram to identify if it is left/right skewed or having a normal distribution.

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

Boxplot can be plotted to detect if there are outliers.

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

We will also plot the distribution of newly derived variables (e.g. Radio penetration rate).

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

The derived variables are plotted below:

```{r}
radio <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

tv <- ggplot(data=ict_derived, 
             aes(x= `TV_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

llphone <- ggplot(data=ict_derived, 
             aes(x= `LLPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

mphone <- ggplot(data=ict_derived, 
             aes(x= `MPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

computer <- ggplot(data=ict_derived, 
             aes(x= `COMPUTER_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

internet <- ggplot(data=ict_derived, 
             aes(x= `INTERNET_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

We use `ggarrange()` of **ggpubr** package to group these histograms together.

```{r}
ggarrange(radio, tv, llphone, mphone, computer, internet, 
          ncol = 3, 
          nrow = 2)
```

### EDA using choropleth map

#### Joining geospatial data with aspatial data

We combine both the geospatial data object (shan_sf) with the aspatial dataframe (ict_derived) into one data frame.

The unique identifier used to join both data objects is *TS_PCODE*.

```{r}
shan_sf <- left_join(shan_sf, 
                     ict_derived, by=c("TS_PCODE"="TS_PCODE"))
  
write_rds(shan_sf, "data/rds/shan_sf.rds")
```

```{r}
shan_sf <- read_rds("data/rds/shan_sf.rds")
```

#### Preparing a choropleth map

Using a choropleth map, we see the distribution of Radio penetration rate of Shan State at a township level.

```{r}
qtm(shan_sf, "RADIO_PR")
```

We can see that the distribution of the number of households with radios and the number of households in the townships are similar.

```{r}
TT_HOUSEHOLDS.map <- tm_shape(shan_sf) + 
  tm_fill(col = "TT_HOUSEHOLDS",
          n = 5,
          style = "jenks", 
          title = "Total households") + 
  tm_borders(alpha = 0.5) 

RADIO.map <- tm_shape(shan_sf) + 
  tm_fill(col = "RADIO",
          n = 5,
          style = "jenks",
          title = "Number Radio ") + 
  tm_borders(alpha = 0.5) 

tmap_arrange(TT_HOUSEHOLDS.map, RADIO.map,
             asp=NA, ncol=2)
```

When we compare the total number of households to Radio peneration rate, we see that the population bias disappears.

```{r}
tm_shape(shan_sf) +
    tm_polygons(c("TT_HOUSEHOLDS", "RADIO_PR"),
                style="jenks") +
    tm_facets(sync = TRUE, ncol = 2) +
  tm_legend(legend.position = c("right", "bottom"))+
  tm_layout(outer.margins=0, asp=0)
```

## Correlation Analysis

Before performing cluster analysis, we need to ensure that the cluster variables are not highly correlated.

We use `corrplot.mixed()` function of the **corrplot** package to visualise and analyse the correlation of the input variables.

```{r}
cluster_vars.cor = cor(ict_derived[,12:17])
corrplot.mixed(cluster_vars.cor,
         lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

The correlation plot above shows that COMPUTER_PR and INTERNET_PR are highly correlated. This suggest that only one of them should be used in the cluster analysis instead of both.

## Hierarchy Cluster Analysis

### Extracting clustering variables

We exclude INTERNET_PR because it is highly correlated with variable COMPUTER_PR.

```{r}
cluster_vars <- shan_sf %>%
  st_set_geometry(NULL) %>%
  select("TS.x", "RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR")
head(cluster_vars,10)
```

We change the rows by township name instead of row number by using the code chunk below.

```{r}
row.names(cluster_vars) <- cluster_vars$"TS.x"
head(cluster_vars,10)
```

Now we delete the TS.x field.

```{r}
shan_ict <- select(cluster_vars, c(2:6))
head(shan_ict, 10)
write_rds(shan_ict, "data/rds/shan_ict.rds")
```

```{r}
shan_ict <- read_rds("data/rds/shan_ict.rds")
```

### Data standardisation

Multiple variables of varying value ranges will be used. In order to avoid cluster analysis result to be biased to clustering variables with large values, it is useful to standardise the input variables before performing clsuter analysis.

#### Mix-Max standardisation

`normalize()` of **heatmaply** package is used to standardise variables using the Min-Max method.

```{r}
shan_ict.std <- normalize(shan_ict)
summary(shan_ict.std)
```

Values are within the ranges of 0 to 1 now.

#### Z-score standardisation

`scale()` is used to perform Z-score standardisation.

```{r}
shan_ict.z <- scale(shan_ict)
describe(shan_ict.z)
```

The mean and standard deviation of the Z-score standardised clustering variables are 0 and 1 respectively.

Z-score standardisation method should only be used if we would assume all variables come from some normal distribution.

### Visualising the standardised clustering variables

We visualise the distribution of the different standardisation methods.

```{r}
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

RADIO_PR seems to follow a normal distribution as the distribution is the same in the non-standardised graph as compared to the Z-score standardisation.

```{r}
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

### Computing proximity matrix

We will compute the proximity matrix using `dist()`.

It supports 6 distance proximity calculations: - **euclidean** - **maximum** - **manhattan** - **canberra** - **binary** - **minkowski**

```{r}
proxmat <- dist(shan_ict, method = 'euclidean')
```

```{r}
proxmat
```

### Computing hierarchical clustering

We use `hclust()` to perform hierarchical clustering.

There are 8 clustering algorithms: - ward.D - ward.D2 - single - complete - average(UPGMA) - mcquitty(WPGMA) - median(WPGMC) - centroid(UPGMC)

```{r}
hclust_ward <- hclust(proxmat, method = 'ward.D')
plot(hclust_ward, cex = 0.6)
```

### Selecting the optimal clustering algorithm

We use `agnes()` function of the **cluster** package that functions like `hclus()` but you can also get the agglomerative coefficient, which measures the amount of clustering structure found (values closer to 1 suggest strong clustering structure).

```{r}
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

ac <- function(x) {
  agnes(shan_ict, method = x)$ac
}

map_dbl(m, ac)
```

### Determining optimal clusters

To determine the optimal number of clusters, we can use 3 commonly used methods: - Elbow method - Average Silhouette method - Gap statistic method

#### Gap statistic method

The gap statistic compares the total within intra-cluster variation for different values of k with their expected values under null rference distribution of the data. The estimate of the optimal clusters will be values that maximise the gap statistic.

To compute the gap statistic, `clusGap()` of **cluster** package will be used.

We also use the *hcut* function from the **factoextra** package.

```{r}
set.seed(12345)
gap_stat <- clusGap(shan_ict, 
                    FUN = hcut, 
                    nstart = 25, 
                    K.max = 10, 
                    B = 50)
# Print the result
print(gap_stat, method = "firstmax")
```

Next, we can visualise the plot by using `fviz_gap_stat()` of **factoextra** package.

```{r}
fviz_gap_stat(gap_stat)
```

From the gap statistic, the recommended number of clusters to retain is 1, but it is not logical to retain only 1 cluster. Hence, we use 6-cluster as it gives the largest gap statistic and is the next best cluster to pick.

### Interpreting the dendograms

The height of the fusion, on the vertical axis, indicates the (dis)similarity between two observations. The higher the height of the fusion, the less similar the observations are. Note that, conclusions about the proximity of two observations can be drawn only based on the height where branches containing those two observations first are fused. We cannot use the proximity of two observations along the horizontal axis as a criteria of their similarity.

It’s also possible to draw the dendrogram with a border around the selected clusters by using `rect.hclust()` of R stats. The argument border is used to specify the border colors for the rectangles.

```{r}
plot(hclust_ward, cex = 0.6)
rect.hclust(hclust_ward, 
            k = 6, 
            border = 2:5)
```

### Visually-driven hierarchical clustering analysis

We perform visually-driven hiearchical clustering analysis by using `heatmaply` package.

#### Transforming the data frame into a matrix

The dataframe has to be converted to a data matrix to plot as a heatmap.

```{r}
shan_ict_mat <- data.matrix(shan_ict)
```

#### Plotting interactive cluster heatmap using heatmaply()

```{r}
heatmaply(normalize(shan_ict_mat),
          Colv=NA,
          dist_method = "euclidean",
          hclust_method = "ward.D",
          seriate = "OLO",
          colors = Blues,
          k_row = 6,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="Geographic Segmentation of Shan State by ICT indicators",
          xlab = "ICT Indicators",
          ylab = "Townships of Shan State"
          )
```

### Mapping the clusters formed

Based on the dendragram above, we have decided to retain six clusters.

`cutree()` of R Base will be used in the code chunk below to derive a 6-cluster model.

```{r}
groups <- as.factor(cutree(hclust_ward, k=6))
```

*groups* is a list object to be appended to *shan_sf*.

The following steps are done in the join function: - the groups list object will be converted into a matrix; - `cbind()` is used to append groups matrix onto *shan_sf* to produce an output simple feature object called *shan_sf_cluster*; - - rename of **dplyr** package is used to rename as.matrix.groups field as CLUSTER.

```{r}
shan_sf_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER`=`as.matrix.groups.`)
```

```{r}
qtm(shan_sf_cluster, "CLUSTER")
```

From the map, we can see that the clusters are fragmented and are not based on location. This is one limitation when using non-spatial clustering algorithms.

## Spatially Constrained Clustering: SKATER approach

Derive spatially constrained cluster using `skater()`.

### Converting into SpatialPolygonsDataFrame

We convert *shan_sf* into SpatialPolygonsDataFrame. This is because SKATER function only support **sp** objects such as SpatialPolygonDataFrame.

```{r}
shan_sp <- as_Spatial(shan_sf)
```

### Computing neighbour list

Using `poly2nd()` of **spdep** package will be used to compute the neighbours list from the polygon list.

```{r}
shan.nb <- poly2nb(shan_sp)
summary(shan.nb)
```

We plot the neighbours list on *shan_sp*.

The first plot command provides the boundaries. The second plot command is the plot of the neighbour list object, with coordinates applied to extract the centroids of the polygons.

These are used as the nodes for the graph representation.

We also set the colour to blue and specify add=TRUE to plot the network on the top of the boundaries.

```{r}
coords <- st_coordinates(
  st_centroid(st_geometry(shan_sf)))
```

```{r}
plot(st_geometry(shan_sf), 
     border=grey(.5))
plot(shan.nb,
     coords, 
     col="blue", 
     add=TRUE)
```

Since the boundary map extends further than the graph, we plot it first.

### Computing minimum spanning tree

#### Calculating edge costs

`nbcosts()` of **spdep** package is used to compute the cost of each edge (distance between each node). This function compute this distance using a data.frame with observations vector in each node.

```{r}
lcosts <- nbcosts(shan.nb, shan_ict)
```

For each observation, this gives the pairwise dissimilarity between its values on the five variables and the values for the neighbouring observation (from the neighbour list). This is a generalised weight for a spatial weights matrix.

Next, we incorporate these costs into a weights object in the same way as we calculated the inverse of distance weights. We convert the neighbour list to a list weights object by specifying just the computed **lcosts** as the weights.

In order to achieve this, `nb2listw()` of **spdep** package is used

We specify the style="B" to make sure the cost values are not row-standardised.

```{r}
shan.w <- nb2listw(shan.nb, 
                   lcosts, 
                   style="B")
summary(shan.w)
```

#### Computing minimum spanning tree

The minimum spanning tree is computed by mean of the `mstree()` of **spdep** package.

We can then check its class and dimension.

```{r}
shan.mst <- mstree(shan.w)
```

```{r}
class(shan.mst)
```

```{r}
dim(shan.mst)
```

Note that dimension is 54 not 55, as the minimum spnning tree consists of n-1 edges (links) in order to traverse all the nodes.

```{r}
head(shan.mst)
```

The plot below shows the observation of the nodes in addition to the edge.

```{r}
plot(st_geometry(shan_sf), 
                 border=gray(.5))
plot.mst(shan.mst, 
         coords, 
         col="blue", 
         cex.lab=0.7, 
         cex.circles=0.005, 
         add=TRUE)
```

### Computing spatially constrained clusters using SKATER method

Compute the spatially constrained cluster using `skater()` of **spdep** package.

```{r}
clust6 <- spdep::skater(edges = shan.mst[,1:2], 
                 data = shan_ict, 
                 method = "euclidean", 
                 ncuts = 5)
```

The `skater()` takes three mandatory arguments: - the first two columns of the MST matrix (i.e. not the cost), - the data matrix (to update the costs as units are being grouped), and - the number of cuts. Note: It is set to one less than the number of clusters. So, the value specified is not the number of clusters, but the number of cuts in the graph, one less than the number of clusters.

The result of the`skater()` is an object of class skater.

```{r}
str(clust6)
```

We can observe that the groups vector contains the labels of the cluster to which each observation belongs.

Sum of squares measures are given as *ssto* for the total and *ssw* to show the effect of each of the cuts on the overall criterion.

```{r}
ccs6 <- clust6$groups
ccs6
```

Show the number of observations in each cluster.

```{r}
table(ccs6)
```

```{r}
plot(st_geometry(shan_sf), 
     border=gray(.5))
plot(clust6, 
     coords, 
     cex.lab=.7,
     groups.colors=c("red","green","blue", "brown", "pink"),
     cex.circles=0.005, 
     add=TRUE)
```

### Visualising the clusters in choropleth map

Plot the clusters using the SKATER method.

```{r}
groups_mat <- as.matrix(clust6$groups)
shan_sf_spatialcluster <- cbind(shan_sf_cluster, as.factor(groups_mat)) %>%
  rename(`SP_CLUSTER`=`as.factor.groups_mat.`)
qtm(shan_sf_spatialcluster, "SP_CLUSTER")
```

We can place both the hierarchical clustering and spatially constrained hierarchical clustering maps next to each other.

```{r}
hclust.map <- qtm(shan_sf_cluster,
                  "CLUSTER") + 
  tm_borders(alpha = 0.5) 

shclust.map <- qtm(shan_sf_spatialcluster,
                   "SP_CLUSTER") + 
  tm_borders(alpha = 0.5) 

tmap_arrange(hclust.map, shclust.map,
             asp=NA, ncol=2)
```

## Spatially Constrained Clustering: ClustGeo Method

Using functions provided by **ClustGeo** package to perform non-spatially constrained hierarchical cluster analysis and spatially constrained cluster analysis.

### About ClustGeo package

**ClustGeo** package is an R package specially designed to support the need of performing spatially constrained cluster analysis. More specifically, it provides a Ward-like hierarchical clustering algorithm called `hclustgeo()` including spatial/geographical constraints.

In the nutshell, the algorithm uses two dissimilarity matrices D0 and D1 along with a mixing parameter alpha, whereby the value of alpha must be a real number between \[0, 1\]. D0 can be non-Euclidean and the weights of the observations can be non-uniform. It gives the dissimilarities in the **attribute/clustering variable space**. D1, on the other hand, gives the dissimilarities in the **constraint space**. The criterion minimised at each stage is a convex combination of the homogeneity criterion calculated with D0 and the homogeneity criterion calculated with D1.

The idea is then to determine a value of alpha which increases the spatial contiguity without deteriorating too much the quality of the solution based on the variables of interest. This need is supported by a function called `choicealpha()`.

### Ward-like hierarchical clustering: ClustGeo

To perform non-spatially constrained hierarchical clustering, we only need to provide the function a dissimilarity matrix as shown in the code chunk below.

```{r}
nongeo_cluster <- hclustgeo(proxmat)
plot(nongeo_cluster, cex = 0.5)
rect.hclust(nongeo_cluster, 
            k = 6, 
            border = 2:5)
```

Note that the dissimilarity matrix must be an object of class *dist*, i.e. an object obtained with the function `dist()`.

#### Mapping the clusters formed

```{r}
groups <- as.factor(cutree(nongeo_cluster, k=6))
```

```{r}
shan_sf_ngeo_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
```

```{r}
qtm(shan_sf_ngeo_cluster, "CLUSTER")
```

## Spatially Constrained Hierarchical Clustering

Before we can performed spatially constrained hierarchical clustering, a spatial distance matrix will be derived by using `st_distance()` of sf package.

```{r}
dist <- st_distance(shan_sf, shan_sf)
distmat <- as.dist(dist)
```

`as.dist()` is used to convert the data frame into matrix. `choicealpha()` will be used to determine a suitable value for the mixing parameter alpha.

```{r}
cr <- choicealpha(proxmat, distmat, range.alpha = seq(0, 1, 0.1), K=6, graph = TRUE)
```

We use alpha = 0.2

```{r}
clustG <- hclustgeo(proxmat, distmat, alpha = 0.2)
```

`cutree()` is used to derive the cluster object.

```{r}
groups <- as.factor(cutree(clustG, k=6))
```

We will then join back the group list with *shan_sf* polygon feature data frame

```{r}
shan_sf_Gcluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
```

```{r}
qtm(shan_sf_Gcluster, "CLUSTER")
```

## Visual Interpretation of Clusters

### Visualising individual clustering variable

We reveal the distribution of a clustering variable (i.e RADIO_PR) by cluster.

```{r}
ggplot(data = shan_sf_ngeo_cluster,
       aes(x = CLUSTER, y = RADIO_PR)) +
  geom_boxplot()
```

The boxplot reveals that Cluster 3 displays highest mean Radio Ownership Per Thousand Household. This is followed by Cluster 2, 1, 4, 6 and 5.

### Multivariate Visualisation

Parallel coordinate plot can be used to reveal clustering variables by cluster very effectively. We use `ggparcoord()` of **GGally** package

```{r}
ggparcoord(data = shan_sf_ngeo_cluster, 
           columns = c(17:21), 
           scale = "globalminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of ICT Variables by Cluster") +
  facet_grid(~ CLUSTER) + 
  theme(axis.text.x = element_text(angle = 30))
```

The parallel coordinate plot above reveals that households in Cluster 4 townships tend to own the highest number of TV and mobile-phone. On the other hand, households in Cluster 5 tends to own the lowest of all the five ICT.

Note that the scale argument of `ggparcoor()` provide several methods to scale the clustering variables. They are:

-   std: univariately, subtract mean and divide by standard deviation.
-   robust: univariately, subtract median and divide by median absolute deviation.
-   uniminmax: univariately, scale so the minimum of the variable is zero, and the maximum is one.
-   globalminmax: no scaling is done; the range of the graphs is defined by the global minimum and the global maximum. center: use uniminmax to standardize vertical height, then center each variable at a value specified by the scaleSummary param.
-   centerObs: use uniminmax to standardize vertical height, then center each variable at the value of the observation specified by the centerObsID param

There is no one best scaling method to use. You should explore them and select the one that best meet your analysis need.

Last but not least, we can also compute the summary statistics such as mean, median, sd, etc to complement the visual interpretation.

In the code chunk below, `group_by()` and `summarise()` of dplyr are used to derive mean values of the clustering variables.

```{r}
shan_sf_ngeo_cluster %>% 
  st_set_geometry(NULL) %>%
  group_by(CLUSTER) %>%
  summarise(mean_RADIO_PR = mean(RADIO_PR),
            mean_TV_PR = mean(TV_PR),
            mean_LLPHONE_PR = mean(LLPHONE_PR),
            mean_MPHONE_PR = mean(MPHONE_PR),
            mean_COMPUTER_PR = mean(COMPUTER_PR))
```
