---
title: "In-class Exercise 8 | Geographically Weighted Predictive Models"
subtitle: "Data Preparation for Take-home Exercise 3"
date: "Oct 25, 2025"
date-modified: "last-modified"
author: Linh Nguyen
execute:
  echo: true
  eval: true
  freeze: auto
  message: false
  warning: false
format:
  html:
    code-link: true
    toc: true
number-sections: true
number-offset: 1
editor: visual
---

## Overview

### The data

The data used in this exercise includes Aspatial and Geospatial datasets, categorized below based on whether they primarily contain attribute information or geometric features.

#### Aspatial

These datasets primarily contain attribute information, some of which do not have direct geographic coordinates.

| Dataset Name | Description | Source | File Type |
|:-----------------|:-----------------|:-----------------|:-----------------|
| HDB Resale data | A list of HDB resale transacted prices in Singapore from Jan 2017 onwards. | Data.gov.sg | csv |
| Primary school data | Extracted from the list on General information of schools from the data.gov portal. | Data.gov.sg | csv |
| CBD coordinates | Coordinates obtained from Google. | Retrieved/Scraped from other sources (Google) | N/A (List data) |
| Shopping malls data | A list of Shopping malls in Singapore. | Retrieved/Scraped from other sources (Wikipedia) | N/A (List data) |
| Good primary schools list | A list of primary schools ordered in ranking in terms of popularity. | Retrieved/Scraped from other sources (Local Salary Forum) | N/A (List data) |

#### Geospatial

These datasets include geometric features and locational factors, with or without explicit geographic coordinates in the primary list/file type.

| Dataset Name | Description | Source | File Type |
|:-----------------|:-----------------|:-----------------|:-----------------|
| *MP14_SUBZONE_WEB_PL* | A polygon feature data providing information of URA 2014 Master Plan Planning Subzone boundary data. | Data.gov.sg | ESRI shapefile |
| Eldercare data | A list of eldercare in Singapore. | Data.gov.sg | shapefile |
| Hawker Centre data | A list of hawker centres in Singapore. | Data.gov.sg | geojson |
| Parks data | A list of parks in Singapore. | Data.gov.sg | geojson |
| Supermarket data | A list of supermarkets in Singapore. | Data.gov.sg | geojson |
| CHAS clinics data | A list of CHAS clinics in Singapore. | Data.gov.sg | geojson |
| Childcare service data | A list of childcare services in Singapore. | Data.gov.sg | geojson |
| Kindergartens data | A list of kindergartens in Singapore. | Data.gov.sg | geojson |
| MRT data | A list of MRT/LRT stations in Singapore with the station names and codes. | Datamall.lta.gov.sg | shapefile |
| Bus stops data | A list of bus stops in Singapore. | Datamall.lta.gov.sg | shapefile |

### Loading R packages

```{r}
pacman::p_load(httr, tidyverse, sf, tmap, jsonlite, progress, spdep, GWmodel,
               rsample, Metrics, SpatialML, knitr, kableExtra, spatialRF, randomForestExplainer)
```

## Geocoding HDB resale data

We will use Reverse ZIPCode API.

### Basic version

```{r}
HDBresale <- read_csv("data/rawdata/HDBresale.csv") %>% 
  filter (flat_type == "4 ROOM") %>% 
  filter (month >= "2025-09" & month <"2025-10")
```

```{r}
geocode <- function(block, streetname) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  address <- paste(block, streetname, sep=" ")
  query <- list("searchVal" = address,
                "returnGeom" = "Y",
                "getAddrDetails" = "N",
                "pageNum" = "1")
  
  res <- GET(base_url, query=query)
  restext <- content(res, as="text")
  
  output <- fromJSON(restext) %>%
    as.data.frame %>%
    select(results.LATITUDE, results.LONGITUDE)
  
  return(output)
}
```

```{r}
HDBresale$street_name <- gsub("ST\\.", "SAINT", HDBresale$street_name)
```

```{r}
HDBresale$LATITUDE <- 0
HDBresale$LONGTITUDE <- 0

for (i in 1:nrow(HDBresale)) {
  temp_output <- geocode(HDBresale[i, 4], HDBresale[i, 5])
  
  HDBresale$LATITUDE[i] <- temp_output$results.LATITUDE
  HDBresale$LONGITUDE[i] <- temp_output$results.LONGITUDE
}
```

```{r}
HDBresale_sf <- HDBresale %>%
  st_as_sf(
    coords = c("LONGITUDE", "LATITUDE"),
    crs=4326
  ) %>%
  st_transform(crs=3414)
```

```{r}
preschool = st_read("data/rawdata/PreSchoolsLocation.kml") %>% st_transform(crs=3414)
```

```{r}
within_350m <- st_is_within_distance(
  HDBresale_sf, preschool, dist = 350
)

HDBresale_sf <- HDBresale_sf %>% 
  mutate(prischool_350m = map_int(within_350m, length))
```

### Data Sampling

```{r}
mdata <- read_rds("data/rawdata/mdata.rds")
```

```{r}
set.seed(1234)
HDB_sample <- mdata %>% sample_n(1500)
```

### Checking of overlapping point

```{r}
overlapping_points <- HDB_sample %>%
  mutate(overlap = lengths(st_equals(., .)) >1)
```

### Spatial jitter

This is to move the point features by 5m to avoid overlapping point features.

```{r}
HDB_sample <- HDB_sample %>%
  st_jitter(amount=1)
```

### Data Split

```{r}
set.seed(1234)
resale_split <- initial_split(HDB_sample, prop = 6.67/10,)

train_data <- training(resale_split)
test_data <- testing(resale_split)

```
