---
title: "Hands-on 2B | Second Order Spatial Point Patterns Analysis"
subtitle: "In this exercise, we will learn to apply 2nd-order spatial point pattern analysis methods in R, including G, F, K, and L functions, to evaluate spatial point distributions and perform hypothesis testing using the spatstat package."
date: "Sep 2, 2025"
date-modified: "last-modified"
author: Linh Nguyen
execute:
  echo: true
  eval: true
  freeze: auto
  message: false
  warning: false
format:
  html:
    code-link: true
    toc: true
number-sections: true
number-offset: 1
editor: visual
---

## Overview

Spatial Point Pattern Analysis is the evaluation of the pattern or distribution, of a set of points on a surface. The point can be location of:

-   events such as crime, traffic accident and disease onset, or
-   business services (coffee and fastfood outlets) or facilities such as childcare and eldercare.

Using appropriate functions of [spatstat](https://cran.r-project.org/web/packages/spatstat/), this hands-on exercise aims to discover the spatial point processes of childecare centres in Singapore.

The specific questions we would like to answer are as follows:

-   are the childcare centres in Singapore randomly distributed throughout the country?\
-   if the answer is not, then the next logical question is where are the locations with higher concentration of childcare centres?

This hands-on exercise continues from [Hands-on Exercise 2A](/hands-on_ex/hands-on_ex02/hands-on_ex02a)

### Learning Outcome

-   Importing and managing geospatial data using `sf` and `spatstat` packages.
-   Converting spatial data formats from `sf` to `spatstat`'s `ppp` format.
-   Handling and correcting duplicate spatial points in datasets.
-   Defining and applying spatial windows (`owin` objects) for focused analysis.
-   Conducting 2nd-order spatial point pattern analyses using G, F, K, and L functions.
-   Performing Monte Carlo simulations and hypothesis testing to assess spatial randomness.
-   Visualizing spatial patterns and statistical results using appropriate plotting functions.

### The Data

| Dataset | Description | Source | Format |
|------------------|------------------|------------------|------------------|
| **ChildCareServices** | Point data containing location and attributes of childcare centers. | Data.gov.sg | GeoJSON |
| **MasterPlan2019SubzoneBoundaryNoSeaKML** | Polygon data with URA 2014 Master Plan Planning Subzone boundaries. | Data.gov.sg | KML |

### Installing and Loading the R Packages

The following R packages will be used in this exercise:

| Package | Purpose | Use Case in Exercise |
|------------------------|------------------------|------------------------|
| **sf** | Imports, manages, and processes vector-based geospatial data. | Handling vector geospatial data in R. |
| **spatstat** | Provides tools for point pattern analysis. | Performing 1st- and 2nd-order spatial point pattern analysis and deriving kernel density estimation (KDE). |
| **tmap** | Creates static and interactive thematic maps using cartographic quality elements. | Plotting cartographic quality static point patterns maps or interactive maps by using [**leaflet**](https://leafletjs.com/) API. |
| **tidyverse** | Supports data manipulation, analysis, and visualization in R a more intuitive and efficient way | Tidying data |
| **rvest** | Scraping (or harvesting) data from web pages, or tidying HTML data. | Tidying data by extracting values from the HTML description |

To install and load these packages in R, use the following code:

```{r}
pacman::p_load(sf, spatstat, tmap, tidyverse, rvest)
```

**Reproducibility**

As this document involves Monte Carlo simulations, we will set the seed to ensure reproducibility

```{r}
set.seed(1234)
```

## Importing and Wrangling Geospatial Data Sets

### Master Plan 2019 Subzone Boundary No Sea KML

#### Import data

```{r}
#| eval: false
mpsz_sf <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
```

#### Inspect and Reproject to Same Projection System

Before we can use these data for analysis, it is important for us to ensure that they are projected in same projection system.

```{r}
#| eval: false
st_crs(mpsz_sf)
```

Since the ID is `EPSG:4326`, we will set the crs to `EPSG:3414`.

```{r}
#| eval: false
mpsz_sf <- st_transform(mpsz_sf, 3414)
st_crs(mpsz_sf)
```

#### Build a function to extract REGION_N, PLN_AREA_N, SUBZONE_N, SUBZONE_C from Description field

```{r}
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}
```

```{r}
#| eval: false
mpsz_sf <- mpsz_sf %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())
```

```{r}
#| eval: false
#| echo: false
write_rds(mpsz_sf, "data/geospatial/mpsz_sf.rds")
```

```{r}
#| echo: false
mpsz_sf <- read_rds("data/geospatial/mpsz_sf.rds")
```

```{r}
#| eval: false
mpsz_cl <- mpsz_sf %>%
  filter(SUBZONE_N != "SOUTHERN GROUP",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS")
```

```{r}
#| eval: false
#| echo: false
write_rds(mpsz_cl, "data/geospatial/mpsz_cl.rds")
```

```{r}
#| echo: false
mpsz_cl <- read_rds("data/geospatial/mpsz_cl.rds")
```

#### Extract study areas

```{r}
pg <- mpsz_cl %>% filter(PLN_AREA_N == "PUNGGOL")
tm <- mpsz_cl %>% filter(PLN_AREA_N == "TAMPINES")
ck <- mpsz_cl %>% filter(PLN_AREA_N == "CHOA CHU KANG")
jw <- mpsz_cl %>% filter(PLN_AREA_N == "JURONG WEST")
```

#### Creating `owin` object

When analysing spatial point patterns, it is a good practice to confine the analysis with a geographical area like Singapore boundary. In `spatstat`, an object called `owin` is specially designed to represent this polygonal region.

The code chunk below, `as.owin()` of spatstat is used to convert `mpsz_cl` as well as `pg`, `tm`, `ck`, `jw` into `owin` object.

```{r}
sg_owin <- as.owin(mpsz_cl)
pg_owin = as.owin(pg)
tm_owin = as.owin(tm)
ck_owin = as.owin(ck)
jw_owin = as.owin(jw)
```

The `owin` objects above can be displayed by using `plot()` function as below:

```{r}
plot(sg_owin, main='Singapore')
```

```{r}
par(mfrow = c(2,2), mar = c(1, 1, 2, 1))
plot(pg_owin, main = 'Punggol')
plot(tm_owin, main = 'Tampines')
plot(ck_owin, main = 'Choa Chu Kang')
plot(jw_owin, main = 'Jurong West')
```

### Child Care Services

#### Import data

```{r}
childcare_sf <- st_read("data/geospatial/ChildCareServices.geojson")
```

#### Inspect and Reproject to Same Projection System\*\*

```{r}
st_crs(childcare_sf)
```

This dataset is using the WGS84 crs. We will reproject it to SVY21 crs for standardization and analysis.

```{r}
childcare_sf <- st_transform(childcare_sf , crs = 3414)
st_crs(childcare_sf)
```

#### Converting sf data frames to ppp class

`spatstat` requires the point event data in ppp object form. The code chunk below uses `as.ppp()` of `spatstat` package to convert `childcare_sf` to ppp format.

```{r}
childcare_ppp <- as.ppp(childcare_sf)
```

### Combining point events object and owin object

#### Combine `ppp` and `owin` objects

In this last step of geospatial data wrangling, we will extract childcare events that are located within Singapore and each regions by using the code chunk below.

```{r}
childcareSG_ppp = childcare_ppp[sg_owin]
childcare_pg_ppp = childcare_ppp[pg_owin]
childcare_tm_ppp = childcare_ppp[tm_owin]
childcare_ck_ppp = childcare_ppp[ck_owin]
childcare_jw_ppp = childcare_ppp[jw_owin]
```

#### Transform measurement to km

```{r}
childcare_pg_ppp.km = rescale.ppp(childcare_pg_ppp, 1000, "km")
childcare_tm_ppp.km = rescale.ppp(childcare_tm_ppp, 1000, "km")
childcare_ck_ppp.km = rescale.ppp(childcare_ck_ppp, 1000, "km")
childcare_jw_ppp.km = rescale.ppp(childcare_jw_ppp, 1000, "km")
```

#### Plot

```{r}
par(mfrow = c(2,2), mar = c(1, 1, 2, 1))
plot(unmark(childcare_pg_ppp.km), main = "Punggol")
plot(unmark(childcare_tm_ppp.km), main = "Tampines")
plot(unmark(childcare_ck_ppp.km), main = "Choa Chu Kang")
plot(unmark(childcare_jw_ppp.km), main = "Jurong West")
```

## Second-order Spatial Point Patterns Analysis

### Analysing Spatial Point Process Using G-Function

The G function measures the distribution of the distances from an arbitrary event to its nearest event. In this section, you will learn how to compute G-function estimation by using `Gest()` of spatstat package. You will also learn how to perform Monte Carlo simulation test using `envelope()` of `spatstat` package.

#### Choa Chu Kang planning area

##### Computing G-function estimation

The code chunk below is used to compute G-function using `Gest()` of `spastat` package.

```{r}
G_CK = Gest(childcare_ck_ppp, correction = "border")
plot(G_CK, xlim=c(0,500))
```

##### Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

-   Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

-   H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

Monte Carlo test with G-fucntion

```{r}
G_CK.csr <- envelope(childcare_ck_ppp, Gest, nsim = 999)
```

```{r}
plot(G_CK.csr)
```

#### Tampines planning area

We do the same steps with Tampines.

##### Computing G-function estimation

```{r}
G_tm = Gest(childcare_tm_ppp, correction = "best")
plot(G_tm)
```

##### Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

-   Ho = The distribution of childcare services at Tampines are randomly distributed.

-   H1= The distribution of childcare services at Tampines are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

Monte Carlo test with G-fucntion

```{r}
G_TM.csr <- envelope(childcare_tm_ppp, Gest, nsim = 999)
```

```{r}
plot(G_TM.csr)
```

### Analysing Spatial Point Process Using F-Function

The F function estimates the empty space function **F(r)** or its hazard rate **h(r)** from a point pattern in a window of arbitrary shape. In this section, you will learn how to compute F-function estimation by using `Fest()` of `spatstat` package. You will also learn how to perform monta carlo simulation test using `envelope()` of `spatstat` package.

#### Choa Chu Kang planning area

##### Computing F-function estimation

The code chunk below is used to compute F-function using `Fest()` of `spatat` package.

```{r}
F_CK = Fest(childcare_ck_ppp)
plot(F_CK)
```

##### Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

-   Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

-   H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

Monte Carlo test with F-fucntion

```{r}
F_CK.csr <- envelope(childcare_ck_ppp, Fest, nsim = 999)
```

```{r}
plot(F_CK.csr)
```

#### Tampines planning area

We do the same for Tampines.

##### Computing F-function estimation

```{r}
F_TM = Fest(childcare_tm_ppp)
plot(F_TM)
```

##### Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

-   Ho = The distribution of childcare services at Tampines are randomly distributed.

-   H1= The distribution of childcare services at Tampines are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

Monte Carlo test with F-fucntion

```{r}
F_TM.csr <- envelope(childcare_tm_ppp, Fest, nsim = 999)
```

```{r}
plot(F_TM.csr)
```

### Analysing Spatial Point Process Using K-Function

K-function measures the number of events found up to a given distance of any particular event. In this section, you will learn how to compute K-function estimates by using `Kest()` of `spatstat` package. You will also learn how to perform Monte Carlo simulation test using `envelope()` of `spatstat` package.

#### Choa Chu Kang planning area

##### Computing K-function estimate

```{r}
K_ck = Kest(childcare_ck_ppp, correction = "Ripley")
plot(K_ck, . -r ~ r, ylab= "K(d)-r", xlab = "d(m)")
```

##### Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

-   Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

-   H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

The code chunk below is used to perform the hypothesis testing.

```{r}
K_ck.csr <- envelope(childcare_ck_ppp, Kest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(K_ck.csr, . - r ~ r, xlab="d", ylab="K(d)-r")
```

#### Tampines planning area

##### Computing K-function estimate

```{r}
K_tm = Kest(childcare_tm_ppp, correction = "Ripley")
plot(K_tm, . -r ~ r, 
     ylab= "K(d)-r", xlab = "d(m)", 
     xlim=c(0,1000))
```

##### Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

-   Ho = The distribution of childcare services at Tampines are randomly distributed.

-   H1= The distribution of childcare services at Tampines are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

The code chunk below is used to perform the hypothesis testing.

```{r}
K_tm.csr <- envelope(childcare_tm_ppp, Kest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(K_tm.csr, . - r ~ r, 
     xlab="d", ylab="K(d)-r", xlim=c(0,500))
```

### Analysing Spatial Point Process Using L-Function

In this section, you will learn how to compute L-function estimation by using `Lest()` of `spatstat` package. You will also learn how to perform Monte Carlo simulation test using `envelope()` of `spatstat` package.

#### Choa Chu Kang planning area

##### Computing L-function estimate

```{r}
L_ck = Lest(childcare_ck_ppp, correction = "Ripley")
plot(L_ck, . -r ~ r, 
     ylab= "L(d)-r", xlab = "d(m)")
```

##### Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

-   Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

-   H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

The code chunk below is used to perform the hypothesis testing.

```{r}
L_ck.csr <- envelope(childcare_ck_ppp, Lest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(L_ck.csr, . - r ~ r, xlab="d", ylab="L(d)-r")
```

#### Tampines planning area

##### Computing K-function estimate

```{r}
L_tm = Lest(childcare_tm_ppp, correction = "Ripley")
plot(L_tm, . -r ~ r, 
     ylab= "L(d)-r", xlab = "d(m)", 
     xlim=c(0,1000))
```

##### Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

-   Ho = The distribution of childcare services at Tampines are randomly distributed.

-   H1= The distribution of childcare services at Tampines are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

The code chunk below is used to perform the hypothesis testing.

```{r}
L_tm.csr <- envelope(childcare_tm_ppp, Lest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(L_tm.csr, . - r ~ r, 
     xlab="d", ylab="L(d)-r", xlim=c(0,500))
```
