---
title: "Hands-on 2A | First-order Spatial Point Patterns Analysis Methods"
subtitle: "In this exercise, we will learn to analyze spatial point patterns in R, including importing geospatial data, performing kernel density estimation and nearest neighbor analysis, and visualizing results using spatstat, sf, and tmap packages."
date: "Sep 2, 2025"
date-modified: "last-modified"
author: Linh Nguyen
execute:
  echo: true
  eval: true
  freeze: auto
  message: false
  warning: false
format:
  html:
    code-link: true
    toc: true
number-sections: true
number-offset: 1
editor: visual
---

## Overview

Spatial Point Pattern Analysis is the evaluation of the pattern or distribution, of a set of points on a surface. The point can be location of:

-   events such as crime, traffic accident and disease onset, or
-   business services (coffee and fastfood outlets) or facilities such as childcare and eldercare.

Using appropriate functions of [spatstat](https://cran.r-project.org/web/packages/spatstat/), this hands-on exercise aims to discover the spatial point processes of childecare centres in Singapore.

The specific questions we would like to answer are as follows:

-   are the childcare centres in Singapore randomly distributed throughout the country?\
-   if the answer is not, then the next logical question is where are the locations with higher concentration of childcare centres?

## Learning Outcome

-   Importing and managing geospatial data using `sf` and `spatstat` packages.
-   Converting simple feature data frames to various spatial data formats.
-   Performing kernel density estimation (KDE) using different methods and bandwidths.
-   Conducting nearest neighbor analysis to test spatial point distribution.
-   Visualizing spatial patterns and KDE results using `tmap` and `spatstat`.
-   Handling duplicate spatial points and creating point patterns for analysis.
-   Applying appropriate statistical methods for confirmatory spatial point pattern analysis.

## The Data

| Dataset | Description | Source | Format |
|-----------------|--------------------|-----------------|-----------------|
| **ChildCareServices** | Point data containing location and attributes of childcare centers. | Data.gov.sg | GeoJSON |
| **MasterPlan2019SubzoneBoundaryNoSeaKML** | Polygon data with URA 2014 Master Plan Planning Subzone boundaries. | Data.gov.sg | KML |

## Installing and Loading the R Packages

The following R packages will be used in this exercise:

| Package | Purpose | Use Case in Exercise |
|-------------------|--------------------------|---------------------------|
| **sf** | Imports, manages, and processes vector-based geospatial data. | Handling vector geospatial data in R. |
| **spatstat** | Provides tools for point pattern analysis. | Performing 1st- and 2nd-order spatial point pattern analysis and deriving kernel density estimation (KDE). |
| **terra** | Creating, reading, manipulating, and writing raster and vector data. | Convert image output generate by `spatstat` into `terra` format. |
| **tmap** | Creates static and interactive thematic maps using cartographic quality elements. | Plotting cartographic quality static point patterns maps or interactive maps by using [**leaflet**](https://leafletjs.com/) API. |
| **rvest** | Scraping (or harvesting) data from web pages, or tidying HTML data. | Tidying data by extracting values from the HTML description |

To install and load these packages in R, use the following code:

```{r}
pacman::p_load(sf, terra, spatstat, tmap, tidyverse, rvest)
```

**Reproducibility**

As this document involves Monte Carlo simulations, we will set the seed to ensure reproducibility

```{r}
set.seed(1234)
```

## Importing and Wrangling Geospatial Data Sets

### Master Plan 2019 Subzone Boundary No Sea KML

#### Import data

```{r}
mpsz_sf <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
```

#### Inspect and Reproject to Same Projection System

Before we can use these data for analysis, it is important for us to ensure that they are projected in same projection system.

```{r}
st_crs(mpsz_sf)
```

Since the ID is `EPSG:4326`, we will set the crs to `EPSG:3414`.

```{r}
mpsz_sf <- st_transform(mpsz_sf, 3414)
st_crs(mpsz_sf)
```

#### Build a function to extract REGION_N, PLN_AREA_N, SUBZONE_N, SUBZONE_C from Description field

```{r}
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}
```

```{r}
mpsz_sf <- mpsz_sf %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())
```

```{r}
mpsz_cl <- mpsz_sf %>%
  filter(SUBZONE_N != "SOUTHERN GROUP",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS")
```

```{r}
write_rds(mpsz_cl, 
          "data/geospatial/mpsz_cl.rds")
```

### Child Care Services

#### Import data

```{r}
childcare_sf <- st_read("data/geospatial/ChildCareServices.geojson")
```

#### Inspect and Reproject to Same Projection System**

```{r}
st_crs(childcare_sf)
```

This dataset is using the WGS84 crs. We will reproject it to SVY21 crs for standardization and analysis.

```{r}
childcare_sf <- st_transform(childcare_sf , crs = 3414)
st_crs(childcare_sf)
```

The childcare dataset has been reprojected to SVY21 successfully.

## Mapping the Geospatial Datasets

After checking the referencing system of each geospatial data data frame, it is also useful for us to plot a map to show their spatial patterns.

### Static Map

```{r}
# add polygon layer of the subzone based on sg masterplan
tm_shape(mpsz_sf) + tm_polygons() +
# add dot layer to show the locations of childcare centres
tm_shape(childcare_sf) + tm_dots() +
tm_layout()
```

When 2 datasets are overlayed together, it shows the locations of childcare centres on the Singapore island. Since all the geospatial layers are within the same map context, it means their referencing system and coordinate values are referred to similar spatial context. This consistency is crucial for accurate geospatial analysis.

### Interactive Map

Alternatively, we can also prepare an interactive point symbol map by using the code chunk below.

```{r}
tmap_mode('view')
tm_shape(childcare_sf)+
  tm_dots()
```

```{r}
tmap_mode('plot')
```

In interactive mode, `tmap` uses the Leaflet for R API, allowing you to freely navigate, zoom, and click on features for detailed information. You can also change the map's background using layers like ESRI.WorldGrayCanvas, OpenStreetMap, and ESRI.WorldTopoMap, with ESRI.WorldGrayCanvas as the default.

```{r}
tmap_mode('view')
tm_basemap("Esri.WorldTopoMap") +
tm_shape(childcare_sf) +
  tm_dots(alpha = 0.5)
```

```{r}
tmap_mode('plot')
```

::: callout-tip
Remember to switch back to plot mode after interacting to avoid connection issues and limit interactive maps to fewer than 10 in RMarkdown documents for Netlify publishing.
:::

## Geospatial Data Wrangling

`spatstat` relies on its own specific data structures like **ppp** (planar point pattern) for point data and **owin** for observation windows. In this section, you will learn how to convert sf (Simple Features) objects into spatstat ppp and owin object.

### Converting sf data frames to ppp class

`spatstat` requires the point event data in ppp object form. The code chunk below uses `as.ppp()` of `spatstat` package to convert `childcare_sf` to ppp format.

```{r}
childcare_ppp <- as.ppp(childcare_sf)
```

Next, `class()` of Base R will be used to verify the object class of `childcare_ppp`.

```{r}
class(childcare_ppp)
```

Great, it is in ppp object class!

You can take a quick look at the summary statistics of the newly converted ppp object by using the code chunk below.

```{r}
summary(childcare_ppp)
```

### Creating `owin` object

When analysing spatial point patterns, it is a good practice to confine the analysis with a geographical area like Singapore boundary. In `spatstat`, an object called `owin` is specially designed to represent this polygonal region.

The code chunk below, `as.owin()` of spatstat is used to covert `mpsz_sf` into `owin` object of `spatstat`.

```{r}
sg_owin <- as.owin(mpsz_cl)
```

Again, `class()` will be used to verify the object class of `sg_owin`.

```{r}
class(sg_owin)
```

The result above confirmed that `sg_owin` is indeed in `owin` object.

`sg_owin` object can be displayed by using `plot()` function.

```{r}
plot(sg_owin)
```

### Combining point events object and owin object

In this last step of geospatial data wrangling, we will extract childcare events that are located within Singapore by using the code chunk below.

```{r}
childcareSG_ppp = childcare_ppp[sg_owin]
```

The output object combined both the point and polygon feature in one ppp object class as shown below.

```{r}
childcareSG_ppp
```

## Clark-Evan Test for Nearest Neighbour Analysis

**Nearest Neighbor Analysis (NNA)** is a spatial statistics method that calculates the average distance between each point and its closest neighbor to determine if a pattern of points is clustered, dispersed, or randomly distributed.

Clark-Evans test is a specific statistical method used within NNA to quantify whether a point pattern is clustered, random, or uniformly spaced, using the Clark-Evans aggregation index (R) to describe this pattern. NNA provides a numerical value that describes the degree of clustering or regularity, and the Clark-Evans test calculates a specific index (R) for this purpose

::: callout-note 
NNA is a broad term for methods that examine the spatial distribution of points, whereas the Clark-Evans test is a specific statistical method used within NNA to quantify whether a point pattern is clustered, random, or uniformly spaced, using the Clark-Evans aggregation index (R) to describe this pattern. NNA provides a numerical value that describes the degree of clustering or regularity, and the Clark-Evans test calculates a specific index (R) for this purpose. 
:::

In this section, we will perform the Clark-Evans test of aggregation for a spatial point pattern by using [clarkevans.test()](https://cran.r-project.org/web/packages/spatstat.explore/refman/spatstat.explore.html#clarkevans.test) of `spatstat.explore` package.

The test hypotheses are:

-   Ho = The distribution of childcare services are randomly distributed.

-   H1= The distribution of childcare services are not randomly distributed.

The 95% confident interval will be used.

### Perform the Clark-Evans test without CSR

`clarkevans.test()` of `spatstat.explore` package support two Clark-Evans test, namely: without CRS and with CRS. In the code chunk below, Clark-Evans test without CSR method is used.

```{r}
clarkevans.test(childcareSG_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"))
```

As p-value is smaller than 0.05, we can reject the null hypothesis and infer that the distribution of childcare centres in Singapore is not random but rather clustered (due to alternative=“clustered”).

### Perform the Clark-Evans test with CSR

In the code chunk below, the argument method = “MonteCarlo” is used. In this case, the p-value for the test is computed by comparing the observed value of R to the results obtained from nsim (i.e. 39, 99, 999) simulated realisations of Complete Spatial Randomness conditional on the observed number of points.

```{r}
clarkevans.test(childcareSG_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                method="MonteCarlo",
                nsim=99)
```

As p-value is still smaller than 0.05, we can still reject the null hypothesis and infer that the distribution of childcare centres in Singapore is not random but rather clustered (due to alternative=“clustered”).

## Kernel Density Estimation Method

**Kernel Density Estimation (KDE)** is a valuable tool for visualising and analyzing first-order spatial point patterns. It is widely considered a method within **Exploratory Spatial Data Analysis (ESDA)** because it’s used to visualize and understand spatial data patterns by transforms discrete point data (like locations of childcare service, crime incidents or disease cases) into continuous density surfaces that reveal clusters and variations in event occurrences, without making prior assumptions about data distribution. It helps to begin understanding data distribution, identify hotspots, and explore relationships between spatial variables before performing more rigorous analysis.

In this section, you will learn how to compute the kernel density estimation (KDE) of childcare services in Singapore.

### Working with automatic bandwidth selection method

The code chunk below computes a kernel density by using the following configurations of `density()` of **spatstat**:

-   `bw.diggle()` automatic bandwidth selection method. Other recommended methods are `bw.CvL()`, `bw.scott()` or `bw.ppl()`.

-   The smoothing kernel used is gaussian, which is the default. Other smoothing methods are: “epanechnikov”, “quartic” or “disc”.

-   The intensity estimate is corrected for edge effect bias by using method described by Jones (1993) and Diggle (2010, equation 18.9). The default is FALSE.

```{r}
kde_SG_diggle <- density(
  childcareSG_ppp,
  sigma=bw.diggle,
  edge=TRUE,
  kernel="gaussian") 
```

::: callout-note
The output of `density()` of spatstat is an **im** class, represents a two-dimensional pixel image. It’s a class used to store and manipulate raster data, where the spatial domain is divided into a grid of rectangular pixels. Each pixel has an associated value, which can be numerical or a factor.
:::

The `plot()` function of Base R is then used to display the kernel density derived.

```{r}
plot(kde_SG_diggle)
```
In the code chunk below, `summary()` of Base R is used to print the summary report of the

```{r}
summary(kde_SG_diggle)
```

::: callout-warning
The output of `density()` of spatstat is an **im** class, represents a two-dimensional pixel image. It’s a class used to store and manipulate raster data, where the spatial domain is divided into a grid of rectangular pixels. Each pixel has an associated value, which can be numerical or a factor.
:::

Before we move on to next section, it is good to know that you can retrieve the bandwidth used to compute the kde layer by using the code chunk below.

```{r}
bw <- bw.diggle(childcareSG_ppp)
bw
```
### Rescalling KDE values

In the code chunk below, rescale.ppp() is used to covert the unit of measurement from meter to kilometer.

```{r}
childcareSG_ppp_km <- rescale.ppp(
  childcareSG_ppp, 1000, "km")
```

Now, we can re-run density() using the resale data set and plot the output kde map.

```{r}
kde_childcareSG_km <- density(childcareSG_ppp_km,
                              sigma=bw.diggle,
                              edge=TRUE,
                              kernel="gaussian")
```

Next, plot() is used to plot the kde object as shown below.

```{r}
plot(kde_childcareSG_km)
```
Notice that output image looks identical to the earlier version, the only changes in the data values (refer to the legend).

### Working with different automatic badwidth methods

Beside `bw.diggle()`, there are three other `spatstat` functions can be used to determine the bandwidth, they are: `bw.CvL()`, `bw.scott()`, and `bw.ppl()`.

Let us take a look at the bandwidth return by these automatic bandwidth calculation methods by using the code chunks below.

::: panel-tabset

#### CvL

```{r}
bw.CvL(childcareSG_ppp_km)
```

#### Scott

```{r}
bw.scott(childcareSG_ppp_km)
```

#### PPL

```{r}
bw.ppl(childcareSG_ppp_km)
```

#### Diggle

```{r}
bw.diggle(childcareSG_ppp_km)
```
:::

Baddeley et. (2016) suggested the use of the `bw.ppl()` algorithm because past experience shown that it tends to produce the more appropriate values when the pattern consists predominantly of tight clusters. But they also insist that if the purpose of once study is to detect a single tight cluster in the midst of random noise then the `bw.diggle()` method seems to work best.

The code chunk beow will be used to compare the output of using `bw.diggle` and `bw.ppl` methods.

```{r}
kde_childcareSG.ppl <- density(childcareSG_ppp_km, 
                               sigma=bw.ppl, 
                               edge=TRUE,
                               kernel="gaussian")
par(mfrow=c(1,2))
plot(kde_childcareSG_km, main = "bw.diggle")
plot(kde_childcareSG.ppl, main = "bw.ppl")
```
### Working with different kernel methods

By default, the kernel method used in `density.ppp()` is gaussian. But there are three other options, namely: Epanechnikov, Quartic and Dics.

The code chunk below will be used to compute three more kernel density estimations by using these three kernel function.

```{r}
par(mfrow=c(2,2))
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="gaussian"), 
     main="Gaussian")
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="epanechnikov"), 
     main="Epanechnikov")
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="quartic"), 
     main="Quartic")
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="disc"), 
     main="Disc")
```
## Fixed and Adaptive KDE

### Computing KDE by using fixed bandwidth

Next, you will compute a KDE layer by defining a bandwidth of 600 meter. Notice that in the code chunk below, the sigma value used is 0.6. This is because the unit of measurement of `childcareSG_ppp_km` object is in kilometer, hence the 600m is 0.6km.

```{r}
kde_childcareSG_fb <- density(childcareSG_ppp_km,
                              sigma=0.6, 
                              edge=TRUE,
                              kernel="gaussian")
plot(kde_childcareSG_fb)
```
### Computing KDE by using adaptive bandwidth

Fixed bandwidth method is very sensitive to highly skew distribution of spatial point patterns over geographical units for example urban versus rural. One way to overcome this problem is by using adaptive bandwidth instead.

In this section, you will learn how to derive adaptive kernel density estimation by using `density.adaptive()` of `spatstat`.

```{r}
kde_childcareSG_ab <- adaptive.density(
  childcareSG_ppp_km, 
  method="kernel")
plot(kde_childcareSG_ab)
```
We can compare the fixed and adaptive kernel density estimation outputs by using the code chunk below.

```{r}
par(mfrow=c(1,2))
plot(kde_childcareSG_fb, main = "Fixed bandwidth")
plot(kde_childcareSG_ab, main = "Adaptive bandwidth")
```
## Plotting cartographic quality KDE map

### Converting gridded output into raster

Next, we will convert the im kernal density objects into SpatRaster object by using rast() of terra package.

```{r}
kde_childcareSG_bw_terra <- rast(kde_childcareSG_km)
```

Again, `class()` is used to verify if `kde_childcareSG_bw_terra` data are belong to `SpatRaster` class.

```{r}
class(kde_childcareSG_bw_terra)
```
Yes, it is indeed in SpatRaster class. Let us take a look at the properties of kde_childcareSG_bw_terra .

```{r}
kde_childcareSG_bw_terra
```
Notice that the crs property is empty.

### Assigning projection systems

In code chunk below, `crs()` of terra is used to assign the CRS information on `kde_childcareSG_bw_terra` layer.

```{r}
crs(kde_childcareSG_bw_terra) <- "EPSG:3414"
```

Let us take a look at the properties of kde_childcareSG_bw_raster RasterLayer.

```{r}
kde_childcareSG_bw_terra
```
Notice that the coordicates reference (i.e. coord. ref.) is in SVY21 now.

### Plotting KDE map with tmap

Finally, we will display the raster in cartographic quality map using **tmap** package.

```{r}
tm_shape(kde_childcareSG_bw_terra) + 
  tm_raster(col.scale = 
              tm_scale_continuous(
                values = "viridis"),
            col.legend = tm_legend(
            title = "Density values",
            title.size = 0.7,
            text.size = 0.7,
            bg.color = "white",
            bg.alpha = 0.7,
            position = tm_pos_in(
              "right", "bottom"),
            frame = TRUE)) +
  tm_graticules(labels.size = 0.7) +
  tm_compass() +
  tm_layout(scale = 1.0)
```
Notice that the raster values are encoded explicitly onto the raster pixel using the values in “layer.1” field.

## First Order SPPA at the Planning Subzone Level

In this section, we would like to further our analysis at the planning area level. For simplicity reason, we will focus on Punggol, Tampines Chua Chu Kand and Jurong West planning areas

### Geospatial data wrangling

#### Extracting study area

The code chunk below will be used to extract the target planning areas.

```{r}
pg <- mpsz_cl %>%
  filter(PLN_AREA_N == "PUNGGOL")
tm <- mpsz_cl %>%
  filter(PLN_AREA_N == "TAMPINES")
ck <- mpsz_cl %>%
  filter(PLN_AREA_N == "CHOA CHU KANG")
jw <- mpsz_cl %>%
  filter(PLN_AREA_N == "JURONG WEST")
```

It is always a good practice to review the extracted areas. The code chunk below will be used to plot the extracted planning areas.

```{r}
par(mfrow=c(2,2))
plot(st_geometry(pg), main = "Ponggol")
plot(st_geometry(tm), main = "Tampines")
plot(st_geometry(ck), main = "Choa Chu Kang")
plot(st_geometry(jw), main = "Jurong West")
```

#### Creating **owin** object

Now, we will convert these sf objects into owin objects that is required by spatstat.

```{r}
pg_owin = as.owin(pg)
tm_owin = as.owin(tm)
ck_owin = as.owin(ck)
jw_owin = as.owin(jw)
```

#### Combining point events object and owin object

```{r}
childcare_pg_ppp = childcare_ppp[pg_owin]
childcare_tm_ppp = childcare_ppp[tm_owin]
childcare_ck_ppp = childcare_ppp[ck_owin]
childcare_jw_ppp = childcare_ppp[jw_owin]
```

Next, rescale.ppp() function is used to trasnform the unit of measurement from metre to kilometre.

```{r}
childcare_pg_ppp.km = rescale.ppp(childcare_pg_ppp, 1000, "km")
childcare_tm_ppp.km = rescale.ppp(childcare_tm_ppp, 1000, "km")
childcare_ck_ppp.km = rescale.ppp(childcare_ck_ppp, 1000, "km")
childcare_jw_ppp.km = rescale.ppp(childcare_jw_ppp, 1000, "km")
```

The code chunk below is used to plot these four study areas and the locations of the childcare centres.

```{r}
par(mfrow=c(2,2))
plot(unmark(childcare_pg_ppp.km), 
  main="Punggol")
plot(unmark(childcare_tm_ppp.km), 
  main="Tampines")
plot(unmark(childcare_ck_ppp.km), 
  main="Choa Chu Kang")
plot(unmark(childcare_jw_ppp.km), 
  main="Jurong West")
```

### Clark and Evans Test

#### Choa Chu Kang planning area

In the code chunk below, `clarkevans.test()` of `spatstat` is used to performs Clark-Evans test of aggregation for childcare centre in Choa Chu Kang planning area.

```{r}
clarkevans.test(childcare_ck_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)
```
#### Tampines planning area

In the code chunk below, the similar test is used to analyse the spatial point patterns of childcare centre in Tampines planning area.

```{r}
clarkevans.test(childcare_tm_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)
```
### Computing KDE surfaces by planning area

The code chunk below will be used to compute the KDE of these four planning area. `bw.diggle` method is used to derive the bandwidth of each

```{r}
par(mfrow=c(2,2))
plot(density(childcare_pg_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Punggol")
plot(density(childcare_tm_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Tempines")
plot(density(childcare_ck_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Choa Chu Kang")
plot(density(childcare_jw_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Jurong West")
```



