---
title: "Hands-on 2B | Second Order Spatial Point Patterns Analysis"
subtitle: "In this exercise, we will learn to apply 2nd-order spatial point pattern analysis methods in R, including G, F, K, and L functions, to evaluate spatial point distributions and perform hypothesis testing using the spatstat package."
date: "Sep 2, 2025"
date-modified: "last-modified"
author: Linh Nguyen
execute:
  echo: true
  eval: true
  freeze: auto
  message: false
  warning: false
format:
  html:
    code-link: true
    toc: true
number-sections: true
number-offset: 1
editor: visual
---
## Overview

Spatial Point Pattern Analysis is the evaluation of the pattern or distribution, of a set of points on a surface. The point can be location of:

-   events such as crime, traffic accident and disease onset, or
-   business services (coffee and fastfood outlets) or facilities such as childcare and eldercare.

Using appropriate functions of [spatstat](https://cran.r-project.org/web/packages/spatstat/), this hands-on exercise aims to discover the spatial point processes of childecare centres in Singapore.

The specific questions we would like to answer are as follows:

-   are the childcare centres in Singapore randomly distributed throughout the country?\
-   if the answer is not, then the next logical question is where are the locations with higher concentration of childcare centres?

This hands-on exercise continues from [Hands-on Exercise 2A](/hands-on_ex/hands-on_ex02/hands-on_ex02a)

## Learning Outcome

-   Importing and managing geospatial data using `sf` and `spatstat` packages.
-   Converting spatial data formats from `sf` to `spatstat`'s `ppp` format.
-   Handling and correcting duplicate spatial points in datasets.
-   Defining and applying spatial windows (`owin` objects) for focused analysis.
-   Conducting 2nd-order spatial point pattern analyses using G, F, K, and L functions.
-   Performing Monte Carlo simulations and hypothesis testing to assess spatial randomness.
-   Visualizing spatial patterns and statistical results using appropriate plotting functions.

## The Data

| Dataset | Description | Source | Format |
|-----------------|--------------------|-----------------|-----------------|
| **ChildCareServices** | Point data containing location and attributes of childcare centers. | Data.gov.sg | GeoJSON |
| **MasterPlan2019SubzoneBoundaryNoSeaKML** | Polygon data with URA 2014 Master Plan Planning Subzone boundaries. | Data.gov.sg | KML |

## Installing and Loading the R Packages

The following R packages will be used in this exercise:

| Package | Purpose | Use Case in Exercise |
|-------------------|--------------------------|---------------------------|
| **sf** | Imports, manages, and processes vector-based geospatial data. | Handling vector geospatial data in R. |
| **spatstat** | Provides tools for point pattern analysis. | Performing 1st- and 2nd-order spatial point pattern analysis and deriving kernel density estimation (KDE). |
| **tmap** | Creates static and interactive thematic maps using cartographic quality elements. | Plotting cartographic quality static point patterns maps or interactive maps by using [**leaflet**](https://leafletjs.com/) API. |
| **tidyverse** | Supports data manipulation, analysis, and visualization in R a more intuitive and efficient way | Tidying data |

To install and load these packages in R, use the following code:

```{r}
pacman::p_load(sf, spatstat, tmap, tidyverse)
```

## Reproducibility

As this document involves Monte Carlo simulations, we will set the seed to ensure reproducibility

```{r}
set.seed(1234)
```

## Spatial Data Wrangling

### Importing the Spatial Data

To import the geographical datasets, we will use `st_read()` from `sf`.

