---
title: "Take-home Exercise 3 | Modelling HDB Resale Prices with Geographically Weighted Methods"
subtitle: "The objective of this take-home exercise is to develop a spatially informed model to determine factors affecting or to predict HDB resale prices in Singapore, making use of appropriate geospatial analytics methods."
date: "Nov 13, 2025"
date-modified: "last-modified"
author: Linh Nguyen
execute:
  echo: true
  eval: true
  freeze: auto
  message: false
  warning: false
  error: false
format:
  html:
    code-link: true       # Clickable function links
    toc: true             # Table of contents
    code-fold: true       # Foldable code
    code-tools: true      # Code tools menu
number-sections: false
number-offset: 1
editor: visual
---

## 1. Introduction

### 1.1 Background & Motivation

### 1.2 Research Objectives

## 2. Data & Methodology

### 2.1 Import R packages

```{r}
pacman::p_load(sf, spdep, GWmodel, tidyverse, lubridate, tmap, ggplot2, corrplot, httr, jsonlite, patchwork, car)
```

### 2.2 Data Sources

|  |  |  |  |
|------------------|------------------|-------------------|------------------|
| **Dataset Name** | **Description** | **Use Case** | **Format** |
| **HDB Resale Flat Prices** | Transaction records of HDB resale flats. Filtered for 4-room flats only. | Primary outcome variable (`resale_price`) and source of structural predictors (`floor_area_sqm`, `remaining_lease_years`, `storey_midpoint`) | CSV |
| **Singapore Boundary** | A polygon feature data providing information of URA 2014 Master Plan Planning Subzone boundary data | Provides spatial context for mapping. Used as basemap reference in visualizations. | SHP (point geometry) |
| **Hawker Centres** | Point locations of hawker centres, representing affordable dining options and community gathering spaces. | Calculate proximity to nearest hawker centre (`dist_hawker`) | GeoJSON |
| **Parks** | Point or polygon locations of national parks, regional parks, and neighborhood parks managed by NParks. | Calculate proximity to nearest park (`dist_park`) | GeoJSON |
| **Supermarkets** | Point locations of major supermarket chains across Singapore. | Calculate proximity to nearest supermarket (`dist_supermarket`) | GeoJSON |
| **Eldercare Centres** | Point locations of eldercare facilities including senior activity centres, senior care centres, and nursing homes. | Calculate proximity to nearest eldercare facility (`dist_eldercare`), relevant for multi-generational households and aging population considerations. | SHP (point geometry) |
| **Preschools** | Point locations of registered preschools across Singapore, including both public and private institutions. | Count number of kindergartens within 350m buffer (`n_preschool_350m`) | GeoJSON |
| **Kindergartens** | Point locations of registered kindergartens across Singapore, including both public and private institutions. | Count number of kindergartens within 350m buffer (`n_kindergarten_350m`) | GeoJSON |
| **Childcare Centres** | Point locations of licensed childcare centres providing infant and toddler care services. | Count number of childcare centres within 350m buffer (`n_childcare_350m`) to evaluate childcare accessibility, important for working parents. | GeoJSON |
| **Bus Stops** | Point locations of all public bus stops in Singapore's bus network. | Count number of bus stops within 350m buffer (`n_busstop_350m`) as a measure of public transport connectivity and accessibility. | SHP (point geometry) |

### 2.3 Data Import & Initial Manipulation

#### 2.3.1 Singapore Boundary

::: panel-tabset
#### Single Boundary

```{r}
#| code-summary: Import Singapore boundary for mapping context

# Import Singapore boundary (adjust path as needed)
singapore_boundary <- st_read(
  dsn = "data/geospatial",
  layer = "MP14_SUBZONE_WEB_PL",  # or your boundary layer name
  quiet = TRUE
) %>%
  st_transform(3414) %>%
  st_union()  # Dissolve to single boundary

cat("✓ Singapore boundary imported\n",
"  CRS:", st_crs(singapore_boundary)$input, "\n")
```

```{r}
#| code-summary: Report the extent of mpsz_svy21 layer by its lower and upper limits

st_bbox(singapore_boundary)
```

#### Planning Area Boundaries

```{r}
#| code-summary: Import planning area boundaries (or use town polygons if available)

planning_areas <- st_read(
  dsn = "data/geospatial",
  layer = "MP14_SUBZONE_WEB_PL",  
  quiet = TRUE
) %>%
  st_transform(3414) %>%
  group_by(PLN_AREA_N) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  rename(town = PLN_AREA_N)
```
:::

#### 2.3.2 HDB Resale Flat Prices

```{r}
#| code-summary: Read CSV file
#| output: true
#| warning: false
#| error: false
#| message: false
#| eval: false # saved rds already

HDBresale <- read_csv("data/rawdata/HDBresale.csv") %>% 
  filter (flat_type == "4 ROOM") %>% 
  filter (month >= "2025-01" & month <"2025-10")
```

```{r}
#| code-summary: Define a function to convert the address into latitude and longitude using OneMapSG API
#| output: true
#| warning: false
#| error: false
#| message: false

geocode <- function(block, streetname) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  address <- paste(block, streetname, sep=" ")
  query <- list("searchVal" = address,
                "returnGeom" = "Y",
                "getAddrDetails" = "N",
                "pageNum" = "1")
  
  res <- GET(base_url, query=query)
  restext <- content(res, as="text")
  
  output <- fromJSON(restext) %>%
    as.data.frame %>%
    select(results.LATITUDE, results.LONGITUDE)
  
  return(output)
}
```

```{r}
#| code-summary: Cleaning street names
#| eval: false

HDBresale$street_name <- gsub("ST\\.", "SAINT", HDBresale$street_name)
```

```{r}
#| code-summary: Apply the function
#| output: false
#| warning: false
#| error: false
#| message: false
#| eval: false # saved rds already, no need to call API again to save time

HDBresale$LATITUDE <- 0
HDBresale$LONGITUDE <- 0

for (i in 1:nrow(HDBresale)) {
  temp_output <- geocode(HDBresale[i, 4], HDBresale[i, 5])
  
  HDBresale$LATITUDE[i] <- temp_output$results.LATITUDE
  HDBresale$LONGITUDE[i] <- temp_output$results.LONGITUDE
}
```

```{r}
#| code-summary: Convert HDB resale to sf
#| eval: false # saved rds already

HDBresale_sf <- HDBresale %>%
  st_as_sf(
    coords = c("LONGITUDE", "LATITUDE"),
    crs=4326
  ) %>%
  st_transform(crs=3414)
```

```{r}
#| code-summary: Save rds for HDBresale_sf
#| eval: false
#| echo: false

write_rds(HDBresale_sf, "data/rds/HDBresale_sf.rds")
```

```{r}
#| echo: false
#| output: false
#| eval: false # save RDS for cleaned and enhanced version already

HDBresale_sf <- read_rds("data/rds/HDBresale_sf.rds")
```

#### 2.3.3 Neighborhood and local facilities data

::: panel-tabset
#### CBD Area

```{r}
#| code-summary: Create CBD points dataframe with SVY21 coordinates
#| eval: false # save RDS for cleaned and enhanced version already

cbd_data <- data.frame(
  name = c("Raffles Place", "Tanjong Pagar", "Marina Bay"),
  x = c(30214.5, 29527.8, 31324.7),
  y = c(29414.2, 28645.1, 29877.3)
)
```

#### Hawker Centres

```{r}
#| code-summary: Import Hawker Centres data
#| eval: false # save RDS for cleaned and enhanced version already

hawkerCenters = st_read(
  "data/geospatial/HawkerCentresGEOJSON.geojson") %>%
  st_transform(crs= 3414)
```

#### Parks

```{r}
#| code-summary: Import Parks data
#| eval: false # save RDS for cleaned and enhanced version already

parks = st_read(
  "data/geospatial/Parks.geojson") %>%
  st_transform(crs= 3414)
```

#### Supermarkets

```{r}
#| code-summary: Import Supermarkets data
#| eval: false # save RDS for cleaned and enhanced version already

supermarkets = st_read(
  "data/geospatial/SupermarketsGEOJSON.geojson") %>%
  st_transform(crs= 3414)
```

#### Eldercare Centres

```{r}
#| code-summary: Import Eldercare Centres data
#| eval: false # save RDS for cleaned and enhanced version already

eldercare <- st_read(
  dsn = "data/geospatial", 
  layer = "ELDERCARE"
) %>% 
  st_transform(crs = 3414)
```

#### Preschools

```{r}
#| code-summary: Import Preschool data
#| eval: false # save RDS for cleaned and enhanced version already

preschool = st_read(
  "data/geospatial/PreSchoolsLocation.kml") %>%
  st_transform(crs= 3414)
```

#### Kindergartens

```{r}
#| code-summary: Import Kindergartens data
#| eval: false # save RDS for cleaned and enhanced version already

kindergartens = st_read(
  "data/geospatial/Kindergartens.geojson") %>%
  st_transform(crs= 3414)
```

#### Childcare Centres

```{r}
#| code-summary: Import Childcare Centres data
#| eval: false # save RDS for cleaned and enhanced version already

childCareServices = st_read(
  "data/geospatial/ChildCareServices.geojson") %>%
  st_transform(crs= 3414)
```

#### Bus Stops

```{r}
#| code-summary: Import Bus Stops data
#| eval: false # save RDS for cleaned and enhanced version already

busStops <- st_read(
  dsn = "data/geospatial", 
  layer = "BusStop"
) %>% 
  st_transform(crs = 3414)
```

```{r}
#| code-summary: Convert to sf object with SVY21 CRS (EPSG:3414)
#| eval: false # save RDS for cleaned and enhanced version already

cbd_sf <- st_as_sf(cbd_data, 
                   coords = c("x", "y"), 
                   crs = 3414)

# View the result
print(cbd_sf)
```
:::

### 2.4 Proximity Analysis

```{r}
#| code-summary: Count number of preschools, kindergartens, childcare services, and bus stops
#| eval: false # save RDS for cleaned and enhanced version already

# Preschool
within_350m <- st_is_within_distance(
  HDBresale_sf, preschool, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(N_350M_PRESCHOOL = map_int(within_350m, length))

# Kindergartens
within_350m <- st_is_within_distance(
  HDBresale_sf, kindergartens, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(N_350M_KINDERG = map_int(within_350m, length))

# Child care
within_350m <- st_is_within_distance(
  HDBresale_sf, childCareServices, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(N_350M_CHILDCARE = map_int(within_350m, length))

# Bus Stop
within_350m <- st_is_within_distance(
  HDBresale_sf, busStops, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(N_350M_BUSSTOP = map_int(within_350m, length))

```

```{r}
#| code-summary: Create function for nearest distance calculation
#| eval: false # save RDS for cleaned and enhanced version already

calc_nearest_dist <- function(from_geom, to_sf) {
  as.numeric(
    st_distance(
      from_geom,
      to_sf[st_nearest_feature(from_geom, to_sf), ],
      by_element = TRUE
    )
  )
}
```

```{r}
#| code-summary: Apply function to get the nearest distance to the facilities
#| eval: false # save RDS for cleaned and enhanced version already

hdb_model <- HDBresale_sf %>%
  mutate(
    PROX_CBD           = calc_nearest_dist(geometry, cbd_sf),
    PROX_HAWKER        = calc_nearest_dist(geometry, hawkerCenters),
    PROX_PARK          = calc_nearest_dist(geometry, parks),
    PROX_SUPERMARKET   = calc_nearest_dist(geometry, supermarkets),
    PROX_ELDERCARE     = calc_nearest_dist(geometry, eldercare)
  )

hdb_model
```

```{r}
#| code-summary: Save rds for hdb_model
#| eval: false
#| echo: false

write_rds(hdb_model, "data/rds/hdb_model.rds")
```

```{r}
#| echo: false
#| output: false

hdb_model <- read_rds("data/rds/hdb_model.rds")
```

### 2.5 Aspatial Data Cleaning

::: panel-tabset
#### Step 1: Initial Data Inspection

```{r}
#| code-summary: Check structure

glimpse(hdb_model)
```

```{r}
#| code-summary: Check dimensions

cat("Dataset dimensions:", nrow(hdb_model), "rows ×", ncol(hdb_model), "columns\n")
```

```{r}
#| code-summary: Check for missing values

missing_summary <- hdb_model %>%
  st_drop_geometry() %>%  # Remove geometry for cleaner summary
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") %>%
  filter(missing_count > 0) %>%
  arrange(desc(missing_count))

if(nrow(missing_summary) > 0) {
  cat("\n⚠ Missing values detected:\n")
  print(missing_summary)
} else {
  cat("\n✓ No missing values detected\n")
}
```

#### Step 2: Clean Structural Variables

Purpose: Prepare structural variables for modeling

-   Convert storey_range to numeric midpoint (e.g., "13 TO 15" → 14)
-   Convert remaining_lease to numeric years
-   Extract month as date object

```{r}
#| code-summary: Clean and transform structural variables

hdb_clean <- hdb_model %>%
  mutate(
    # Convert storey_range (e.g., "13 TO 15" → 14)
    storey_midpoint = as.numeric(str_extract(storey_range, "^\\d+")) + 1,
    
    # Convert remaining_lease to numeric years
    # Original format like "56 years 03 months"
    remaining_lease_years = case_when(
      str_detect(remaining_lease, "years") & str_detect(remaining_lease, "months") ~
        as.numeric(str_extract(remaining_lease, "^\\d+")) + 
        as.numeric(str_extract(remaining_lease, "(?<=years )\\d+")) / 12,
      str_detect(remaining_lease, "years") ~
        as.numeric(str_extract(remaining_lease, "^\\d+")),
      TRUE ~ NA_real_
    ),
    
    # Convert month to date
    month_date = as.Date(paste0(month, "-01")),
    
    # Extract year and month for temporal analysis
    year = year(month_date),
    month_num = month(month_date)
  )

# Check transformations
cat("✓ Storey midpoint range:", 
    min(hdb_clean$storey_midpoint, na.rm = TRUE), "to", 
    max(hdb_clean$storey_midpoint, na.rm = TRUE), "\n",

"✓ Remaining lease range:", 
    round(min(hdb_clean$remaining_lease_years, na.rm = TRUE), 1), "to", 
    round(max(hdb_clean$remaining_lease_years, na.rm = TRUE), 1), "years\n")
```

#### Step 3: Check for Outliers

```{r}
#| code-summary: Identify outliers in key variables of `resale_price` and `resale_price`

# Function to detect outliers using IQR method
detect_outliers <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  return(x < lower_bound | x > upper_bound)
}
```

```{r}
#| code-summary: Check outliers in resale_price

outlier_price <- detect_outliers(hdb_clean$resale_price)
cat("Outliers in resale_price:", sum(outlier_price), 
    "(", round(sum(outlier_price)/nrow(hdb_clean)*100, 2), "%)\n")
```

```{r}
#| code-summary: Check outliers in floor_area_sqm

outlier_area <- detect_outliers(hdb_clean$floor_area_sqm)
cat("Outliers in floor_area_sqm:", sum(outlier_area), 
    "(", round(sum(outlier_area)/nrow(hdb_clean)*100, 2), "%)\n")
```

```{r}
#| code-summary: Visual inspection with boxplots

par(mfrow = c(1, 2))
boxplot(hdb_clean$resale_price, main = "Resale Price", ylab = "Price (SGD)")
boxplot(hdb_clean$floor_area_sqm, main = "Floor Area", ylab = "Square Meters")
par(mfrow = c(1, 1))
```

For now, we don't remove outliers yet. We just flag them using 2 additional columns of `outlier_price` and `outlier_area`, and keep in the dataset. Those detected will be report "TRUE" in `outlier_price` or `outlier_area` columns.

```{r}
#| code-summary: Flag outliers

hdb_clean <- hdb_clean %>%
  mutate(
    outlier_price = detect_outliers(resale_price),
    outlier_area = detect_outliers(floor_area_sqm)
  )

hdb_clean
```
:::

```{r}
#| code-summary: Save rds for hdb_clean
#| eval: false
#| echo: false

write_rds(hdb_clean, "data/rds/hdb_clean.rds")
```

```{r}
#| echo: false
#| output: false

hdb_clean <- read_rds("data/rds/hdb_clean.rds")
```

## 3. Exploratory Spatial Data Analysis

### 3.1 Descriptive Statistics

#### 3.1.1 Summary of Variables

::: panel-tabset

#### Structural variables

```{r}
#| code-summary: Summary statistics for structural variables

structural_vars <- c("resale_price", "floor_area_sqm", "storey_midpoint", 
                     "remaining_lease_years")

structural_summary <- hdb_clean %>%
  st_drop_geometry() %>%
  select(all_of(structural_vars)) %>%
  summarise(across(everything(), 
                   list(mean = ~mean(., na.rm = TRUE),
                        median = ~median(., na.rm = TRUE),
                        sd = ~sd(., na.rm = TRUE),
                        min = ~min(., na.rm = TRUE),
                        max = ~max(., na.rm = TRUE)),
                   .names = "{.col}_{.fn}")) %>%
  pivot_longer(everything(), names_to = "stat", values_to = "value") %>%
  separate(stat, into = c("variable", "statistic"), sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = statistic, values_from = value)

knitr::kable(structural_summary, 
             digits = 2,
             caption = "Descriptive Statistics: Structural Variables",
             col.names = c("Variable", "Mean", "Median", "SD", "Min", "Max"))
```


```{r}
#| code-summary: Visualize distributions of key variables
#| fig-width: 12
#| fig-height: 8

# Create histogram panels for key variables

# Resale price distribution
p1 <- ggplot(hdb_clean, aes(x = resale_price)) +
  geom_histogram(bins = 50, fill = "#440154", alpha = 0.7) +
  geom_vline(aes(xintercept = median(resale_price)), 
             color = "red", linetype = "dashed", linewidth = 1) +
  labs(title = "Distribution of Resale Prices",
       x = "Resale Price (SGD)", y = "Frequency") +
  scale_x_continuous(labels = scales::comma) +
  theme_minimal()

# Floor area distribution
p2 <- ggplot(hdb_clean, aes(x = floor_area_sqm)) +
  geom_histogram(bins = 30, fill = "#31688e", alpha = 0.7) +
  geom_vline(aes(xintercept = median(floor_area_sqm)), 
             color = "red", linetype = "dashed", linewidth = 1) +
  labs(title = "Distribution of Floor Area",
       x = "Floor Area (sqm)", y = "Frequency") +
  theme_minimal()

# Remaining lease distribution
p3 <- ggplot(hdb_clean, aes(x = remaining_lease_years)) +
  geom_histogram(bins = 40, fill = "#35b779", alpha = 0.7) +
  geom_vline(aes(xintercept = median(remaining_lease_years)), 
             color = "red", linetype = "dashed", linewidth = 1) +
  labs(title = "Distribution of Remaining Lease",
       x = "Remaining Lease (years)", y = "Frequency") +
  theme_minimal()

# Storey distribution
p4 <- ggplot(hdb_clean, aes(x = storey_midpoint)) +
  geom_histogram(bins = 20, fill = "#fde724", alpha = 0.7) +
  geom_vline(aes(xintercept = median(storey_midpoint)), 
             color = "red", linetype = "dashed", linewidth = 1) +
  labs(title = "Distribution of Storey Level",
       x = "Storey Midpoint", y = "Frequency") +
  theme_minimal()

# Combine plots
(p1 + p2) / (p3 + p4) +
  plot_annotation(
    title = "Distribution of Structural Variables",
    subtitle = "Red dashed line indicates median value"
  )
```

#### COUNT variables

```{r}
#| code-summary: Descriptive statistics for count variables

# --- Location Variables: Counts (within buffer) ---
count_vars <- names(hdb_clean)[str_detect(names(hdb_clean), "^N_")]

count_summary <- hdb_clean %>%
  st_drop_geometry() %>%
  select(all_of(count_vars)) %>%
  summarise(across(everything(), 
                   list(mean = ~mean(., na.rm = TRUE),
                        median = ~median(., na.rm = TRUE),
                        sd = ~sd(., na.rm = TRUE),
                        min = ~min(., na.rm = TRUE),
                        max = ~max(., na.rm = TRUE)),
                   .names = "{.col}_{.fn}")) %>%
  pivot_longer(everything(), names_to = "stat", values_to = "value") %>%
  separate(stat, into = c("variable", "statistic"), sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = statistic, values_from = value)

knitr::kable(count_summary, 
             digits = 1,
             caption = "Table 3: Descriptive Statistics - Count Variables (facilities within buffer)",
             col.names = c("Variable", "Mean", "Median", "SD", "Min", "Max"))
```


```{r}
#| code-summary: Visualization of Distribution of count variables  
#| fig-width: 12
#| fig-height: 10

# Count variables distributions
c1 <- ggplot(hdb_clean, aes(x = N_350M_PRESCHOOL)) +
  geom_bar(fill = "#440154", alpha = 0.7) +
  labs(title = "Preschools within 350m", x = "Count", y = "Frequency") +
  theme_minimal()

c2 <- ggplot(hdb_clean, aes(x = N_350M_KINDERG)) +
  geom_bar(fill = "#31688e", alpha = 0.7) +
  labs(title = "Kindergartens within 350m", x = "Count", y = "Frequency") +
  theme_minimal()

c3 <- ggplot(hdb_clean, aes(x = N_350M_CHILDCARE)) +
  geom_bar(fill = "#35b779", alpha = 0.7) +
  labs(title = "Childcare Centres within 350m", x = "Count", y = "Frequency") +
  theme_minimal()

c4 <- ggplot(hdb_clean, aes(x = N_350M_BUSSTOP)) +
  geom_bar(fill = "#fde724", alpha = 0.7) +
  labs(title = "Bus Stops within 350m", x = "Count", y = "Frequency") +
  theme_minimal()

(c1 + c2) / (c3 + c4) +
  plot_annotation(
    title = "Figure 2: Distribution of Count Variables",
    subtitle = "Number of facilities within buffer zones"
  )
```


#### Proximity variables

```{r}
#| code-summary: Summary statistics for Proximity variables summary (in meters)

prox_vars <- hdb_clean %>%
  st_drop_geometry() %>%
  select(starts_with("PROX_")) %>%
  names()

location_summary <- hdb_clean %>%
  st_drop_geometry() %>%
  select(all_of(prox_vars)) %>%
  summarise(across(everything(), 
                   list(mean = ~mean(., na.rm = TRUE),
                        median = ~median(., na.rm = TRUE),
                        sd = ~sd(., na.rm = TRUE),
                        min = ~min(., na.rm = TRUE),
                        max = ~max(., na.rm = TRUE)),
                   .names = "{.col}_{.fn}")) %>%
  pivot_longer(everything(), names_to = "stat", values_to = "value") %>%
  separate(stat, into = c("variable", "statistic"), sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = statistic, values_from = value)

knitr::kable(location_summary, 
             digits = 2,
             caption = "Descriptive Statistics: Location Variables (Proximity in meters)",
             col.names = c("Variable", "Mean", "Median", "SD", "Min", "Max"))
```

:::

#### 3.1.2 Exploration on different dimensions

::: panel-tabset

#### Temporal Distribution

```{r}
#| code-summary: Transaction volume and prices over time
#| fig-width: 12
#| fig-height: 6

# Transaction volume by month
t1 <- hdb_clean %>%
  st_drop_geometry() %>%
  count(month_date) %>%
  ggplot(aes(x = month_date, y = n)) +
  geom_line(color = "#440154", linewidth = 1) +
  geom_point(color = "#440154", size = 2) +
  labs(title = "Transaction Volume Over Time",
       x = "Month", y = "Number of Transactions") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Median price by month
t2 <- hdb_clean %>%
  st_drop_geometry() %>%
  group_by(month_date) %>%
  summarise(
    median_price = median(resale_price),
    mean_price = mean(resale_price),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = month_date)) +
  geom_line(aes(y = median_price, color = "Median"), linewidth = 1) +
  geom_line(aes(y = mean_price, color = "Mean"), linewidth = 1) +
  geom_point(aes(y = median_price, color = "Median"), size = 2) +
  geom_point(aes(y = mean_price, color = "Mean"), size = 2) +
  labs(title = "Resale Price Trends Over Time",
       x = "Month", y = "Resale Price (SGD)", color = "Statistic") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +
  scale_color_manual(values = c("Median" = "#440154", "Mean" = "#fde724")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

t1 / t2
```

#### Summary by Town

```{r}
#| code-summary: Resale prices by town

# Calculate summary statistics by town
town_summary <- hdb_clean %>%
  st_drop_geometry() %>%
  group_by(town) %>%
  summarise(
    n_transactions = n(),
    median_price = median(resale_price),
    mean_price = mean(resale_price),
    min_price = min(resale_price),
    max_price = max(resale_price),
    .groups = "drop"
  ) %>%
  arrange(desc(median_price))

# Display table
knitr::kable(town_summary, 
             digits = 0,
             caption = "Resale Price Summary by Town (4-Room Flats)",
             col.names = c("Town", "Transactions", "Median Price", 
                          "Mean Price", "Min Price", "Max Price"),
             format.args = list(big.mark = ","))

# Top 10 most expensive towns
top10_towns <- town_summary %>%
  slice_max(median_price, n = 10) %>%
  ggplot(aes(x = reorder(town, median_price), y = median_price)) +
  geom_col(fill = "#440154", alpha = 0.8) +
  geom_text(aes(label = scales::comma(median_price)), 
            hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "Top 10 Towns by Median Resale Price",
       subtitle = "4-Room Flats (Jan-Sep 2025)",
       x = "Town", y = "Median Resale Price (SGD)") +
  scale_y_continuous(labels = scales::comma, 
                     expand = expansion(mult = c(0, 0.1))) +
  theme_minimal()

print(top10_towns)
```

#### Summary by Flat Model

```{r}
#| code-summary: Resale prices by flat model

# Summary by flat_model
model_summary <- hdb_clean %>%
  st_drop_geometry() %>%
  group_by(flat_model) %>%
  summarise(
    n_transactions = n(),
    median_price = median(resale_price),
    pct_of_total = n() / nrow(hdb_clean) * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(n_transactions))

knitr::kable(model_summary, 
             digits = 1,
             caption = "Resale Price Summary by Flat Model",
             col.names = c("Flat Model", "Transactions", 
                          "Median Price", "% of Total"),
             format.args = list(big.mark = ","))
```
:::

#### 3.1.3 Bivariate Relationships

```{r}
#| code-summary: Key relationships between price and predictors
#| fig-width: 12
#| fig-height: 10

# Price vs Floor Area
b1 <- ggplot(hdb_clean, aes(x = floor_area_sqm, y = resale_price)) +
  geom_point(alpha = 0.3, color = "#440154") +
  geom_smooth(method = "lm", color = "#fde724", linewidth = 1.5) +
  labs(title = "Price vs Floor Area",
       x = "Floor Area (sqm)", y = "Resale Price (SGD)") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()

# Price vs Remaining Lease
b2 <- ggplot(hdb_clean, aes(x = remaining_lease_years, y = resale_price)) +
  geom_point(alpha = 0.3, color = "#31688e") +
  geom_smooth(method = "lm", color = "#fde724", linewidth = 1.5) +
  labs(title = "Price vs Remaining Lease",
       x = "Remaining Lease (years)", y = "Resale Price (SGD)") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()

# Price vs Storey
b3 <- ggplot(hdb_clean, aes(x = storey_midpoint, y = resale_price)) +
  geom_point(alpha = 0.3, color = "#35b779") +
  geom_smooth(method = "lm", color = "#fde724", linewidth = 1.5) +
  labs(title = "Price vs Storey Level",
       x = "Storey Midpoint", y = "Resale Price (SGD)") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()

# Price vs Distance to CBD
b4 <- ggplot(hdb_clean, aes(x = PROX_CBD, y = resale_price)) +
  geom_point(alpha = 0.3, color = "#fde724") +
  geom_smooth(method = "lm", color = "#440154", linewidth = 1.5) +
  labs(title = "Price vs Distance to CBD",
       x = "Distance to CBD (meters)", y = "Resale Price (SGD)") +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()

(b1 + b2) / (b3 + b4) +
  plot_annotation(
    title = "Figure 6: Bivariate Relationships with Resale Price",
    subtitle = "Linear trend lines shown in yellow/purple"
  )
```

Let's specifically dive in to the predictor of Distance to CBD:

```{r}
#| code-summary: Relationship between price and CBD proximity
#| fig-width: 12
#| fig-height: 8

# Create distance bands
hdb_clean <- hdb_clean %>%
  mutate(
    cbd_band = cut(PROX_CBD,
                   breaks = c(0, 5000, 10000, 15000, 20000, 25000, Inf),
                   labels = c("0-5km", "5-10km", "10-15km", 
                             "15-20km", "20-25km", "> 25km"))
  )

# Boxplot: Price by distance band
p1 <- ggplot(hdb_clean, aes(x = cbd_band, y = resale_price, fill = cbd_band)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_viridis_d(option = "plasma") +
  labs(title = "Resale Price by Distance to CBD",
       x = "Distance Band", y = "Resale Price (SGD)") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

# Scatterplot with smooth
p2 <- ggplot(hdb_clean, aes(x = PROX_CBD, y = resale_price)) +
  geom_point(alpha = 0.2, color = "#440154") +
  geom_smooth(method = "loess", color = "#fde724", linewidth = 1.5, se = TRUE) +
  labs(title = "Price Decay with Distance from CBD",
       x = "Distance to CBD (meters)", y = "Resale Price (SGD)") +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()

p1 + p2 +
  plot_annotation(
    title = "Figure 8: Relationship Between Resale Price and CBD Proximity",
    subtitle = "Clear price premium for central locations"
  )
```

Distance to the CBD exhibits a clear inverse relationship with resale prices, demonstrating the persistent importance of accessibility in Singapore's housing market. Properties within 5km of the CBD command median prices exceeding \$750,000, while those beyond 25km average below \$600,000—a premium gap of over 25%. The LOESS curve reveals a non-linear decay pattern: prices drop steeply within the first 10km from the CBD, then gradually flatten in peripheral areas, suggesting diminishing marginal effects of distance beyond a certain threshold. High variance in the 10-15km band reflects diverse submarkets where both mature estates (Ang Mo Kio, Bishan) and developing towns coexist. Interestingly, some peripheral transactions (\>20km) still achieve high prices, likely due to compensating factors like larger floor areas, higher floors, or proximity to regional centers (Tampines, Jurong). This spatial price gradient underscores why geographically weighted regression is necessary—the impact of CBD distance on price likely varies across locations, with stronger effects in central areas where marginal improvements in accessibility are more valuable.

### 3.2 Spatial Distribution of Resale Prices

::: panel-tabset
#### Basic Point Map of Transactions

```{r}
#| code-summary: Point map showing all HDB resale transactions
#| fig-width: 10
#| fig-height: 12

tmap_mode("plot")  # Static map mode

# Create basic point map
tm_shape(singapore_boundary) +
  tm_borders(col = "grey50", lwd = 1) +
  tm_fill(col = "grey95") +
tm_shape(hdb_clean) +
  tm_dots(col = "resale_price",
          palette = "YlOrRd",
          size = 0.05,
          alpha = 0.6,
          title = "Resale Price (SGD)",
          breaks = c(0, 500000, 600000, 700000, 800000, 900000, 1500000),
          labels = c("< $500k", "$500k-$600k", "$600k-$700k", 
                     "$700k-$800k", "$800k-$900k", "> $900k")) +
  tm_layout(
    main.title = "Map 1: Spatial Distribution of HDB 4-Room Resale Prices",
    main.title.size = 1.2,
    main.title.position = "center",
    legend.outside = TRUE,
    legend.outside.position = "right",
    legend.title.size = 1,
    legend.text.size = 0.7,
    frame = FALSE
  ) +
  tm_compass(position = c("left", "top"), size = 2)
```

The spatial distribution reveals distinct price clustering across Singapore, with higher-priced transactions (orange to red) concentrated in central and mature estates. The western (Clementi, Bukit Timah) and eastern (Marine Parade, Bedok) regions show elevated prices, while peripheral areas like Woodlands and Yishun display lower price points (yellow). This pattern reflects Singapore's geographical price gradient radiating from the CBD, where accessibility to employment centers, established infrastructure, and proximity to amenities command premiums. Notable price outliers appear in mature estates like Queenstown and Toa Payoh, likely reflecting recently renovated units or prime locations within these towns. The dense clustering of transactions in Tampines, Jurong West, and Ang Mo Kio indicates active resale markets in these high-density HDB towns. This spatial heterogeneity validates the need for geographically weighted regression, as the relationship between price and location factors clearly varies across Singapore's urban landscape.

#### Hexbin Aggregation Map

```{r}
#| code-summary: Hexagonal binning to show transaction density and median prices
#| fig-width: 10
#| fig-height: 12

# Create hexagonal grid (1km cell size)
hex_grid <- st_make_grid(
  singapore_boundary,
  cellsize = 1000,  # 1km hexagons
  square = FALSE,   # hexagonal
  what = "polygons"
) %>%
  st_sf() %>%
  st_intersection(singapore_boundary)

# Spatial join: count transactions and calculate median price per hexagon
hex_summary <- hex_grid %>%
  st_join(hdb_clean) %>%
  group_by(geometry) %>%
  summarise(
    n_transactions = sum(!is.na(resale_price)),
    median_price = median(resale_price, na.rm = TRUE),
    mean_price = mean(resale_price, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_transactions > 0)  # Keep only hexagons with transactions
```

```{r}
#| code-summary: Hexbin Aggregation Map
#| fig-width: 10
#| fig-height: 12

# Map: Transaction Density
map_density <- tm_shape(singapore_boundary) +
  tm_borders(col = "grey50", lwd = 1) +
tm_shape(hex_summary) +
  tm_fill(col = "n_transactions",
          palette = "brewer.blues",
          style = "jenks",
          n = 6,
          title = "Number of\nTransactions",
          alpha = 0.8) +
  tm_borders(col = "white", lwd = 0.5) +
  tm_layout(
    main.title = "Map 2a: Transaction Density (Hexagonal Binning)",
    main.title.size = 1.2,
    legend.outside = TRUE,
    legend.outside.position = "right",
    frame = FALSE
  ) +
  tm_compass(position = c("left", "top"), size = 2) 

# Map - Median Price by Hexagon
map_price_hex <- tm_shape(singapore_boundary) +
  tm_borders(col = "grey50", lwd = 1) +
tm_shape(hex_summary) +
  tm_fill(col = "median_price",
          palette = "RdYlGn",
          style = "jenks",
          n = 6,
          title = "Median Price\n(SGD)",
          alpha = 0.8,
          labels = c("< $550k", "$550k-$600k", "$600k-$650k", 
                     "$650k-$700k", "$700k-$750k", "> $750k")) +
  tm_borders(col = "white", lwd = 0.5) +
  tm_layout(
    main.title = "Map 2b: Median Resale Price (Hexagonal Aggregation)",
    main.title.size = 1.2,
    legend.outside = TRUE,
    legend.outside.position = "right",
    frame = FALSE
  ) +
  tm_compass(position = c("left", "top"), size = 2) 

# Display both maps
tmap_arrange(map_density, map_price_hex, ncol = 1)
```

Hexagonal binning reveals clear spatial patterns in both transaction activity and pricing. The transaction density map (2a) shows the most active resale markets in mature estates like Tampines, Ang Mo Kio, and Jurong West, where large HDB populations generate consistent turnover. Lower activity appears in the CBD fringe and newer peripheral estates. The median price map (2b) demonstrates a pronounced core-periphery gradient, with central hexagons (Bukit Timah, Queenstown, Marine Parade) commanding premiums above \$700,000, while outer regions (Woodlands, Sembawang, Jurong) show prices below \$600,000. Interestingly, some high-transaction areas display moderate prices, suggesting affordability-driven demand in towns like Yishun and Woodlands. Conversely, low-transaction but high-price hexagons appear near the CBD, reflecting limited 4-room stock in central districts. This aggregated view smooths individual outliers and clarifies macro-level spatial trends, supporting the hypothesis that location significantly influences HDB resale values across Singapore's geography.

#### Price Map by Planning Area/Town

```{r}
#| code-summary: Calculate median price by town

town_prices <- hdb_clean %>%
  st_drop_geometry() %>%
  group_by(town) %>%
  summarise(
    median_price = median(resale_price),
    n_transactions = n(),
    .groups = "drop"
  )

# Join planning area with price data
planning_areas_price <- planning_areas %>%
  left_join(town_prices, by = "town")
```

```{r}
#| code-summary: Choropleth map to for Median resale price aggregated by town
#| fig-width: 10
#| fig-height: 12

tm_shape(planning_areas_price) +
  tm_fill(col = "median_price",
          palette = "RdYlGn",
          style = "quantile",
          n = 7,
          title = "Median Price (SGD)",
          textNA = "No transactions",
          colorNA = "grey90",
          alpha = 0.8) +
  tm_borders(col = "white", lwd = 1) +
  tm_text("town", size = 0.4, col = "black", alpha = 0.7) +
  tm_layout(
    main.title = "Map 3: Median Resale Price by Planning Area",
    main.title.size = 1.2,
    legend.outside = TRUE,
    legend.outside.position = "right",
    frame = FALSE
  ) +
  tm_compass(position = c("left", "top"), size = 2) 
```

The planning area choropleth provides a clear administrative view of price disparities across Singapore's towns. Central planning areas (Bukit Timah, Novena, Queenstown) display the highest median prices, benefiting from CBD proximity, mature infrastructure, and excellent connectivity. Eastern towns like Bedok and Marine Parade also command premiums due to coastal amenities and established neighborhoods. Mid-tier pricing appears in towns like Ang Mo Kio, Bishan, and Clementi—mature estates with good amenities but located further from the core. The lowest prices concentrate in peripheral northern (Woodlands, Sembawang, Yishun) and western (Choa Chu Kang, Bukit Batok) towns, where longer commute times and developing infrastructure reduce demand premiums. Some towns show no transactions (grey areas), indicating limited 4-room stock or very low turnover. This administrative aggregation reveals policy-relevant insights: price gaps between mature and non-mature estates persist, suggesting continued importance of geographical accessibility in Singapore's housing market despite government efforts to improve connectivity across all regions.
:::

### 3.3 Spatial Autocorrelation

::: panel-tabset
#### Global Moran's I test

```{r}
#| code-summary: Test for spatial clustering using Moran's I

# Create spatial weights matrix (k-nearest neighbors) using k=8 nearest neighbors
coords <- st_coordinates(st_centroid(hdb_clean))
knn_weights <- knn2nb(knearneigh(coords, k = 8))
knn_listw <- nb2listw(knn_weights, style = "W")

# Global Moran's I test
moran_test <- moran.test(hdb_clean$resale_price, knn_listw)
```

```{r}
#| code-summary: GLOBAL MORAN'S I TEST FOR RESALE PRICE

cat("Moran's I statistic:", round(moran_test$estimate[1], 4), "\n",
"Expected value:", round(moran_test$estimate[2], 4), "\n",
"Variance:", round(moran_test$estimate[3], 6), "\n",
"Standard deviate:", round(moran_test$statistic, 4), "\n",
"P-value:", format.pval(moran_test$p.value, eps = 0.001), "\n\n")

if(moran_test$p.value < 0.001) {
  cat("Interpretation: Strong evidence of positive spatial autocorrelation\n",
  "→ High prices cluster near high prices, low near low\n",
  "→ Spatial non-stationarity present\n",
  "→ GWR is justified over OLS\n")
} else if(moran_test$p.value < 0.05) {
  cat("Interpretation: Evidence of spatial autocorrelation\n")
} else {
  cat("Interpretation: No significant spatial autocorrelation\n")
}
```

```{r}
#| code-summary: Moran's I scatterplot showing spatial lag relationship
#| fig-width: 10
#| fig-height: 8

# Create Moran's I scatterplot
moran_plot <- moran.plot(hdb_clean$resale_price, knn_listw,
                         xlab = "Resale Price (standardized)",
                         ylab = "Spatially Lagged Price (standardized)",
                         main = "Figure 7: Moran's I Scatterplot")

# Add interpretation text
mtext(paste0("Moran's I = ", round(moran_test$estimate[1], 4), 
             ", p < 0.001"), 
      side = 3, line = 0.5, cex = 0.9)
```

The Global Moran's I test confirms significant positive spatial autocorrelation (I = \[value\], p \< 0.001), indicating that HDB resale prices cluster spatially across Singapore. The Moran scatterplot reveals the relationship between each transaction's price and the average price of its spatial neighbors. Points in the upper-right quadrant (High-High) represent expensive flats surrounded by expensive neighbors, typical of central and mature estates. The lower-left quadrant (Low-Low) captures affordable flats in peripheral neighborhoods. Observations in the upper-left (Low-High) and lower-right (High-Low) quadrants are spatial outliers—affordable flats in expensive neighborhoods or vice versa—often reflecting unique property characteristics or localized market dynamics. The strong positive relationship (clustered around the regression line) demonstrates that a property's price is not independent of nearby prices, violating the independence assumption of ordinary least squares regression. This spatial dependency justifies the use of geographically weighted regression, which explicitly accounts for spatial non-stationarity by allowing model coefficients to vary across geographic space.

#### Local Indicators of Spatial Association (LISA)

```{r}
#| code-summary: Identify local clusters and outliers using LISA

# Calculate Local Moran's I
local_moran <- localmoran(hdb_clean$resale_price, knn_listw)

# Add LISA results to dataset
hdb_clean <- hdb_clean %>%
  mutate(
    local_I = local_moran[, 1],  # Local Moran's I statistic
    local_pval = local_moran[, 5],  # P-value
    
    # Standardized price
    z_price = scale(resale_price)[, 1],
    
    # Spatially lagged standardized price
    lag_z_price = lag.listw(knn_listw, z_price),
    
    # LISA quadrant classification
    lisa_category = case_when(
      local_pval > 0.05 ~ "Not Significant",
      z_price > 0 & lag_z_price > 0 ~ "High-High",
      z_price < 0 & lag_z_price < 0 ~ "Low-Low",
      z_price < 0 & lag_z_price > 0 ~ "Low-High (outlier)",
      z_price > 0 & lag_z_price < 0 ~ "High-Low (outlier)",
      TRUE ~ "Not Significant"
    ),
    
    # Factor with specific order for legend
    lisa_category = factor(lisa_category, 
                          levels = c("High-High", "Low-Low", 
                                    "High-Low (outlier)", "Low-High (outlier)", 
                                    "Not Significant"))
  )

```

```{r}
#| code-summary: LISA cluster map
#| fig-width: 10
#| fig-height: 12

tm_shape(singapore_boundary) +
  tm_borders(col = "grey50", lwd = 1) +
  tm_fill(col = "grey95") +
tm_shape(hdb_clean) +
  tm_dots(col = "lisa_category",
          palette = c("High-High" = "#d7191c",      # Red
                     "Low-Low" = "#2b83ba",         # Blue
                     "High-Low (outlier)" = "#fdae61",  # Orange
                     "Low-High (outlier)" = "#abd9e9",  # Light blue
                     "Not Significant" = "grey80"),
          size = 0.08,
          alpha = 0.7,
          title = "LISA Category") +
  tm_layout(
    main.title = "Map 4: Local Indicators of Spatial Association (LISA)",
    main.title.size = 1.2,
    main.title.position = "center",
    legend.outside = TRUE,
    legend.outside.position = "right",
    legend.title.size = 1,
    legend.text.size = 0.7,
    frame = FALSE
  ) +
  tm_compass(position = c("left", "top"), size = 2)
  # + tm_scale_bar(position = c("left", "bottom"), width = 0.15)
```

```{r}
#| code-summary: Summary statistics
lisa_summary <- hdb_clean %>%
  st_drop_geometry() %>%
  count(lisa_category) %>%
  mutate(percentage = n / sum(n) * 100)

knitr::kable(lisa_summary,
             digits = 1,
             caption = "Table 6: Distribution of LISA Categories",
             col.names = c("LISA Category", "Count", "Percentage (%)"))
```

The LISA map identifies statistically significant local clusters and spatial outliers in the HDB resale market. High-High clusters (red) concentrate in central and mature estates like Bukit Timah, Queenstown, and Marine Parade, where expensive properties are surrounded by expensive neighbors—representing premium submarkets with sustained high demand. Low-Low clusters (blue) dominate peripheral towns like Woodlands, Yishun, and Choa Chu Kang, indicating consistently affordable neighborhoods with similar pricing dynamics. Spatial outliers are particularly revealing: Low-High outliers (light blue) represent affordable flats in expensive neighborhoods, potentially older units or ground-floor flats in prime locations, while High-Low outliers (orange) capture premium properties in moderate-price areas, possibly reflecting recent renovations or unique features. The large proportion of non-significant observations (grey) suggests that while broad spatial patterns exist, many transactions fall within normal local variation. This local analysis confirms that Singapore's resale market exhibits distinct spatial regimes, supporting the use of geographically weighted regression to capture varying price determinants across these heterogeneous submarkets.
:::

### 3.4 Correlation Structure

```{r}
#| code-summary: Select and prepare variables for correlation analysis

# Select all predictor variables and outcome variable
# Drop geometry and categorical/text variables

correlation_vars <- hdb_clean %>%
  st_drop_geometry() %>%
  select(
    # Outcome variable
    resale_price,
    
    # Structural variables
    floor_area_sqm,
    storey_midpoint,
    remaining_lease_years,
    
    # Proximity variables (all PROX_ columns)
    starts_with("PROX_"),
    
    # Count variables (all N_ columns)
    starts_with("N_")
  )

# Check for any remaining NA values
na_summary <- correlation_vars %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "na_count") %>%
  filter(na_count > 0)

if(nrow(na_summary) > 0) {
  cat("⚠ Warning: Missing values detected:\n")
  print(na_summary)
  cat("\nRemoving rows with missing values...\n")
  correlation_vars <- na.omit(correlation_vars)
} else {
  cat("✓ No missing values detected\n")
}
```

```{r}
#| code-summary: Summary on data

cat("Variables for correlation analysis:", ncol(correlation_vars), "\n",
"Observations:", nrow(correlation_vars), "\n")
```

#### 3.4.1 Pearson correlation coefficients

```{r}
#| code-summary: Calculate Pearson correlation coefficients and Visualize full correlation matrix
#| fig-width: 12
#| fig-height: 12

# Calculate correlation matrix
cor_matrix <- cor(correlation_vars, use = "complete.obs", method = "pearson")

# Display correlation with resale_price (sorted by absolute value)
price_correlations <- cor_matrix[, "resale_price"] %>%
  as.data.frame() %>%
  rownames_to_column(var = "variable") %>%
  rename(correlation = ".") %>%
  filter(variable != "resale_price") %>%
  arrange(desc(abs(correlation)))

# Create correlation plot with all variables
corrplot(cor_matrix,
         method = "color",
         type = "upper",
         order = "hclust",  # Hierarchical clustering
         tl.col = "black",
         tl.srt = 45,
         tl.cex = 0.7,
         cl.cex = 0.7,
         addCoef.col = "black",
         number.cex = 0.5,
         col = COL2('RdBu', 200),
         title = "Figure 9: Correlation Matrix - All Variables",
         mar = c(0, 0, 2, 0))
```

```{r}
#| code-summary: Heatmap showing correlations with resale price
#| fig-width: 10
#| fig-height: 8

# Extract correlations with resale_price
price_cor_data <- price_correlations %>%
  mutate(
    abs_cor = abs(correlation),
    variable = factor(variable, levels = variable[order(correlation)])
  )

# Create heatmap-style barplot
ggplot(price_cor_data, aes(x = variable, y = correlation, fill = correlation)) +
  geom_col(alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_hline(yintercept = c(-0.3, 0.3), linetype = "dotted", color = "grey50") +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                       midpoint = 0, limit = c(-1, 1),
                       name = "Correlation") +
  coord_flip() +
  labs(title = "Figure 10: Correlation of Predictors with Resale Price",
       subtitle = "Positive correlations increase price | Negative correlations decrease price",
       x = NULL, y = "Pearson Correlation Coefficient") +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.grid.major.y = element_blank())
```

Distance to CBD exhibits the strongest negative correlation with resale prices (r = -0.612), confirming central location premiums. Storey level (r = 0.542) and remaining lease (r = 0.488) show moderate positive correlations, validating their importance as price determinants. Surprisingly, floor area and proximity to amenities (supermarkets, parks, eldercare) demonstrate weak correlations, suggesting their effects may vary spatially across Singapore's submarkets.

Severe multicollinearity exists between childcare centers and preschools within 350m buffers (r = 0.967), indicating these facilities co-locate spatially. Kindergartens show moderate correlation with preschools (r = 0.656). To address multicollinearity, one childcare-related variable should be retained based on stronger theoretical relevance or predictive power, while others are excluded from modeling.

#### 3.4.2 Variance Inflation Factor (VIF) analysis

While correlation analysis reveals pairwise relationships between predictors, it cannot capture the cumulative multicollinearity effect when multiple variables interact simultaneously in a regression model. Variance Inflation Factor (VIF) analysis should address this limitation by quantifying how much each predictor's variance is inflated due to linear dependencies with all other predictors, enabling more informed variable selection decisions.

```{r}
#| code-summary: Calculate VIF to assess multicollinearity severity

# Prepare data for VIF (need to use actual data, not correlation matrix)
# First, fit a preliminary linear model with all predictors

# Create formula with all predictors
predictor_vars <- names(correlation_vars)[names(correlation_vars) != "resale_price"]
formula_str <- paste("resale_price ~", paste(predictor_vars, collapse = " + "))
prelim_formula <- as.formula(formula_str)

# Fit preliminary model
prelim_model <- lm(prelim_formula, data = correlation_vars)

# Calculate VIF
vif_values <- vif(prelim_model)

# Create VIF dataframe
vif_df <- data.frame(
  variable = names(vif_values),
  VIF = vif_values
) %>%
  arrange(desc(VIF)) %>%
  mutate(
    severity = case_when(
      VIF < 5 ~ "Low (acceptable)",
      VIF < 10 ~ "Moderate (caution)",
      TRUE ~ "High (problematic)"
    )
  )

# VIF visualization
ggplot(vif_df, aes(x = reorder(variable, VIF), y = VIF, fill = severity)) +
  geom_col(alpha = 0.8) +
  geom_hline(yintercept = 5, linetype = "dashed", color = "orange", linewidth = 1) +
  geom_hline(yintercept = 10, linetype = "dashed", color = "red", linewidth = 1) +
  scale_fill_manual(values = c("Low (acceptable)" = "#2166ac",
                               "Moderate (caution)" = "#f4a582",
                               "High (problematic)" = "#b2182b"),
                    name = "Severity") +
  coord_flip() +
  labs(title = "Figure 12: Variance Inflation Factors (VIF)",
       subtitle = "VIF > 10 indicates severe multicollinearity | VIF > 5 suggests caution",
       x = NULL, y = "VIF Value") +
  annotate("text", x = 1, y = 5, label = "Threshold = 5", 
           hjust = -0.1, vjust = -0.5, color = "orange", size = 3) +
  annotate("text", x = 1, y = 10, label = "Threshold = 10", 
           hjust = -0.1, vjust = -2, color = "red", size = 3) +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.grid.major.y = element_blank())
```

VIF diagnostics reveal critical multicollinearity issues exclusively among childcare-related variables: preschools (VIF \> 80) and childcare centers (VIF \~ 30) demonstrate severe variance inflation, while kindergartens show moderate concern (VIF \> 5). All structural variables and proximity measures exhibit VIF values below 2, well within acceptable limits. These results confirm that only early childhood education counts require exclusion, ensuring the final model maintains statistical stability without sacrificing theoretical completeness.

#### 3.4.3 Final Variable Selection for GWR Modeling

Based on correlation analysis, multicollinearity diagnostics, and VIF assessment, the following variables are selected for the geographically weighted regression model:

Variables Included (n = 10)

**1. Outcome Variable:** resale_price

**2. Structural Variables (n = 3):**

-   `floor_area_sqm` (VIF = 1.11, r = -0.052)

-   `storey_midpoint` (VIF = 1.21, r = 0.542)

-   `remaining_lease_years` (VIF = 1.25, r = 0.488)

**3. Location Variables - Proximity (n = 5):**

-   `PROX_CBD` (VIF = 1.71, r = -0.612) \[strongest predictor\]

-   `PROX_ELDERCARE` (VIF = 1.26, r = -0.275)

-   `PROX_HAWKER` (VIF = 1.17, r = -0.274)

-   `PROX_PARK` (VIF = 1.26, r = -0.200)

-   `PROX_SUPERMARKET` (VIF = 1.11, r = -0.000)

**4. Location Variables - Count (n = 1):**

-   `N_350M_BUSSTOP` (VIF = 1.11, r = -0.087)

**Variables to be Excluded:**

*Removed due to severe multicollinearity (VIF \> 10):*

-   `N_350M_PRESCHOOL` (VIF = 80.22, r = -0.053)

-   `N_350M_CHILDCARE` (VIF = 59.05, r = -0.041)

*Removed due to moderate multicollinearity:*

-   `N_350M_KINDERG` (VIF = 6.52, r = -0.019)

`N_350M_BUSSTOP` will serve as a proxy for overall accessibility.

## 4. Model Calibration

### 4.1 OLS Baseline Model

#### 4.1.1 Model Specification

#### 4.1.2 Results & Diagnostics

#### 4.1.3 Spatial Autocorrelation Test

### 4.2 Geographically Weighted Regression

#### 4.2.1 Bandwidth Selection

\[Explain CV vs AIC choice\]\

#### 4.2.2 Model Results

#### 4.2.3 Model Comparison

## 5. Spatial Patterns & Interpretation

### 5.1 Overview of Local Variations

### 5.2 Key Findings by Variable

\[Subsection for each important coefficient\]\

### 5.3 Policy & Market Implications

## 6. Limitations & Conclusions

### 6.1 Study Limitations

\[Data period, 4-room only, modifiable areal unit problem, etc.\]\

### 6.2 Conclusions

### 6.3 Future Research Directions

## Appendix

### References
