---
title: "Take-home Exercise 3 | Modelling HDB Resale Prices with Geographically Weighted Methods"
subtitle: "The objective of this take-home exercise is to develop a spatially informed model to determine factors affecting or to predict HDB resale prices in Singapore, making use of appropriate geospatial analytics methods."
date: "Nov 13, 2025"
date-modified: "last-modified"
author: Linh Nguyen
execute:
  echo: true
  eval: true
  freeze: auto
  message: false
  warning: false
  error: false
format:
  html:
    code-link: true       # Clickable function links
    toc: true             # Table of contents
    code-fold: true       # Foldable code
    code-tools: true      # Code tools menu
number-sections: false
number-offset: 1
editor: visual
---

## 1. Introduction

### 1.1 Background & Motivation

### 1.2 Research Objectives

## 2. Data & Methodology

### 2.1 Import R packages

```{r}
pacman::p_load(sf, spdep, GWmodel, tidyverse, lubridate, tmap, ggplot2, corrplot, httr, jsonlite)
```

### 2.2 Data Sources

|  |  |  |  |
|-----------------|------------------|--------------------|-----------------|
| **Dataset Name** | **Description** | **Use Case** | **Format** |
| **HDB Resale Flat Prices** | Transaction records of HDB resale flats. Filtered for 4-room flats only. | Primary outcome variable (`resale_price`) and source of structural predictors (`floor_area_sqm`, `remaining_lease_years`, `storey_midpoint`) | CSV |
| **Singapore Boundary** | A polygon feature data providing information of URA 2014 Master Plan Planning Subzone boundary data | Provides spatial context for mapping. Used as basemap reference in visualizations. | SHP (point geometry) |
| **Hawker Centres** | Point locations of hawker centres, representing affordable dining options and community gathering spaces. | Calculate proximity to nearest hawker centre (`dist_hawker`) | GeoJSON |
| **Parks** | Point or polygon locations of national parks, regional parks, and neighborhood parks managed by NParks. | Calculate proximity to nearest park (`dist_park`) | GeoJSON |
| **Supermarkets** | Point locations of major supermarket chains across Singapore. | Calculate proximity to nearest supermarket (`dist_supermarket`) | GeoJSON |
| **Eldercare Centres** | Point locations of eldercare facilities including senior activity centres, senior care centres, and nursing homes. | Calculate proximity to nearest eldercare facility (`dist_eldercare`), relevant for multi-generational households and aging population considerations. | SHP (point geometry) |
| **Preschools** | Point locations of registered preschools across Singapore, including both public and private institutions. | Count number of kindergartens within 350m buffer (`n_preschool_350m`) | GeoJSON |
| **Kindergartens** | Point locations of registered kindergartens across Singapore, including both public and private institutions. | Count number of kindergartens within 350m buffer (`n_kindergarten_350m`) | GeoJSON |
| **Childcare Centres** | Point locations of licensed childcare centres providing infant and toddler care services. | Count number of childcare centres within 350m buffer (`n_childcare_350m`) to evaluate childcare accessibility, important for working parents. | GeoJSON |
| **Bus Stops** | Point locations of all public bus stops in Singapore's bus network. | Count number of bus stops within 350m buffer (`n_busstop_350m`) as a measure of public transport connectivity and accessibility. | SHP (point geometry) |

### 2.3 Data Import & Initial Manipulation

#### 2.3.1 Singapore Boundary

```{r}
#| label: import-singapore-boundary
#| code-summary: Import Singapore boundary for mapping context

# Import Singapore boundary (adjust path as needed)
singapore_boundary <- st_read(
  dsn = "data/geospatial",
  layer = "MP14_SUBZONE_WEB_PL",  # or your boundary layer name
  quiet = TRUE
) %>%
  st_transform(3414) %>%
  st_union()  # Dissolve to single boundary

cat("✓ Singapore boundary imported\n")
cat("  CRS:", st_crs(singapore_boundary)$input, "\n")
```

```{r}
#| code-summary: Report the extent of mpsz_svy21 layer by its lower and upper limits

st_bbox(singapore_boundary)
```

#### 2.3.2 HDB Resale Flat Prices

```{r}
#| code-summary: Read CSV file
#| output: true
#| warning: false
#| error: false
#| message: false
#| eval: false # saved rds already

HDBresale <- read_csv("data/rawdata/HDBresale.csv") %>% 
  filter (flat_type == "4 ROOM") %>% 
  filter (month >= "2025-01" & month <"2025-10")
```

```{r}
#| code-summary: Define a function to convert the address into latitude and longitude using OneMapSG API
#| output: true
#| warning: false
#| error: false
#| message: false

geocode <- function(block, streetname) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  address <- paste(block, streetname, sep=" ")
  query <- list("searchVal" = address,
                "returnGeom" = "Y",
                "getAddrDetails" = "N",
                "pageNum" = "1")
  
  res <- GET(base_url, query=query)
  restext <- content(res, as="text")
  
  output <- fromJSON(restext) %>%
    as.data.frame %>%
    select(results.LATITUDE, results.LONGITUDE)
  
  return(output)
}
```

```{r}
#| code-summary: Cleaning street names
#| eval: false

HDBresale$street_name <- gsub("ST\\.", "SAINT", HDBresale$street_name)
```

```{r}
#| code-summary: Apply the function
#| output: false
#| warning: false
#| error: false
#| message: false
#| eval: false # saved rds already, no need to call API again to save time

HDBresale$LATITUDE <- 0
HDBresale$LONGITUDE <- 0

for (i in 1:nrow(HDBresale)) {
  temp_output <- geocode(HDBresale[i, 4], HDBresale[i, 5])
  
  HDBresale$LATITUDE[i] <- temp_output$results.LATITUDE
  HDBresale$LONGITUDE[i] <- temp_output$results.LONGITUDE
}
```

```{r}
#| code-summary: Convert HDB resale to sf
#| eval: false # saved rds already

HDBresale_sf <- HDBresale %>%
  st_as_sf(
    coords = c("LONGITUDE", "LATITUDE"),
    crs=4326
  ) %>%
  st_transform(crs=3414)
```

```{r}
#| code-summary: Save rds for HDBresale_sf
#| eval: false
#| echo: false

write_rds(HDBresale_sf, "data/rds/HDBresale_sf.rds")
```

```{r}
#| echo: false
#| output: false
HDBresale_sf <- read_rds("data/rds/HDBresale_sf.rds")
```

#### 2.3.3 Neighborhood and local facilities data

::: panel-tabset

#### CBD Area

```{r}
#| code-summary: Create CBD points dataframe with SVY21 coordinates

cbd_data <- data.frame(
  name = c("Raffles Place", "Tanjong Pagar", "Marina Bay"),
  x = c(30214.5, 29527.8, 31324.7),
  y = c(29414.2, 28645.1, 29877.3)
)
```

#### Hawker Centres

```{r}
#| code-summary: Import Hawker Centres data

hawkerCenters = st_read(
  "data/geospatial/HawkerCentresGEOJSON.geojson") %>%
  st_transform(crs= 3414)
```
#### Parks

```{r}
#| code-summary: Import Parks data

parks = st_read(
  "data/geospatial/Parks.geojson") %>%
  st_transform(crs= 3414)
```

#### Supermarkets

```{r}
#| code-summary: Import Supermarkets data

supermarkets = st_read(
  "data/geospatial/SupermarketsGEOJSON.geojson") %>%
  st_transform(crs= 3414)
```

#### Eldercare Centres

```{r}
#| code-summary: Import Eldercare Centres data

eldercare <- st_read(
  dsn = "data/geospatial", 
  layer = "ELDERCARE"
) %>% 
  st_transform(crs = 3414)
```

#### Preschools

```{r}
#| code-summary: Import Preschool data

preschool = st_read(
  "data/geospatial/PreSchoolsLocation.kml") %>%
  st_transform(crs= 3414)
```

#### Kindergartens

```{r}
#| code-summary: Import Kindergartens data

kindergartens = st_read(
  "data/geospatial/Kindergartens.geojson") %>%
  st_transform(crs= 3414)
```

#### Childcare Centres

```{r}
#| code-summary: Import Childcare Centres data

childCareServices = st_read(
  "data/geospatial/ChildCareServices.geojson") %>%
  st_transform(crs= 3414)
```

#### Bus Stops

```{r}
#| code-summary: Import Bus Stops data

busStops <- st_read(
  dsn = "data/geospatial", 
  layer = "BusStop"
) %>% 
  st_transform(crs = 3414)
```

```{r}
#| code-summary: Convert to sf object with SVY21 CRS (EPSG:3414)

cbd_sf <- st_as_sf(cbd_data, 
                   coords = c("x", "y"), 
                   crs = 3414)

# View the result
print(cbd_sf)
```
:::

### 2.4 Proximity Analysis

```{r}
#| code-summary: Count number of preschools, kindergartens, childcare services, and bus stops

# Preschool
within_350m <- st_is_within_distance(
  HDBresale_sf, preschool, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(N_350M_PRESCHOOL = map_int(within_350m, length))

# Kindergartens
within_350m <- st_is_within_distance(
  HDBresale_sf, kindergartens, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(N_350M_KINDERG = map_int(within_350m, length))

# Child care
within_350m <- st_is_within_distance(
  HDBresale_sf, childCareServices, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(N_350M_CHILDCARE = map_int(within_350m, length))

# Bus Stop
within_350m <- st_is_within_distance(
  HDBresale_sf, busStops, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(N_350M_BUSSTOP = map_int(within_350m, length))

```

```{r}
#| code-summary: Create function for nearest distance calculation

calc_nearest_dist <- function(from_geom, to_sf) {
  as.numeric(
    st_distance(
      from_geom,
      to_sf[st_nearest_feature(from_geom, to_sf), ],
      by_element = TRUE
    )
  )
}
```

```{r}
#| code-summary: Apply function to get the nearest distance to the facilities

hdb_model <- HDBresale_sf %>%
  mutate(
    PROX_CBD           = calc_nearest_dist(geometry, cbd_sf),
    PROX_HAWKER        = calc_nearest_dist(geometry, hawkerCenters),
    PROX_PARK          = calc_nearest_dist(geometry, parks),
    PROX_SUPERMARKET   = calc_nearest_dist(geometry, supermarkets),
    PROX_ELDERCARE     = calc_nearest_dist(geometry, eldercare)
  )

hdb_model
```

```{r}
#| code-summary: Save rds for hdb_model
#| eval: false
#| echo: false

write_rds(hdb_model, "data/rds/hdb_model.rds")
```

```{r}
#| echo: false
#| output: false

hdb_model <- read_rds("data/rds/hdb_model.rds")
```

### 2.5 Aspatial Data Cleaning

::: panel-tabset

#### Step 1: Initial Data Inspection

```{r}
#| code-summary: Check structure

glimpse(hdb_model)
```

```{r}
#| code-summary: Check dimensions

cat("Dataset dimensions:", nrow(hdb_model), "rows ×", ncol(hdb_model), "columns\n")
```

```{r}
#| code-summary: Check for missing values

missing_summary <- hdb_model %>%
  st_drop_geometry() %>%  # Remove geometry for cleaner summary
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") %>%
  filter(missing_count > 0) %>%
  arrange(desc(missing_count))

if(nrow(missing_summary) > 0) {
  cat("\n⚠ Missing values detected:\n")
  print(missing_summary)
} else {
  cat("\n✓ No missing values detected\n")
}
```

#### Step 2: Clean Structural Variables

Purpose: Prepare structural variables for modeling

 - Convert storey_range to numeric midpoint (e.g., "13 TO 15" → 14)
 - Convert remaining_lease to numeric years
 - Extract month as date object

```{r}
#| code-summary: Clean and transform structural variables

hdb_clean <- hdb_model %>%
  mutate(
    # Convert storey_range (e.g., "13 TO 15" → 14)
    storey_midpoint = as.numeric(str_extract(storey_range, "^\\d+")) + 1,
    
    # Convert remaining_lease to numeric years
    # Original format like "56 years 03 months"
    remaining_lease_years = case_when(
      str_detect(remaining_lease, "years") & str_detect(remaining_lease, "months") ~
        as.numeric(str_extract(remaining_lease, "^\\d+")) + 
        as.numeric(str_extract(remaining_lease, "(?<=years )\\d+")) / 12,
      str_detect(remaining_lease, "years") ~
        as.numeric(str_extract(remaining_lease, "^\\d+")),
      TRUE ~ NA_real_
    ),
    
    # Convert month to date
    month_date = as.Date(paste0(month, "-01")),
    
    # Extract year and month for temporal analysis
    year = year(month_date),
    month_num = month(month_date)
  )

# Check transformations
cat("✓ Storey midpoint range:", 
    min(hdb_clean$storey_midpoint, na.rm = TRUE), "to", 
    max(hdb_clean$storey_midpoint, na.rm = TRUE), "\n")

cat("✓ Remaining lease range:", 
    round(min(hdb_clean$remaining_lease_years, na.rm = TRUE), 1), "to", 
    round(max(hdb_clean$remaining_lease_years, na.rm = TRUE), 1), "years\n")
```
#### Step 3: Check for Outliers

```{r}
#| code-summary: Identify outliers in key variables of `resale_price` and `resale_price`

# Function to detect outliers using IQR method
detect_outliers <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  return(x < lower_bound | x > upper_bound)
}

# Check outliers in resale_price
outlier_price <- detect_outliers(hdb_clean$resale_price)
cat("Outliers in resale_price:", sum(outlier_price), 
    "(", round(sum(outlier_price)/nrow(hdb_clean)*100, 2), "%)\n")

# Check outliers in resale_price
outlier_area <- detect_outliers(hdb_clean$floor_area_sqm)
cat("Outliers in floor_area_sqm:", sum(outlier_area), 
    "(", round(sum(outlier_area)/nrow(hdb_clean)*100, 2), "%)\n")

# Visual inspection with boxplots
par(mfrow = c(1, 2))
boxplot(hdb_clean$resale_price, main = "Resale Price", ylab = "Price (SGD)")
boxplot(hdb_clean$floor_area_sqm, main = "Floor Area", ylab = "Square Meters")
par(mfrow = c(1, 1))
```
For now, we don't remove outliers yet. We just flag them using 2 additional columns of `outlier_price` and `outlier_area`, and keep in the dataset. Those detected will be report "TRUE" in `outlier_price` or `outlier_area` columns.

```{r}
#| code-summary: Flag outliers

hdb_clean <- hdb_clean %>%
  mutate(
    outlier_price = detect_outliers(resale_price),
    outlier_area = detect_outliers(floor_area_sqm)
  )

hdb_clean
```

:::

```{r}
#| code-summary: Save rds for hdb_clean
#| eval: false
#| echo: false

write_rds(hdb_clean, "data/rds/hdb_clean.rds")
```

```{r}
#| echo: false
#| output: false

hdb_clean <- read_rds("data/rds/hdb_clean.rds")
```

## 3. Exploratory Spatial Data Analysis

### 3.1 Descriptive Statistics

#### 3.1.1 Summary in tables

::: panel-tabset

#### Structural variables

```{r}
#| code-summary: Summary statistics for structural variables

structural_vars <- c("resale_price", "floor_area_sqm", "storey_midpoint", 
                     "remaining_lease_years")

structural_summary <- hdb_clean %>%
  st_drop_geometry() %>%
  select(all_of(structural_vars)) %>%
  summarise(across(everything(), 
                   list(mean = ~mean(., na.rm = TRUE),
                        median = ~median(., na.rm = TRUE),
                        sd = ~sd(., na.rm = TRUE),
                        min = ~min(., na.rm = TRUE),
                        max = ~max(., na.rm = TRUE)),
                   .names = "{.col}_{.fn}")) %>%
  pivot_longer(everything(), names_to = "stat", values_to = "value") %>%
  separate(stat, into = c("variable", "statistic"), sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = statistic, values_from = value)

knitr::kable(structural_summary, 
             digits = 2,
             caption = "Descriptive Statistics: Structural Variables",
             col.names = c("Variable", "Mean", "Median", "SD", "Min", "Max"))
```

#### Proximity variables

```{r}
#| code-summary: Summary statistics for Proximity variables summary (in meters)

prox_vars <- hdb_clean %>%
  st_drop_geometry() %>%
  select(starts_with("PROX_")) %>%
  names()

location_summary <- hdb_clean %>%
  st_drop_geometry() %>%
  select(all_of(prox_vars)) %>%
  summarise(across(everything(), 
                   list(mean = ~mean(., na.rm = TRUE),
                        median = ~median(., na.rm = TRUE),
                        sd = ~sd(., na.rm = TRUE),
                        min = ~min(., na.rm = TRUE),
                        max = ~max(., na.rm = TRUE)),
                   .names = "{.col}_{.fn}")) %>%
  pivot_longer(everything(), names_to = "stat", values_to = "value") %>%
  separate(stat, into = c("variable", "statistic"), sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = statistic, values_from = value)

knitr::kable(location_summary, 
             digits = 2,
             caption = "Descriptive Statistics: Location Variables (Proximity in meters)",
             col.names = c("Variable", "Mean", "Median", "SD", "Min", "Max"))
```

#### COUNT variables

```{r}
#| code-summary: Descriptive statistics for count variables

# --- Location Variables: Counts (within buffer) ---
count_vars <- names(hdb_clean)[str_detect(names(hdb_clean), "^N_")]

count_summary <- hdb_clean %>%
  st_drop_geometry() %>%
  select(all_of(count_vars)) %>%
  summarise(across(everything(), 
                   list(mean = ~mean(., na.rm = TRUE),
                        median = ~median(., na.rm = TRUE),
                        sd = ~sd(., na.rm = TRUE),
                        min = ~min(., na.rm = TRUE),
                        max = ~max(., na.rm = TRUE)),
                   .names = "{.col}_{.fn}")) %>%
  pivot_longer(everything(), names_to = "stat", values_to = "value") %>%
  separate(stat, into = c("variable", "statistic"), sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = statistic, values_from = value)

knitr::kable(count_summary, 
             digits = 1,
             caption = "Table 3: Descriptive Statistics - Count Variables (facilities within buffer)",
             col.names = c("Variable", "Mean", "Median", "SD", "Min", "Max"))
```

:::

#### 3.1.2 Summary with Visualization 

::: panel-tabset

#### Distribution Analysis

```{r}
#| code-summary: Visualize distributions of key variables
#| fig-width: 12
#| fig-height: 8

# Create histogram panels for key variables
library(patchwork)

# Resale price distribution
p1 <- ggplot(hdb_clean, aes(x = resale_price)) +
  geom_histogram(bins = 50, fill = "#440154", alpha = 0.7) +
  geom_vline(aes(xintercept = median(resale_price)), 
             color = "red", linetype = "dashed", linewidth = 1) +
  labs(title = "Distribution of Resale Prices",
       x = "Resale Price (SGD)", y = "Frequency") +
  scale_x_continuous(labels = scales::comma) +
  theme_minimal()

# Floor area distribution
p2 <- ggplot(hdb_clean, aes(x = floor_area_sqm)) +
  geom_histogram(bins = 30, fill = "#31688e", alpha = 0.7) +
  geom_vline(aes(xintercept = median(floor_area_sqm)), 
             color = "red", linetype = "dashed", linewidth = 1) +
  labs(title = "Distribution of Floor Area",
       x = "Floor Area (sqm)", y = "Frequency") +
  theme_minimal()

# Remaining lease distribution
p3 <- ggplot(hdb_clean, aes(x = remaining_lease_years)) +
  geom_histogram(bins = 40, fill = "#35b779", alpha = 0.7) +
  geom_vline(aes(xintercept = median(remaining_lease_years)), 
             color = "red", linetype = "dashed", linewidth = 1) +
  labs(title = "Distribution of Remaining Lease",
       x = "Remaining Lease (years)", y = "Frequency") +
  theme_minimal()

# Storey distribution
p4 <- ggplot(hdb_clean, aes(x = storey_midpoint)) +
  geom_histogram(bins = 20, fill = "#fde724", alpha = 0.7) +
  geom_vline(aes(xintercept = median(storey_midpoint)), 
             color = "red", linetype = "dashed", linewidth = 1) +
  labs(title = "Distribution of Storey Level",
       x = "Storey Midpoint", y = "Frequency") +
  theme_minimal()

# Combine plots
(p1 + p2) / (p3 + p4) +
  plot_annotation(
    title = "Distribution of Structural Variables",
    subtitle = "Red dashed line indicates median value"
  )
```

#### Distribution of Count variables

```{r}
#| code-summary: Distribution of count variables  
#| fig-width: 12
#| fig-height: 10

# Count variables distributions
c1 <- ggplot(hdb_clean, aes(x = N_350M_PRESCHOOL)) +
  geom_bar(fill = "#440154", alpha = 0.7) +
  labs(title = "Preschools within 350m", x = "Count", y = "Frequency") +
  theme_minimal()

c2 <- ggplot(hdb_clean, aes(x = N_350M_KINDERG)) +
  geom_bar(fill = "#31688e", alpha = 0.7) +
  labs(title = "Kindergartens within 350m", x = "Count", y = "Frequency") +
  theme_minimal()

c3 <- ggplot(hdb_clean, aes(x = N_350M_CHILDCARE)) +
  geom_bar(fill = "#35b779", alpha = 0.7) +
  labs(title = "Childcare Centres within 350m", x = "Count", y = "Frequency") +
  theme_minimal()

c4 <- ggplot(hdb_clean, aes(x = N_350M_BUSSTOP)) +
  geom_bar(fill = "#fde724", alpha = 0.7) +
  labs(title = "Bus Stops within 350m", x = "Count", y = "Frequency") +
  theme_minimal()

(c1 + c2) / (c3 + c4) +
  plot_annotation(
    title = "Figure 2: Distribution of Count Variables",
    subtitle = "Number of facilities within buffer zones"
  )
```

#### Temporal Distribution

```{r}
#| code-summary: Transaction volume and prices over time
#| fig-width: 12
#| fig-height: 6

# Transaction volume by month
t1 <- hdb_clean %>%
  st_drop_geometry() %>%
  count(month_date) %>%
  ggplot(aes(x = month_date, y = n)) +
  geom_line(color = "#440154", linewidth = 1) +
  geom_point(color = "#440154", size = 2) +
  labs(title = "Transaction Volume Over Time",
       x = "Month", y = "Number of Transactions") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Median price by month
t2 <- hdb_clean %>%
  st_drop_geometry() %>%
  group_by(month_date) %>%
  summarise(
    median_price = median(resale_price),
    mean_price = mean(resale_price),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = month_date)) +
  geom_line(aes(y = median_price, color = "Median"), linewidth = 1) +
  geom_line(aes(y = mean_price, color = "Mean"), linewidth = 1) +
  geom_point(aes(y = median_price, color = "Median"), size = 2) +
  geom_point(aes(y = mean_price, color = "Mean"), size = 2) +
  labs(title = "Resale Price Trends Over Time",
       x = "Month", y = "Resale Price (SGD)", color = "Statistic") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +
  scale_color_manual(values = c("Median" = "#440154", "Mean" = "#fde724")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

t1 / t2
```

#### Summary by Town

```{r}
#| code-summary: Resale prices by town

# Calculate summary statistics by town
town_summary <- hdb_clean %>%
  st_drop_geometry() %>%
  group_by(town) %>%
  summarise(
    n_transactions = n(),
    median_price = median(resale_price),
    mean_price = mean(resale_price),
    min_price = min(resale_price),
    max_price = max(resale_price),
    .groups = "drop"
  ) %>%
  arrange(desc(median_price))

# Display table
knitr::kable(town_summary, 
             digits = 0,
             caption = "Resale Price Summary by Town (4-Room Flats)",
             col.names = c("Town", "Transactions", "Median Price", 
                          "Mean Price", "Min Price", "Max Price"),
             format.args = list(big.mark = ","))

# Top 10 most expensive towns
top10_towns <- town_summary %>%
  slice_max(median_price, n = 10) %>%
  ggplot(aes(x = reorder(town, median_price), y = median_price)) +
  geom_col(fill = "#440154", alpha = 0.8) +
  geom_text(aes(label = scales::comma(median_price)), 
            hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "Top 10 Towns by Median Resale Price",
       subtitle = "4-Room Flats (Jan-Sep 2025)",
       x = "Town", y = "Median Resale Price (SGD)") +
  scale_y_continuous(labels = scales::comma, 
                     expand = expansion(mult = c(0, 0.1))) +
  theme_minimal()

print(top10_towns)
```

#### Summary by Flat Model

```{r}
#| code-summary: Resale prices by flat model

# Summary by flat_model
model_summary <- hdb_clean %>%
  st_drop_geometry() %>%
  group_by(flat_model) %>%
  summarise(
    n_transactions = n(),
    median_price = median(resale_price),
    pct_of_total = n() / nrow(hdb_clean) * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(n_transactions))

knitr::kable(model_summary, 
             digits = 1,
             caption = "Resale Price Summary by Flat Model",
             col.names = c("Flat Model", "Transactions", 
                          "Median Price", "% of Total"),
             format.args = list(big.mark = ","))
```

:::

#### 3.1.3 Bivariate Relationships

```{r}
#| code-summary: Key relationships between price and predictors
#| fig-width: 12
#| fig-height: 10

# Price vs Floor Area
b1 <- ggplot(hdb_clean, aes(x = floor_area_sqm, y = resale_price)) +
  geom_point(alpha = 0.3, color = "#440154") +
  geom_smooth(method = "lm", color = "#fde724", linewidth = 1.5) +
  labs(title = "Price vs Floor Area",
       x = "Floor Area (sqm)", y = "Resale Price (SGD)") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()

# Price vs Remaining Lease
b2 <- ggplot(hdb_clean, aes(x = remaining_lease_years, y = resale_price)) +
  geom_point(alpha = 0.3, color = "#31688e") +
  geom_smooth(method = "lm", color = "#fde724", linewidth = 1.5) +
  labs(title = "Price vs Remaining Lease",
       x = "Remaining Lease (years)", y = "Resale Price (SGD)") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()

# Price vs Storey
b3 <- ggplot(hdb_clean, aes(x = storey_midpoint, y = resale_price)) +
  geom_point(alpha = 0.3, color = "#35b779") +
  geom_smooth(method = "lm", color = "#fde724", linewidth = 1.5) +
  labs(title = "Price vs Storey Level",
       x = "Storey Midpoint", y = "Resale Price (SGD)") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()

# Price vs Distance to CBD
b4 <- ggplot(hdb_clean, aes(x = PROX_CBD, y = resale_price)) +
  geom_point(alpha = 0.3, color = "#fde724") +
  geom_smooth(method = "lm", color = "#440154", linewidth = 1.5) +
  labs(title = "Price vs Distance to CBD",
       x = "Distance to CBD (meters)", y = "Resale Price (SGD)") +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()

(b1 + b2) / (b3 + b4) +
  plot_annotation(
    title = "Figure 6: Bivariate Relationships with Resale Price",
    subtitle = "Linear trend lines shown in yellow/purple"
  )
```

### 3.2 Spatial Distribution of Resale Prices

::: panel-tabset

#### Basic Point Map of Transactions

```{r}
#| label: map-transaction-points
#| code-summary: Point map showing all HDB resale transactions
#| fig-width: 10
#| fig-height: 12

tmap_mode("plot")  # Static map mode

# Create basic point map
tm_shape(singapore_boundary) +
  tm_borders(col = "grey50", lwd = 1) +
  tm_fill(col = "grey95") +
tm_shape(hdb_clean) +
  tm_dots(col = "resale_price",
          palette = "YlOrRd",
          size = 0.05,
          alpha = 0.6,
          title = "Resale Price (SGD)",
          breaks = c(0, 500000, 600000, 700000, 800000, 900000, 1500000),
          labels = c("< $500k", "$500k-$600k", "$600k-$700k", 
                     "$700k-$800k", "$800k-$900k", "> $900k")) +
  tm_layout(
    main.title = "Map 1: Spatial Distribution of HDB 4-Room Resale Prices",
    main.title.size = 1.2,
    main.title.position = "center",
    legend.outside = TRUE,
    legend.outside.position = "right",
    legend.title.size = 1,
    legend.text.size = 0.7,
    frame = FALSE
  ) +
  tm_compass(position = c("left", "top"), size = 2)
  # Scale bar removed due to tmap bug
```

### 3.3 Correlation Structure

### 3.4 Spatial Autocorrelation

## 4. Model Calibration

### 4.1 OLS Baseline Model

#### 4.1.1 Model Specification

#### 4.1.2 Results & Diagnostics

#### 4.1.3 Spatial Autocorrelation Test

### 4.2 Geographically Weighted Regression

#### 4.2.1 Bandwidth Selection

\[Explain CV vs AIC choice\] \

#### 4.2.2 Model Results 

#### 4.2.3 Model Comparison

## 5. Spatial Patterns & Interpretation

### 5.1 Overview of Local Variations

### 5.2 Key Findings by Variable

\[Subsection for each important coefficient\] \

### 5.3 Policy & Market Implications

## 6. Limitations & Conclusions

### 6.1 Study Limitations

\[Data period, 4-room only, modifiable areal unit problem, etc.\] \

### 6.2 Conclusions \

### 6.3 Future Research Directions

## Appendix

### References
