---
title: "Hands-on Exercise 9 | Modelling Geographical Accessibility"
subtitle: "In this exercise, we learn about Accessibility Measurement, with the aim of quantifying accessibility using established spatial models."
author: Linh Nguyen
execute:
  echo: true
  eval: true
  freeze: auto
  message: false
  warning: false
format:
  html:
    code-link: true
    toc: true
number-sections: false
number-offset: 1
editor: visual
---

## 1 Introduction

**Geospatial accessibility** is a core concept in spatial analysis that measures the ease with which people can reach desired opportunities or services (like jobs, healthcare, schools, or grocery stores) from a given location, considering the friction of distance (usually measured in travel time or distance).

It goes beyond simply counting the number of services available by incorporating the spatial reality of **travel impedance**. For instance, an area might have many hospitals, but if they are all a two-hour drive away, accessibility is low. It's fundamentally about opportunity and equity—ensuring that everyone, regardless of where they live, has reasonable access to necessary resources.

**Key Skills:**

**Hansen's Potential Model:** Measures accessibility as a weighted sum of opportunities surrounding a location, where the opportunities are dampened (decayed) by travel time. It gives a single, relative "potential" score for each location.

**Spatial Accessibility Measure (SAM):** Refers to a class of models, often focusing on supply-demand balance, such as the Two-Step Floating Catchment Area (2SFCA) method. These measures calculate accessibility while considering both the supply capacity and the demand from surrounding areas, aiming to provide a more realistic access ratio.

## 2 The dataset

Four data sets will be used in this hands-on exercise, they are:

`MP14_SUBZONE_NO_SEA_PL:` URA Master Plan 2014 subzone boundary GIS data. This data set is downloaded from data.gov.sg.

`hexagons:` A 250m radius hexagons GIS data. This data set was created by using st_make_grid() of sf package. It is in ESRI shapefile format.

`ELDERCARE:` GIS data showing location of eldercare service. This data is downloaded from data.gov.sg. There are two versions. One in ESRI shapefile format. The other one in Google kml file format. For the purpose of this hands-on exercise, ESRI shapefile format is provided.

`OD_Matrix:` a distance matrix in csv format. There are six fields in the data file. They are:

-   origin_id: the unique id values of the origin (i.e. fid of hexagon data set.),
-   destination_id: the unique id values of the destination (i.e. fid of ELDERCARE data set.),
-   entry_cost: the perpendicular distance between the origins and the nearest road),
-   network_cost: the actual network distance from the origin and destination,
-   exit_cost: the perpendicular distance between the destination and the nearest road), and
-   total_cost: the summation of entry_cost, network_cost and exit_cost.

All the values of the cost related fields are in **metres**.

## 3 Installing R packages

Packages used: - `spatialAcc`: This is used to model geographical accessibility

```{r}
pacman::p_load(tmap, SpatialAcc, sf, 
               ggstatsplot, reshape2,
               tidyverse)
```

## 4 Geospatial Data Wranglign

### 4.1 Importing geospatial data

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_NO_SEA_PL")
```

```{r}
hexagons <- st_read(dsn = "data/geospatial", layer = "hexagons") 
```

```{r}
eldercare <- st_read(dsn = "data/geospatial", layer = "ELDERCARE") 
```

The report above shows that the R object used to contain the imported *MP14_SUBZONE_WEB_PL* shapefile is called **mpsz** and it is a simple feature object. The geometry type is multipolygon. it is also important to note that mpsz simple feature object does not have EPSG information.

### 4.2 Updating CRS information

We update the mpsz data with correct ESPG code (i.e. 3414)

```{r}
mpsz <- st_transform(mpsz, 3414)
eldercare <- st_transform(eldercare, 3414)
hexagons <- st_transform(hexagons, 3414)
```

After the transformation, we will verify the projection of the newly transformed *mpsz_svy21* by using `st_crs()` of **sf** package.

```{r}
st_crs(mpsz)
```

EPSG has become 3414 now.

### 4.3 Cleaning and updating attribute fields of the geospatial data

There are many redundant fields in the data tables of both *eldercare* and *hexagons*. The code chunks below will be used to exclude those redundant fields. At the same time, a new field called *demand* and a new field called *capacity* will be added into the data table of hexagons and eldercare sf data frame respectively. Both fields are derive using mutate() of dplyr package.

```{r}
eldercare <- eldercare %>%
  select(fid, ADDRESSPOS) %>%
  mutate(capacity = 100)
```

```{r}
hexagons <- hexagons %>%
  select(fid) %>%
  mutate(demand = 100)
```

For this exercise, we use a constant value of 100, but actual demand of the hexagon and capacity of the eldercare centre should be used in practice.

## 5 Aspatial Data Handling and Wrangling

### 5.1 Importing Distance Matrix

The code chunk below uses `read_cvs()` of **readr** package to import OD_Matrix.csv into RStudio. The imported object is a tibble data.frame called ODMatrix.

```{r}
ODMatrix <- read_csv("data/aspatial/OD_Matrix.csv", skip = 0)
```

### 5.2 Tidying distance matrix

We use `spread()` of **tidyr** package to transform the O-D matrix from a thin format into a fat format.

```{r}
distmat <- ODMatrix %>%
  select(origin_id, destination_id, total_cost) %>%
  spread(destination_id, total_cost)%>%
  select(c(-c('origin_id')))
```

Currently, the distance is measured in metres because SVY21 projected coordinate system is used. We will then convert the unit f measurement from metre to kilometre.

```{r}
distmat_km <- as.matrix(distmat/1000)
```

## 6 Modelling and Visualising Accessibility using Hansen Method

### 6.1 Computing Hansen's accessibility

We will compute Hansen's accessibility using `ac()` of **SpatialAcc** package.

```{r}
acc_Hansen <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 2, 
                            family = "Hansen"))
```

Changing the column name to "accHansen" for better readability.

```{r}
colnames(acc_Hansen) <- "accHansen"
```

Next, we will convert the data table into tibble format.

```{r}
acc_Hansen <- as_tibble(acc_Hansen)
```

Lastly, `bind_cols()` of **dplyr** will be used to join the *acc_Hansen* tibble data frame with the hexagons simple feature data frame. The output is called *hexagon_Hansen*.

```{r}
hexagon_Hansen <- bind_cols(hexagons, acc_Hansen)
```

```{r}
acc_Hansen <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 0.5, 
                            family = "Hansen"))

colnames(acc_Hansen) <- "accHansen"
acc_Hansen <- as_tibble(acc_Hansen)
hexagon_Hansen <- bind_cols(hexagons, acc_Hansen)
```

### 6.2 Visualising Hansen’s accessibility

#### 6.2.1 Extracting map extend

Firstly, we will extract the extend of hexagons simple feature data frame by by using `st_bbox()` of **sf** package.

```{r}
mapex <- st_bbox(hexagons)
```

We use a collection of mapping fucntions of tmap package to create a high cartographic quality accessibility to eldercare centres in Singapore.

```{r}
tm_shape(hexagon_Hansen,
         bbox = mapex) + 
  tm_polygons(fill = "accHansen", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Accessibility to eldercare: Hansen method") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)
```

### 6.3 Statistical graphic visualisation

We are going to compare the distribution of Hansen’s accessibility values by URA Planning Region.

Firstly, we need to add the planning region field into the *hexagon_Hansen* simple feature data frame.

```{r}
hexagon_Hansen <- st_join(hexagon_Hansen, mpsz, 
                          join = st_intersects)
```

Next, `ggplot()` will be used to plot the distribution by using boxplot graphical method.

```{r}
ggplot(data=hexagon_Hansen, 
       aes(y = log(accHansen), 
           x= REGION_N)) +
  geom_boxplot() +
  geom_point(stat="summary", 
             fun.y="mean", 
             colour ="red", 
             size=2)
```

## 7 Modelling and Visualising Accessibility using KD2SFCA Method

### 7.1 Computing KD2SFCA’s accessibility

We will calculate Hansen’s accessibility using `ac()` of **SpatialAcc** and `data.frame()` is used to save the output in a data frame called *acc_KD2SFCA*. Notice that KD2SFCA is used for family argument.

```{r}
acc_KD2SFCA <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            d0 = 5,
                            power = 2, 
                            family = "KD2SFCA"))

colnames(acc_KD2SFCA) <- "accKD2SFCA"
acc_KD2SFCA <- as_tibble(acc_KD2SFCA)
hexagon_KD2SFCA <- bind_cols(hexagons, acc_KD2SFCA)
```

### 7.2 Visualising KD2SFCA’s accessibility

We use a collection of mapping functions of **tmap** package to create a high cartographic quality accessibility to eldercare centres in Singapore. Notice that mapex is reused for bbox argument.

```{r}
tm_shape(hexagon_KD2SFCA,
         bbox = mapex) + 
  tm_polygons(fill = "accKD2SFCA", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Accessibility to eldercare: KD2SFCA method") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)
```

### 7.3 Statistical graphic visualisation

Now, we are going to compare the distribution of KD2CFA accessibility values by URA Planning Region.

Firstly, we need to add the planning region field into *hexagon_KD2SFCA* simple feature data frame.

```{r}
hexagon_KD2SFCA <- st_join(hexagon_KD2SFCA, mpsz, 
                          join = st_intersects)
```

Next, `ggplot()` will be used to plot the distribution by using boxplot graphical method.

```{r}
ggplot(data=hexagon_KD2SFCA, 
       aes(y = accKD2SFCA, 
           x= REGION_N)) +
  geom_boxplot() +
  geom_point(stat="summary", 
             fun.y="mean", 
             colour ="red", 
             size=2)
```

## 8 Modelling and Visualising Accessibility using Spatial Accessibility Measure (SAM) Method

### 8.1 Computing SAM accessibility

We will calculate Hansen’s accessibility using `ac()` of **SpatialAcc** and `data.frame()` is used to save the output in a data frame called *acc_SAM*. Notice that SAM is used for family argument.

```{r}
acc_SAM <- data.frame(ac(hexagons$demand,
                         eldercare$capacity,
                         distmat_km, 
                         d0 = 5,
                         power = 2, 
                         family = "SAM"))

colnames(acc_SAM) <- "accSAM"
acc_SAM <- as_tibble(acc_SAM)
hexagon_SAM <- bind_cols(hexagons, acc_SAM)
```

### 8.2 Visualising SAM’s accessibility

We use a collection of mapping fucntions of **tmap** package to create a high cartographic quality accessibility to eldercare centres in Singapore. Notice that mapex is reused for bbox argument.

```{r}
tm_shape(hexagon_SAM,
         bbox = mapex) + 
  tm_polygons(fill = "accSAM", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Accessibility to eldercare: SAM method") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)
```

### 8.3 Statistical graphic visualisation

Now, we are going to compare the distribution of SAM accessibility values by URA Planning Region.

Firstly, we need to add the planning region field into *hexagon_SAM* simple feature data frame by using the code chunk below.

```{r}
hexagon_SAM <- st_join(hexagon_SAM, mpsz, 
                       join = st_intersects)
```

Next, `ggplot()` will be used to plot the distribution by using boxplot graphical method.

```{r}
ggplot(data=hexagon_SAM, 
       aes(y = accSAM, 
           x= REGION_N)) +
  geom_boxplot() +
  geom_point(stat="summary", 
             fun.y="mean", 
             colour ="red", 
             size=2)
```
