---
title: "In-class Exercise 5 | Global & Local Measures of Spatial Autocorrelation"
subtitle: "In this in-class exercise, you will learn how to compute Global & Local Measures of Spatial Autocorrelation (GMSA) by using sfdep package"
date: "Sep 27, 2025"
date-modified: "last-modified"
author: Linh Nguyen
execute:
  echo: true
  eval: true
  freeze: auto
  message: false
  warning: false
format:
  html:
    code-link: true
    toc: true
number-sections: true
number-offset: 1
editor: visual
---

## Preparation

```{r}
#| code-summary: Install and Load R packages
#| output: false

#install.packages("sfdep")

pacman::p_load(sf, tidyverse, sfdep, tmap)
```

```{r}
#| code-summary: Import data
#| output: false

hunan <- st_read(dsn = "data/geospatial", layer = "Hunan")
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

```{r}
#| code-summary: Join data

hunan_GDPPC <- left_join(hunan, hunan2012) %>% 
  select(1:4, 7, 15)
```

## Chloropleth map to show the distribution of GDPPC of Hunan Province

```{r}
tmap_mode("plot")
tm_shape(hunan_GDPPC) +
  tm_fill("GDPPC", 
          style = "quantile", 
          palette = "Blues",
          title = "GDPPC") +
  tm_layout(main.title = "Distribution of GDP per capita by county, Hunan Province",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha =0.2)
```

## Global Measures of Spatial Association

### Deriving Queen's contiguity weights: sfdep methods

```{r}
wm_q <- hunan_GDPPC %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(nb,
                         style = "W"),
         .before = 1) 

wm_q
```

### Computing Global Moran' I test

```{r}
moranI <- global_moran(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt)
glimpse(moranI)
```

### Performing Global Moran'I permutation test

```{r}
#| code-summary: Set seed for the simulation

set.seed(1234)
```

```{r}
global_moran_perm(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt,
                  nsim = 99)
```

## Hot Spot and Cold Spot Area Analysis (HCSA)

Deriving special weight matrix based on distances.

```{r}
wm_idw <- hunan_GDPPC %>%
  mutate(nb = include_self(
    st_contiguity(geometry)),
    wts = st_inverse_distance(nb, 
                              geometry, 
                              scale = 1,
                              alpha = 1),
         .before = 1)
```

Compute local Gi\* statistics.

```{r}
HCSA <- wm_idw %>% 
  mutate(local_Gi = local_gstar_perm(
    GDPPC, nb, wts, nsim = 99),
         .before = 1) %>%
  unnest(local_Gi)
HCSA
```

### Visualising Gi\*

```{r}
tmap_mode("plot")
tm_shape(HCSA) +
  tm_fill("gi_star") + 
  tm_borders(alpha = 0.5) +
  tm_view(set.zoom.limits = c(6,8))
```

### Visualising p-value of HCSA

```{r}
tmap_mode("plot")
tm_shape(HCSA) +
  tm_fill("p_sim") + 
  tm_borders(alpha = 0.5)
```

### Visualising local HCSA

```{r}
tmap_mode("plot")
map1 <- tm_shape(HCSA) +
  tm_fill("gi_star") + 
  tm_borders(alpha = 0.5) +
  tm_view(set.zoom.limits = c(6,8)) +
  tm_layout(main.title = "Gi* of GDPPC",
            main.title.size = 0.8)

map2 <- tm_shape(HCSA) +
  tm_fill("p_value",
          breaks = c(0, 0.001, 0.01, 0.05, 1),
              labels = c("0.001", "0.01", "0.05", "Not sig")) + 
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "p-value of Gi*",
            main.title.size = 0.8)

tmap_arrange(map1, map2, ncol = 2)
```

### Visualising hot spot and cold spot areas

```{r}
HCSA_sig <- HCSA  %>%
  filter(p_sim < 0.05)
tmap_mode("plot")
tm_shape(HCSA) +
  tm_polygons() +
  tm_borders(alpha = 0.5) +
tm_shape(HCSA_sig) +
  tm_fill("cluster") + 
  tm_borders(alpha = 0.4)
```

Figure above reveals that there is one hot spot area and two cold spot areas. Interestingly, the hot spot areas coincide with the High-high cluster identifies by using local Moran's I method in the earlier sub-section.
