---
title: "In-class Exercise 9: Geography of Accessibility"
date: "Nov 09, 2025"
date-modified: "last-modified"
author: Linh Nguyen
execute:
  echo: true
  eval: true
  freeze: auto
  message: false
  warning: false
format:
  html:
    code-link: true
    toc: true
number-sections: true
number-offset: 1
editor: visual
---

## Getting Started

```{r}
pacman::p_load(SpatialAcc, sf, tidyverse, 
               tmap, ggstatsplot)
```

## Count Number of Points within a Distance

The task is to count number of point features (i.e. CHAS clinics) within 1km of another point features (i.e. eldercare centre).

::::::::::::: panel-tabset

### Read data

```{r}
#| output: false
eldercare <- st_read(dsn = "data/rawdata/ELDERCARE",
                     layer = "ELDERCARE") %>%
  st_transform(crs = 3414)
```

```{r}
#| output: false
CHAS <- st_read("data/rawdata/CHASClinics.kml") %>%
  st_transform(crs = 3414)
```
### Buffering

Next, `st_buffer()` of sf package is used to create a buffer of 1km around each eldercare features

::: {style="font-size: 1.50em"}
```{r}
buffer_1km <- st_buffer(eldercare, 
                        dist = 1000)
```
:::

### Visualising

::::::: columns
:::: {.column width="40%"}
The code chunk below is used to plot the newly created buffers and the CHAS clinics.

```{r}
#| eval: false
tmap_mode("view")
tm_shape(buffer_1km) +
  tm_polygons() +
tm_shape(CHAS) +
  tm_dots()
```

::::

:::: {.column width="60%"}

```{r}
#| echo: false
tmap_mode("view")
tm_shape(buffer_1km) +
  tm_polygons() +
tm_shape(CHAS) +
  tm_dots()
```
::::
:::::::

### Counting points

Lastly, the code chunk below is used to count the number of CHAS clinics with 1km of each eldercare centre.


```{r}
buffer_1km$pts_count <- lengths(
  st_intersects(buffer_1km, CHAS))
```
:::::::::::::

## Importing Data

::::: panel-tabset
### Geospatail data


```{r}
#| output: false
mpsz <- st_read(dsn = "data/geospatial",
                layer = "MP14_SUBZONE_NO_SEA_PL") %>%
  st_transform(crs = 3414)

hexagons <- st_read(dsn = "data/geospatial",
                   layer = "hexagons") %>%
  st_transform(crs = 3414)

eldercare <- st_read(dsn = "data/geospatial",
                     layer = "ELDERCARE") %>%
  st_transform(csr = 3414)
```

### OD Matrix

```{r}
ODMatrix <- read_csv("data/aspatial/OD_Matrix.csv", 
                     skip = 0)
```
:::::

## Data cleaning and Updating Attributes

:::::: panel-tabset
### Supply

```{r}
eldercare <- eldercare %>%
  select(fid, ADDRESSPOS) %>%
  mutate(capacity = 100)
```


### Demand

```{r}
hexagons <- hexagons %>%
  select(fid) %>%
  mutate(demand = 100)
```


### OD Matrix

```{r}
distmat <- ODMatrix %>%
  select(origin_id, destination_id, total_cost) %>%
  spread(destination_id, total_cost)%>%
  select(c(-c('origin_id')))
```

```{r}
distmat_km <- as.matrix(distmat/1000)
```

::::::

## Computing Handsen's Accessibility

:::::: panel-tabset

### power=0.5

```{r}
acc_Hansen_05 <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 0.5, 
                            family = "Hansen"))

colnames(acc_Hansen_05) <- "accHansen_05"
acc_Hansen_05 <- as_tibble(acc_Hansen_05)
hexagon_Hansen_05 <- bind_cols(hexagons, acc_Hansen_05)
```

### power=1

```{r}
acc_Hansen_10 <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 1, 
                            family = "Hansen"))

colnames(acc_Hansen_10) <- "accHansen_10"
acc_Hansen_10 <- as_tibble(acc_Hansen_10)
hexagon_Hansen_10 <- bind_cols(hexagons, acc_Hansen_10)
```

### power=1.5

```{r}
acc_Hansen_15 <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 0.5, 
                            family = "Hansen"))

colnames(acc_Hansen_15) <- "accHansen_15"
acc_Hansen_15 <- as_tibble(acc_Hansen_15)
hexagon_Hansen_15 <- bind_cols(hexagons, acc_Hansen_15)
```

### power=2

```{r}
acc_Hansen_20 <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 2, 
                            family = "Hansen"))
#  Tidy the output
colnames(acc_Hansen_20) <- "accHansen_20"
acc_Hansen_20 <- as_tibble(acc_Hansen_20)
hexagon_Hansen_20 <- bind_cols(hexagons, acc_Hansen_20)
```

### power=2.5


```{r}
acc_Hansen_25 <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 2.5, 
                            family = "Hansen"))
#  Tidy the output
colnames(acc_Hansen_25) <- "accHansen_25"
acc_Hansen_25 <- as_tibble(acc_Hansen_25)
hexagon_Hansen_25 <- bind_cols(hexagons, acc_Hansen_25)
```

::::::

------------------------------------------------------------------------

## Visualising Accessibility

:::: panel-tabset
### power=0.5

```{r}
#| fig-width: 12
#| fig-height: 7
#| echo: false
mapex <- st_bbox(hexagons)

tmap_mode("plot")
tm_shape(hexagon_Hansen_05,
         bbox = mapex) + 
  tm_fill(col = "accHansen_05",
          n = 10,
          style = "quantile",
          border.col = "black",
          border.lwd = 1) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to eldercare: Hansen method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 6),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(lwd = 0.1, alpha = 0.5)
```


### power=1

```{r}
#| fig-width: 12
#| fig-height: 7
#| echo: false
mapex <- st_bbox(hexagons)

tmap_mode("plot")
tm_shape(hexagon_Hansen_10,
         bbox = mapex) + 
  tm_fill(col = "accHansen_10",
          n = 10,
          style = "quantile",
          border.col = "black",
          border.lwd = 1) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to eldercare: Hansen method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 6),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(lwd = 0.1, alpha = 0.5)
```
### power=1.5

```{r}
#| fig-width: 12
#| fig-height: 7
#| echo: false
mapex <- st_bbox(hexagons)

tmap_mode("plot")
tm_shape(hexagon_Hansen_15,
         bbox = mapex) + 
  tm_fill(col = "accHansen_15",
          n = 10,
          style = "quantile",
          border.col = "black",
          border.lwd = 1) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to eldercare: Hansen method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 6),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(lwd = 0.1, alpha = 0.5)
```

### power=2

```{r}
#| fig-width: 12
#| fig-height: 7
#| echo: false
mapex <- st_bbox(hexagons)

tmap_mode("plot")
tm_shape(hexagon_Hansen_20,
         bbox = mapex) + 
  tm_fill(col = "accHansen_20",
          n = 10,
          style = "quantile",
          border.col = "black",
          border.lwd = 1) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to eldercare: Hansen method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 6),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(lwd = 0.1, alpha = 0.5)
```

### power=2.5

```{r}
#| fig-width: 12
#| fig-height: 7
#| echo: false
mapex <- st_bbox(hexagons)

tmap_mode("plot")
tm_shape(hexagon_Hansen_25,
         bbox = mapex) + 
  tm_fill(col = "accHansen_25",
          n = 10,
          style = "quantile",
          border.col = "black",
          border.lwd = 1) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to eldercare: Hansen method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 6),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(lwd = 0.1, alpha = 0.5)
```
::::

------------------------------------------------------------------------

## Statistical graphic

::::: panel-tabset

### power=0.5

```{r}
hexagon_Hansen_05 <- st_join(hexagon_Hansen_05, mpsz, 
                          join = st_intersects)
```

```{r}
#| fig-width: 12
#| fig-height: 7
#| echo: false

ggbetweenstats(
  data = hexagon_Hansen_05,
  x = REGION_N,
  y = accHansen_05,
  type = "p")
```

### power=1

```{r}
hexagon_Hansen_10 <- st_join(hexagon_Hansen_10, mpsz, 
                          join = st_intersects)
```

```{r}
#| fig-width: 12
#| fig-height: 7
#| echo: false

ggbetweenstats(
  data = hexagon_Hansen_10,
  x = REGION_N,
  y = accHansen_10,
  type = "p")
```

### power=1.5

```{r}
hexagon_Hansen_15 <- st_join(hexagon_Hansen_15, mpsz, 
                          join = st_intersects)
```

```{r}
#| fig-width: 12
#| fig-height: 7
#| echo: false

ggbetweenstats(
  data = hexagon_Hansen_15,
  x = REGION_N,
  y = accHansen_15,
  type = "p")
```

### power=2

```{r}
hexagon_Hansen_20 <- st_join(hexagon_Hansen_20, mpsz, 
                          join = st_intersects)
```

```{r}
#| fig-width: 12
#| fig-height: 7
#| echo: false

ggbetweenstats(
  data = hexagon_Hansen_20,
  x = REGION_N,
  y = accHansen_20,
  type = "p")
```

### power=2.5

```{r}
hexagon_Hansen_25 <- st_join(hexagon_Hansen_25, mpsz, 
                          join = st_intersects)
```

```{r}
#| fig-width: 12
#| fig-height: 7
#| echo: false

ggbetweenstats(
  data = hexagon_Hansen_25,
  x = REGION_N,
  y = accHansen_25,
  type = "p")
```

:::::