---
title: "Take-home Exercise 1 | Geospatial Analytics for Public Good"
subtitle: "In this take-home exercise, we are interested to investigate the spatial and spatio-temporal patterns of new businesses and firms established in Singapore for the first six months of 2025."
date: "Sep 13, 2025"
date-modified: "last-modified"
author: Linh Nguyen
execute:
  echo: true
  eval: true
  freeze: auto
  message: false
  warning: false
format:
  html:
    code-link: true        # Clickable function links
    toc: true             # Table of contents
    code-fold: true       # Foldable code
    code-tools: true      # Code tools menu
number-sections: true
number-offset: 1
editor: visual
---

## Overview

### Objectives

-   To identify where and when new businesses are established in Singapore, revealing clusters, patterns, and trends
-   To uncover complex relationships in how new businesses are distributed across the island

## Analysis Questions

- What spatial distribution patterns do different business types exhibit across Singapore?
- How do survival rates vary between business categories and specific SSIC codes?
- Which business types show clustering behavior versus random spatial distribution?
- What factors influence business location density and geographic concentration patterns?

## Set up R environment

```{r}
# Import R Packages
pacman::p_load(tidyverse, sf, tmap, httr, readxl, dplyr, stringr, rvest, knitr, spatstat, terra, raster, sparr, tidyr, kableExtra, ggplot2)
```

```{r}
set.seed(1234)
```

### The Data

+-------------------------------------------------------------+----------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------+------------+
| Dataset                                                     | Description                                                                                        | Source                                                                                                                                       | Format     |
+=============================================================+====================================================================================================+==============================================================================================================================================+============+
| **ACRA Information on Corporate Entities**                  | Information of registered entities in Singapore, including location data                           | [data.gov.sg](https://data.gov.sg/datasets?query=ACRA+Information+on+Corporate+Entities&page=1&resultId=2)                                   | CSV        |
+-------------------------------------------------------------+----------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------+------------+
| **SSIC: Singapore Standard Industrial Classification 2020** | List of codes represented economic activity of business establishments in Singapore, 2020 version. | [SSIC: Singapore Standard Industrial Classification](https://classification.codes/classifications/industry/ssic-singapore)                   | Excel      |
+-------------------------------------------------------------+----------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------+------------+
| **MasterPlan2019SubzoneBoundaryNoSeaKML**                   | Polygon data with URA 2019 Master Plan Planning Subzone boundaries                                 | [data.gov.sg](https://data.gov.sg/datasets?query=Master+Plan+Planning+Subzone+boundaries&page=1&resultId=d_6c6d7361dd826d97b91bac914ca6b2ac) | KML        |
+-------------------------------------------------------------+----------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------+------------+

## Data Import & Preparation

### ACRA Data

```{r}
#| eval: false
folder_path <- "data/aspatial/ACRA"
file_list <- list.files(path = folder_path, 
                        pattern = "^ACRA*.*\\.csv$", 
                        full.names = TRUE)

acra_data <- file_list %>%
  map_dfr(read_csv)
```

```{r}
#| echo: false
#| eval: false
write_rds(acra_data, "data/rds/acra_data.rds")
```

```{r}
#| echo: false
#| eval: false
acra_data <- read_rds("data/rds/acra_data.rds")
```

#### Extracting data of chosen business types

The chosen business types are as the following table:

```{r}
#| echo: false

# Create the data frame
ssic_table <- data.frame(
  `SSIC 3-digit` = c(
    "471",
    "",
    "",
    "",
    "561", 
    "",
    "",
    "",
    "",
    "",
    "",
    "",
    "",
    "681",
    "",
    "",
    "",
    "",
    ""
  ),
  Description = c(
    "Retail Sale in Non-Specialised Stores",
    "",
    "",
    "",
    "Restaurants, Cafes, Food Courts, Food Kiosks And Bars",
    "",
    "",
    "",
    "",
    "",
    "",
    "",
    "",
    "Real Estate Activities With Own Or Leased Property",
    "",
    "",
    "",
    "",
    ""
  ),
  Alias = c(
    "**Retail**",
    "",
    "",
    "",
    "**Food & Beverage**",
    "",
    "",
    "",
    "",
    "",
    "",
    "",
    "",
    "**Real Estate**",
    "",
    "",
    "",
    "",
    ""
  ),
  `SSIC 4-digit` = c(
    "4710: Retail Sale in Non-Specialised Stores",
    "",
    "",
    "",
    "5611: Restaurants and Cafes",
    "5612: Fast Food Outlets, Food Courts and Food Kiosks",
    "5613: Pubs", 
    "5614: Stalls Selling Cooked Food and Prepared Drinks",
    "",
    "",
    "",
    "",
    "",
    "6810: Real Estate Activities With Own Or Leased Property",
    "",
    "",
    "",
    "",
    ""
  ),
  `SSIC 5-digit` = c(
    "47101: Supermarkets and hypermarkets",
    "47102: Mini-marts, convenience stores and provision shops",
    "47103: Department stores", 
    "47104: Retail sale in other non-specialised stores n.e.c.",
    "56111: Restaurants",
    "56112: Cafes",
    "56121: Fast food outlets",
    "56122: Food courts, coffee shops and canteens (with mainly food and beverage income)",
    "56123: Food kiosks mainly for takeaway and delivery",
    "56130: Pubs",
    "56140: Stalls Selling Cooked Food and Prepared Drinks",
    "",
    "",
    "68101: Real estate developers",
    "68102: Operating of serviced apartments",
    "68103: Letting of self-owned or leased real estate property except food courts, coffee shops and canteens",
    "68104: Letting and operating of self-owned or leased food courts, coffee shops and canteens",
    "68105: Collective portfolio investment funds with rental income",
    "68106: Management of self-owned strata titled property (i.e. Management Corporation Strata Title)"
  ),
  stringsAsFactors = FALSE
)

# Create the kable
ssic_table %>%
  kable(
    col.names = c("SSIC 3-digit prefix", "Description", "Alias for this analysis", 
                  "SSIC 4-digit sub-types", "SSIC 5-digit sub-types"),
    caption = "SSIC Classification Structure for Analysis",
    escape = FALSE  # Allow HTML formatting like **bold**
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    font_size = 11,
    full_width = TRUE
  ) %>%
  # Group rows by main categories
  pack_rows("Retail (471)", 1, 4, label_row_css = "background-color: #f0f0f0; font-weight: bold;") %>%
  pack_rows("Food & Beverage (561)", 5, 13, label_row_css = "background-color: #f0f0f0; font-weight: bold;") %>%
  pack_rows("Real Estate (681)", 14, 19, label_row_css = "background-color: #f0f0f0; font-weight: bold;") %>%
  # Style specific columns
  column_spec(1, bold = TRUE, width = "8%") %>%
  column_spec(2, width = "25%") %>%
  column_spec(3, bold = TRUE, width = "12%") %>%
  column_spec(4, width = "25%") %>%
  column_spec(5, width = "30%") %>%
  # Add scroll for better mobile viewing
  scroll_box(width = "100%", height = "600px")
```

```{r}
#| label: Extracting data of chosen business types
#| echo: false
#| eval: false
biz_selected <- acra_data %>%
  select(1:24) %>%
  filter(str_detect(primary_ssic_code, "^(561|471|681)")) %>%
  rename(date = registration_incorporation_date) %>%
  mutate(date = as.Date(date),
         YEAR = year(date),
         MONTH_NUM = month(date),
         MONTH_ABBR = month(date, 
                            label = TRUE, 
                            abbr = TRUE)) %>% 
  mutate(
    postal_code = str_pad(postal_code, 
    width = 6, side = "left", pad = "0")) %>%
  filter(YEAR >= 2024 & YEAR <= 2025) %>%  
  filter(date >= as.Date("2024-01-01") & date <= as.Date("2025-06-30")) 
```

```{r}
#| echo: false
#| eval: false
write_rds(biz_selected, "data/rds/biz_selected.rds")
```

```{r}
#| echo: false
#| eval: false
biz_selected <- read_rds("data/rds/biz_selected.rds")
```

### SSIC Data

```{r}
# Read the Excel file
ssic_data <- read_excel("data/ssic2020-classification-structure.xlsx", 
                        sheet = "SSIC 2020 Structure",
                        skip = 4)  # Skip the header rows

# Clean the data - remove any rows where SSIC 2020 is NA
ssic_data <- ssic_data %>%
  filter(!is.na(`SSIC 2020`))

# Filter to get only level 5 codes (5-character codes)
level5_data <- ssic_data %>%
  filter(str_length(`SSIC 2020`) == 5)

# Create the hierarchy columns
ssic_hierarchy <- level5_data %>%
  mutate(
    # Level 2: First 2 characters (2-digit number)
    level_2 = str_sub(`SSIC 2020`, 1, 2),
    
    # Level 3: First 3 characters (3-digit number)
    level_3 = str_sub(`SSIC 2020`, 1, 3),
    
    # Level 4: First 4 characters (4-digit number)
    level_4 = str_sub(`SSIC 2020`, 1, 4),
    
    # Level 5: Full 5-character code
    level_5 = `SSIC 2020`,
    
    # Keep the title for reference
    level_5_description = `SSIC 2020 Title`
  ) %>%
  dplyr::select(level_2, level_3, level_4, level_5, level_5_description)
```

```{r}
# Add the descriptions for each level by joining with the original data
# Create lookup tables for each level
level_2_lookup <- ssic_data %>%
  filter(str_length(`SSIC 2020`) == 2) %>%
  dplyr::select(level_2 = `SSIC 2020`, level_2_description = `SSIC 2020 Title`)

level_3_lookup <- ssic_data %>%
  filter(str_length(`SSIC 2020`) == 3) %>%
  dplyr::select(level_3 = `SSIC 2020`, level_3_description = `SSIC 2020 Title`)

level_4_lookup <- ssic_data %>%
  filter(str_length(`SSIC 2020`) == 4) %>%
  dplyr::select(level_4 = `SSIC 2020`, level_4_description = `SSIC 2020 Title`)

# Join all the descriptions
ssic_final <- ssic_hierarchy %>%
  left_join(level_2_lookup, by = "level_2") %>%
  left_join(level_3_lookup, by = "level_3") %>%
  left_join(level_4_lookup, by = "level_4") %>%
  dplyr::select(level_2, level_2_description, 
         level_3, level_3_description,
         level_4, level_4_description,
         level_5, level_5_description)
```

```{r}
cat("\nSummary:\n")
cat("Total level 5 business types:", nrow(ssic_final), "\n")
cat("Number of level 2 categories:", length(unique(ssic_final$level_2)), "\n")
cat("Number of level 3 categories:", length(unique(ssic_final$level_3)), "\n")
cat("Number of level 4 categories:", length(unique(ssic_final$level_4)), "\n")
```


#### Join ACRA and SSIC Data

```{r}
#| eval: false
# Left join ssic_final to biz_selected
# Convert primary_ssic_code to character and pad with leading zeros if needed
biz_with_hierarchy <- biz_selected %>%
  mutate(primary_ssic_code = sprintf("%05d", primary_ssic_code)) %>%  # Ensures 5-digit format
  left_join(ssic_final, by = c("primary_ssic_code" = "level_5"))

# Check the result
str(biz_with_hierarchy)
head(biz_with_hierarchy)
```

```{r}
#| echo: false
#| eval: false
write_rds(biz_with_hierarchy, "data/rds/biz_with_hierarchy.rds")
```

```{r}
#| echo: false
biz_with_hierarchy <- read_rds("data/rds/biz_with_hierarchy.rds")
```

### Geospatial data

The Master Plan 2019 Boundary (No Sea) (KML) was downloaded from data.gov.sg that consists of the geographical boundary of Singapore at the planning subzone level. The data is based on URA Master Plan 2019. We will use this to plot the boundaries of the subzones.

::: panel-tabset
#### Import data

```{r}
#| eval: false
mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
```

#### Tidy data

```{r}
#| eval: false
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}
```

```{r}
#| eval: false
mpsz <- mpsz %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  dplyr::select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())

mpsz_cl <- mpsz %>%
  filter(SUBZONE_N != "SOUTHERN GROUP",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS")
```

```{r}
#| echo: false
mpsz_cl <- read_rds("data/rds/mpsz_cl.rds")
```
:::

### Geocoding the data

```{r}
#| eval: false
# Initiate 2 empty data frames
found <- data.frame() # to store Postal codes that were successfully geocoded with their coordinates
not_found <- data.frame(postcode = character()) # to store postal codes that couldn't be matched
```

```{r}
#| eval: false

postcodes <- unique(biz_selected$postal_code)

url <- "https://onemap.gov.sg/api/common/elastic/search"

for (pc in postcodes) {
  query <- list(
    searchVal = pc,
    returnGeom = "Y",
    getAddrDetails = "Y",
    pageNum = "1"
  )
  
  res <- GET(url, query = query)
  json <- content(res)
  
  if (json$found != 0) {
    df <- as.data.frame(json$results, stringsAsFactors = FALSE)
    df$input_postcode <- pc
    found <- bind_rows(found, df)
  } else {
    not_found <- bind_rows(not_found, data.frame(postcode = pc))
  }
}
```

```{r}
#| eval: false

# Tidying the geocoded data
found <- found %>%
  select(1:10)
```

### Appending the location information

```{r}
#| eval: false
biz_selected_geocoded = biz_with_hierarchy %>%
  left_join(found, 
            by = c('postal_code' = 'POSTAL'))
```

```{r}
#| eval: false
#| echo: false
write_rds(biz_selected_geocoded, "data/rds/biz_selected_geocoded.rds")
```

```{r}
#| echo: false
biz_selected_geocoded <- read_rds("data/rds/biz_selected_geocoded.rds")
```

### Create separate datasets for each business type

```{r}
#| eval: false
# Create separate datasets for each business type to count each type
biz_restaurant_cafe <- biz_with_hierarchy %>% filter(str_detect(primary_ssic_code, "^561"))
biz_retail_nonspec <- biz_with_hierarchy %>% filter(str_detect(primary_ssic_code, "^471"))    
biz_real_estate <- biz_with_hierarchy %>% filter(str_detect(primary_ssic_code, "^681"))
```

```{r}
#| eval: false
#| echo: false
write_rds(biz_restaurant_cafe, "data/rds/biz_restaurant_cafe.rds")
write_rds(biz_retail_nonspec, "data/rds/biz_retail_nonspec.rds")
write_rds(biz_real_estate, "data/rds/biz_real_estate.rds")
```

```{r}
#| echo: false
biz_restaurant_cafe <- read_rds("data/rds/biz_restaurant_cafe.rds")
biz_retail_nonspec <- read_rds("data/rds/biz_retail_nonspec.rds")
biz_real_estate <- read_rds("data/rds/biz_real_estate.rds")
```

### Converting into simple point features - SF data frame

Before applying the `st_as_sf` function to convert the data into simple point features, we check for missing values first and remove such records.

```{r}
biz_with_coords <- biz_selected_geocoded %>%
  # Remove records without coordinates
  filter(!is.na(X) & !is.na(Y)) %>%
  # Convert LATITUDE and LONGITUDE to numeric if they're character
  mutate(
    LATITUDE = as.numeric(X),
    LONGITUDE = as.numeric(Y)
  )
```

```{r}
biz_sf <- st_as_sf(biz_with_coords, 
                   coords = c("X","Y"), 
                   crs=3414)
```

```{r}
#| eval: false
# Create separate datasets for each business type for later individual analysis
biz_restaurant_cafe_sf <- biz_sf %>% filter(str_detect(primary_ssic_code, "^561"))
biz_retail_nonspec_sf <- biz_sf %>% filter(str_detect(primary_ssic_code, "^471"))    
biz_real_estate_sf <- biz_sf %>% filter(str_detect(primary_ssic_code, "^681"))
```

```{r}
#| eval: false
#| echo: false
# Save point features as GeoPackage for future use
st_write(biz_sf, "data/sf/biz_sf.gpkg", delete_dsn = TRUE)
st_write(biz_restaurant_cafe_sf, "data/sf/biz_restaurant_cafe_sf.gpkg", delete_dsn = TRUE)
st_write(biz_retail_nonspec_sf, "data/sf/biz_retail_nonspec_sf.gpkg", delete_dsn = TRUE)
st_write(biz_real_estate_sf, "data/sf/biz_real_estate_sf.gpkg", delete_dsn = TRUE)
```

```{r}
#| echo: false
biz_sf <- st_read("data/sf/biz_sf.gpkg")
biz_restaurant_cafe_sf <- st_read("data/sf/biz_restaurant_cafe_sf.gpkg")
biz_retail_nonspec_sf <- st_read("data/sf/biz_retail_nonspec_sf.gpkg")
biz_real_estate_sf <- st_read("data/sf/biz_real_estate_sf.gpkg")
```

### Verify the point features

::: panel-tabset

#### Retail

```{r}
plot(st_geometry(biz_retail_nonspec_sf), main = "Business Point Features", pch = 16, cex = 0.5)
```

#### Food & Beverages

```{r}
plot(st_geometry(biz_restaurant_cafe_sf), main = "Business Point Features", pch = 16, cex = 0.5)
```
#### Real Estate

```{r}
plot(st_geometry(biz_real_estate_sf), main = "Business Point Features", pch = 16, cex = 0.5)
```
:::

## Exploratory Data Analysis

::: panel-tabset

### Summary on the chosen business types

```{r}
# Step 1: Create summary table from biz_with_hierarchy with both total and live counts
biz_summary <- biz_with_hierarchy %>%
  group_by(primary_ssic_code) %>%
  summarise(
    business_count = n(),
    live_business_count = sum(entity_status_description == "Live", na.rm = TRUE),
    live_percentage = round(live_business_count / business_count * 100, 1),
    .groups = 'drop'
  ) %>%
  filter(!is.na(primary_ssic_code))

# Step 2: Join with ssic_final on primary_ssic_code = level_5
biz_hierarchy_summary <- biz_summary %>%
  left_join(ssic_final, by = c("primary_ssic_code" = "level_5"))

# Step 3: Create concatenated descriptions
biz_pivot_table <- biz_hierarchy_summary %>%
  mutate(
    level_3_concat = paste(level_3, level_3_description, sep = " - "),
    level_4_concat = paste(level_4, level_4_description, sep = " - "),
    level_5_concat = paste(primary_ssic_code, level_5_description, sep = " - ")
  ) %>%
  arrange(level_3_concat, level_4_concat, level_5_concat)

# Step 4: Keep only the columns we need
biz_pivot_table <- biz_pivot_table[, c("level_3_concat", "level_4_concat", "level_5_concat", 
                                       "business_count", "live_business_count", "live_percentage")]

# Step 5: Display with kable including the new column
biz_pivot_table %>%
  kable(caption = "Business Distribution by SSIC Hierarchy",
        col.names = c("Level 3", "Level 4", "Level 5", "Total", "Live", "Live %")) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(4:6, bold = TRUE) %>%
  pack_rows(index = table(biz_pivot_table$level_3_concat)) %>%
  scroll_box(height = "500px")
```

Based on the business summary above, we can draw some key insights about business types and survival rates:

-   Food & Beverage dominates with 5,542 businesses (69.1% of total), reflecting Singapore's vibrant food scene but challenging market conditions with 27.4% survival rate
-   Traditional food stalls excel (52.3% survival) vs. restaurants struggle (8% survival), suggesting lower overhead and established customer bases favor traditional formats
-   Real estate sector shows poor resilience with only 6.7% survival rate, likely due to high capital requirements and market volatility affecting property investments
-   Retail stores perform moderately (25.9% survival) with department stores outperforming (43.3%) supermarkets (13.8%), indicating differentiated positioning beats commodity retail
-   Overall 24.4% survival rate suggests high business turnover in Singapore, reflecting competitive market dynamics and regulatory compliance challenges across sectors

**Key insight:** Asset-light, established food formats and specialized retail show better survival than capital-intensive businesses.

### Plotting business points on Singapore boundary

As the number of business points per business type is large itself, plotting all of them on `mpsz` - Singapore boundary will be too packed and we will not be able to extract any pattern out of it. Therefore, it is a good idea to plot them separately as below.

```{r}
ggplot() +
  geom_sf(data = mpsz_cl, fill = "lightgray", color = "white") +
  geom_sf(data = biz_sf, aes(color = as.factor(primary_ssic_code)), 
          size = 0.3, alpha = 0.7) +
  facet_wrap(~ primary_ssic_code) +  # Remove scales = "free"
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none") +  # Hide legend since facets show the codes
  labs(title = "Business Locations by SSIC Code")
```

Based on the business location plots, here are key observations about spatial density patterns:

**Food & Beverage Sector (Food & Beverages):**

-   56111 (Restaurants) and 56140 (Food stalls) show the highest density and widest distribution across Singapore, indicating ubiquitous presence in residential and commercial areas
-   56122 (Food courts/coffee shops) displays concentrated clusters, likely in HDB estates and commercial centers where these establishments serve local communities
-   56123 (Food kiosks) shows moderate density with scattered distribution, reflecting their mobile/flexible nature
-   56112 (Cafes) and 56130 (Pubs) are more selectively located, probably in trendy districts and entertainment areas

**Retail Sector (Retail):**

-   47109 (Other non-specialised stores) shows broad distribution but lower density than food businesses
-   47101-47103 display very sparse, clustered patterns, suggesting these larger retail formats locate in specific commercial zones

**Real Estate Sector (Real Estate):**

- 68103 (Property letting) shows moderate clustering in business districts Other real estate activities (68101, 68102, 68104-68106) are extremely sparse, reflecting their specialized, capital-intensive nature

:::

## CORE SPATIAL ANALYSIS - REGISTRATION INTENT ANALYSIS

### Spatial KDE Analysis

#### Individual Density Surfaces

**Create Owin object**

```{r}
sg_owin <- as.owin(mpsz_cl)
```

:::panel-tabset

##### Retail

```{r}
# Convert sf data frames to ppp class
ppp_retail_all <- as.ppp(biz_retail_nonspec_sf)
ppp_retail_all = ppp_retail_all[sg_owin]

# Calculate KDE using spatstat
kde_retail_all <- density(ppp_retail_all,
                         sigma = bw.diggle,
                         edge = TRUE,
                         kernel="gaussian")
plot(kde_retail_all)
```

##### Food & Beverages

```{r}
# Convert sf data frames to ppp class
ppp_restaurants_all <- as.ppp(biz_restaurant_cafe_sf)
ppp_restaurants_all = ppp_restaurants_all[sg_owin]

# Calculate KDE using spatstat
kde_restaurants_all <- density(ppp_restaurants_all, 
                              sigma = bw.diggle,  # Automatic bandwidth selection
                              edge = TRUE,
                              kernel="gaussian")        # Edge correction
plot(kde_restaurants_all)
```

##### Real Estate

```{r}
# Convert sf data frames to ppp class
ppp_realestate_all <- as.ppp(biz_real_estate_sf)
ppp_realestate_all = ppp_realestate_all[sg_owin]

# Calculate KDE using spatstat
kde_realestate_all <- density(ppp_realestate_all,
                             sigma = bw.diggle,
                             edge = TRUE,
                             kernel="gaussian")
plot(kde_realestate_all)
```

:::

#### Perform the Clark-Evans test with CSR

In this section, we will perform Clark-Evans tests for the 3 business entities separately, to quantify whether a point pattern is clustered, random, or uniformly spaced. The result will help us understand the underlying spatial patterns of each business type. This will inform our interpretation of the subsequent KDE analysis.

The confidence level chosen for the test is 95%. Therefore, we can set the number simulation at the minimum of `nsim = 39`.

::: panel-tabset

##### Food & Beverages

The test hypotheses are: 
- Ho = The distribution of Food & Beverages are randomly distributed. 
- H1 = The distribution of Food & Beverages are not randomly distributed. The 95% confident interval will be used.

```{r}
#| cache: true
clarkevans.test(ppp_restaurants_all,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                method="MonteCarlo",
                nsim=39)
```

The Clark-Evans test reveals that restaurants and cafes (SSIC Food & Beverages) in the study area exhibit a significantly clustered spatial distribution (R = 0.35908, p < 0.05). This strong clustering pattern suggests that food service establishments tend to co-locate rather than distribute randomly across the landscape. The clustering likely reflects agglomeration benefits where restaurants benefit from proximity to competitors through shared customer traffic, as well as urban planning policies that concentrate commercial activities in designated zones. This spatial concentration creates identifiable dining districts and food hubs within the study area, which has important implications for business location strategy and urban commercial planning.

##### Retail

The test hypotheses are: 
- Ho = The distribution of Retail are randomly distributed. 
- H1 = The distribution of Retail are not randomly distributed. The 95% confident interval will be used.

```{r}
#| cache: true
clarkevans.test(ppp_retail_all,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                method="MonteCarlo",
                nsim=39)
```

The Clark-Evans test indicates that non-specialised retail stores (SSIC Retail) exhibit a significantly clustered spatial distribution (R = 0.46, p < 0.05). While still showing clustering behavior, these retail establishments are somewhat more dispersed than restaurants and cafes, reflecting their need to serve broader geographic markets while still benefiting from commercial agglomeration in shopping districts and retail centers.

##### Real Estate

The test hypotheses are: 
- Ho = The distribution of Real Estate are randomly distributed. 
- H1 = The distribution of Real Estate are not randomly distributed. The 95% confident interval will be used.

```{r}
#| cache: true
clarkevans.test(ppp_realestate_all,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                method="MonteCarlo",
                nsim=39)
```

The Clark-Evans test shows that real estate businesses (SSIC Real Estate) exhibit a significantly clustered spatial distribution (R = 0.45, p < 0.05). This clustering pattern suggests that real estate agencies tend to co-locate in commercial districts and business centers, likely benefiting from proximity to financial services, legal offices, and high foot traffic areas where property transactions commonly occur.
:::

#### Hotspot Identification using spatstat quantile method

```{r}
# Extract quantile-based hotspots
hotspot_threshold <- 0.95  # Top 5%

# For restaurants
restaurants_quantile <- quantile(kde_restaurants_all, probs = hotspot_threshold)
hotspots_restaurants_mask <- kde_restaurants_all >= restaurants_quantile

# For retail  
retail_quantile <- quantile(kde_retail_all, probs = hotspot_threshold)
hotspots_retail_mask <- kde_retail_all >= retail_quantile

# For real estate
realestate_quantile <- quantile(kde_realestate_all, probs = hotspot_threshold)
hotspots_realestate_mask <- kde_realestate_all >= realestate_quantile

# Plot hotspots overlay
plot(kde_restaurants_all, main = "Registration Intent Hotspots", ribbon = FALSE)
contour(hotspots_restaurants_mask, add = TRUE, col = "red", lwd = 2)
contour(hotspots_retail_mask, add = TRUE, col = "yellow", lwd = 2) 
contour(hotspots_realestate_mask, add = TRUE, col = "green", lwd = 2)
legend("topright", 
       legend = c("Restaurants", "Retail", "Real Estate"),
       col = c("red", "yellow", "green"),
       lty = 1, lwd = 2)
```

### Spatio-Temporal KDE Analysis

#### 2025 Monthly Progression

```{r}
# Function to create monthly ppp objects and KDE
create_monthly_ppp_kde <- function(data_sf, year_filter = 2025) {
  monthly_results <- list()
  
  for(month in 1:6) {
    # Filter sf object by year and month
    monthly_data_sf <- data_sf %>%
      filter(YEAR == year_filter, MONTH_NUM == month)
    
    if(nrow(monthly_data_sf) >= 10) {  # Minimum sample size
      # Create ppp object from sf (your method)
      monthly_ppp <- as.ppp(monthly_data_sf)
      monthly_ppp <- monthly_ppp[sg_owin]  
      
      # Calculate KDE using spatstat
      monthly_kde <- density(monthly_ppp, sigma = bw.diggle, edge = TRUE)
      
      monthly_results[[paste0("month_", month)]] <- list(
        ppp = monthly_ppp,
        kde = monthly_kde,
        count = nrow(monthly_data_sf)
      )
    }
  }
  
  return(monthly_results)
}
```

```{r}
# Function to plot monthly progression
plot_monthly_kde <- function(monthly_list, title) {
  par(mfrow = c(2, 3), mar = c(3, 3, 3, 1))
  for(i in 1:6) {
    month_name <- paste0("month_", i)
    if(month_name %in% names(monthly_list)) {
      plot(monthly_list[[month_name]]$kde, 
           main = paste0(title, "\nMonth ", i, " (n=", monthly_list[[month_name]]$count, ")"),
           ribbon = FALSE, cex.main = 0.8) 
    } else {
      plot.new()
      text(0.5, 0.5, paste("Month", i, "\nInsufficient data"), 
           cex = 1.2, adj = c(0.5, 0.5))
    }
  }
  par(mfrow = c(1, 1))
}
```

```{r}
# Create monthly analysis for each business type
monthly_retail <- create_monthly_ppp_kde(biz_retail_nonspec_sf)
monthly_restaurants <- create_monthly_ppp_kde(biz_restaurant_cafe_sf)
monthly_realestate <- create_monthly_ppp_kde(biz_real_estate_sf)
```

::: panel-tabset

##### Retail

```{r fig.width=12, fig.height=8}
#| cache: true
plot_monthly_kde(monthly_retail, "Retail Business Registration")
```
Retail registrations fluctuate between 57-92 establishments monthly in early 2025, showing more variation than restaurants. Primary hotspot concentrates in central Singapore's CBD and Orchard Road commercial belt, with strong secondary clusters in northeastern areas likely Tampines/Bedok and northwestern regions around Jurong. This dispersed pattern reflects retail's need to serve residential populations across multiple regional commercial centers.

##### Food & Beverages

```{r fig.width=12, fig.height=8}
plot_monthly_kde(monthly_restaurants, "Restaurant Registration")
```
The visualization above represents restaurant registrations in the first half of 2025, which show stable patterns with 283-344 monthly establishments consistently concentrating in the CBD and central urban core. Spatial distribution remains unchanged across months, indicating established downtown commercial districts maintain attractiveness for new ventures with minimal geographic diversification into outer residential areas.

##### Real Estate

```{r fig.width=12, fig.height=8}
plot_monthly_kde(monthly_realestate, "Real Estate Business Registration")
```
Real estate registrations vary between 61-74 establishments monthly across the first half of 2025. Activity concentrates heavily in the central business district and downtown core, with prominent hotspots in the CBD area and Marina Bay region. Secondary clusters appear in northern Singapore, likely Woodlands/Yishun areas, indicating diversified real estate development beyond traditional commercial centers.
:::

#### 2024 vs 2025 Year-over-Year Comparison

```{r}
# Create year comparison function
create_annual_ppp_comparison <- function(data_sf) {
  annual_results <- list()
  
  # 2024 data (first half)
  data_2024_h1_sf <- data_sf %>%
    filter(YEAR == 2024, MONTH_NUM %in% 1:6)
  
  # 2025 data (first half)
  data_2025_h1_sf <- data_sf %>%
    filter(YEAR == 2025, MONTH_NUM %in% 1:6)
  
  if(nrow(data_2024_h1_sf) >= 10) {
    ppp_2024 <- as.ppp(data_2024_h1_sf)
    ppp_2024 <- ppp_2024[sg_owin]  # Apply Singapore window
    
    annual_results$year_2024 <- list(
      ppp = ppp_2024,
      kde = density(ppp_2024, sigma = bw.diggle, edge = TRUE),
      count = nrow(data_2024_h1_sf)
    )
  }
  
  if(nrow(data_2025_h1_sf) >= 10) {
    ppp_2025 <- as.ppp(data_2025_h1_sf)
    ppp_2025 <- ppp_2025[sg_owin]  # Apply Singapore window
    
    annual_results$year_2025 <- list(
      ppp = ppp_2025,
      kde = density(ppp_2025, sigma = bw.diggle, edge = TRUE),
      count = nrow(data_2025_h1_sf)
    )
  }
  
  return(annual_results)
}
```

```{r}
# Create a function to plot year-over-year comparison
plot_annual_comparison <- function(annual_list, title) {
  if(length(annual_list) == 2) {
    par(mfrow = c(1, 2), mar = c(3, 3, 3, 1))
    
    plot(annual_list$year_2024$kde, 
         main = paste0(title, "\n2024 H1 (n=", annual_list$year_2024$count, ")"),
         ribbon = TRUE, 
         cex.main = 1,
         ribargs = list(cex.axis = 0.7))  # Adjust ribbon text size
    
    plot(annual_list$year_2025$kde,
         main = paste0(title, "\n2025 H1 (n=", annual_list$year_2025$count, ")"),
         ribbon = TRUE, 
         cex.main = 1,
         ribargs = list(cex.axis = 0.7))  # Adjust ribbon text size
    
    par(mfrow = c(1, 1))
  }
}
```

```{r}
# Annual comparison for each business type (using sf objects)
annual_restaurants <- create_annual_ppp_comparison(biz_restaurant_cafe_sf)
annual_retail <- create_annual_ppp_comparison(biz_retail_nonspec_sf)
annual_realestate <- create_annual_ppp_comparison(biz_real_estate_sf)
```

Based on the visualization below, Year-over-year analysis reveals significant growth and spatial diversification across all three business sectors in H1 2025. More notably, the spatial distribution patterns demonstrate a clear shift from concentrated urban clustering to geographic diversification. The intense density hotspots visible in H1 2024—particularly around the CBD and Marina Bay areas—have diffused considerably in 2025, indicating businesses are expanding into secondary commercial districts and suburban locations. This spatial deconcentration suggests either market saturation in prime central areas driving entrepreneurs to seek opportunities in emerging neighborhoods, or improved infrastructure and zoning policies making previously less attractive areas more viable for business establishment. 

::: panel-tabset

##### Retail

```{r}
plot_annual_comparison(annual_retail, "Retail Registration")
```

##### Food & Beverages

```{r}
plot_annual_comparison(annual_restaurants, "Restaurant Registration")
```

##### Real Estate

```{r}
plot_annual_comparison(annual_realestate, "Real Estate Registration")
```

:::

## Second Order Spatial Point Patterns Analysis

This section is to examine spatial relationships beyond simple density mapping to reveal underlying clustering processes and business interaction patterns. My approach for this analysis is as follows:

- F-function Analysis: Measures empty space accessibility from random locations to identify underserved areas
- L-function Analysis: Detects clustering intensity using the clear above/below envelope interpretation
- Cross L-functions: Quantify inter-business type spatial interactions to determine attraction, repulsion, or independence between business types
- Combined F-L Analysis: Provides comprehensive pattern validation through dual complementary perspectives on spatial distribution

### L-Function Analysis

I employ L-function analysis for its superior interpretability over K-functions, where deviations above/below the theoretical line clearly indicate clustering or regularity. 

The Complete Spatial Randomness test with nsim = 39 provides statistically valid significance testing (p < 0.05) while maintaining computational efficiency for this exploratory spatial analysis of business registration patterns.

::: panel-tabset

##### Retail

Calculate L-functions with border correction method, then plot

```{r}
#| cache: true
L_retail_all = Lest(ppp_retail_all, correction = "border")
plot(L_retail_all, . -r ~ r, 
     ylab= "L(d)-r", xlab = "d(m)")
```

Performing Complete Spatial Randomness Test

```{r}
#| cache: true
l_env_retail.csr <- envelope(ppp_retail_all, Lest, nsim = 39,
                             correction = "border", rank = 1, glocal=TRUE)
```

##### Food & Beverages

Calculate L-functions with border correction method, then plot

```{r}
#| cache: true
L_restaurants_all = Lest(ppp_restaurants_all, correction = "border")
plot(L_restaurants_all, . -r ~ r, 
     ylab= "L(d)-r", xlab = "d(m)")
```

Performing Complete Spatial Randomness Test

```{r}
#| cache: true
l_env_restaurants.csr <- envelope(ppp_restaurants_all, Lest, nsim = 39,
                             correction = "border", rank = 1, glocal=TRUE)
```

##### Real Estate

Calculate L-functions with border correction method, then plot

```{r}
L_realestate_all = Lest(ppp_realestate_all, correction = "border")
plot(L_realestate_all, . -r ~ r, 
     ylab= "L(d)-r", xlab = "d(m)")
```

Performing Complete Spatial Randomness Test

```{r}
#| cache: true
l_env_realestate.csr <- envelope(ppp_realestate_all, Lest, nsim = 39,
                             correction = "border", rank = 1, glocal=TRUE)
```

:::

### F-Function Analysis

The F-function (Empty Space Function) measures the distance from randomly selected locations to the nearest business point, quantifying spatial accessibility and identifying areas with insufficient business coverage. Interpretation:

- Below envelope = Clustered pattern (less empty space, high accessibility)
- Above envelope = Regular pattern (more empty space, potential service gaps)
- Within envelope = Random distribution

::: panel-tabset

##### Retail

Computing F-function estimation, then plot

```{r}
#| cache: true
F_retail_all = Lest(ppp_retail_all, correction = "border")
plot(F_retail_all)
```

Performing Complete Spatial Randomness Test

```{r}
#| cache: true
F_retail_all.csr <- envelope(ppp_retail_all, Fest, nsim = 39, rank = 1, glocal=TRUE)
plot(F_retail_all.csr, . - r ~ r, ylab= "F(d)-r", xlab = "d(m)")
```

##### Food & Beverages

Computing F-function estimation, then plot

```{r}
#| cache: true
F_restaurants_all = Lest(ppp_restaurants_all, correction = "border")
plot(F_restaurants_all)
```

Performing Complete Spatial Randomness Test

```{r}
#| cache: true
F_restaurants_all.csr <- envelope(ppp_restaurants_all, Fest, nsim = 39, rank = 1, glocal=TRUE)
plot(F_restaurants_all.csr, . - r ~ r, ylab= "F(d)-r", xlab = "d(m)")
```

##### Real Estate

Computing F-function estimation, then plot

```{r}
#| cache: true
F_realestate_all = Lest(ppp_realestate_all, correction = "border")
plot(F_realestate_all)
```

Performing Complete Spatial Randomness Test

```{r}
#| cache: true
F_realestate_all.csr <- envelope(ppp_realestate_all, Fest, nsim = 39, rank = 1, glocal=TRUE)
plot(F_realestate_all.csr, . - r ~ r, ylab= "F(d)-r", xlab = "d(m)")
```

:::

### Cross L-Function Analysis (Inter-business Type Interactions)

The cross L-function envelope method tests spatial interactions between business types using Monte Carlo simulations to create confidence envelopes under spatial independence. The observed cross L-function is compared against these envelopes to determine statistical significance. Interpretation:

- Above envelope = positive interaction (attraction/co-location between business types)
- Below envelope = negative interaction (repulsion/avoidance between business types)
- Within envelope = spatial independence (no significant interaction)

First, we create **Multitype Point Pattern** for Cross-Function Analysis

```{r}
# Get coordinates and create data frame with marks
restaurants_coords <- st_coordinates(biz_restaurant_cafe_sf)
retail_coords <- st_coordinates(biz_retail_nonspec_sf) 
realestate_coords <- st_coordinates(biz_real_estate_sf)

# Combine all coordinates with business type marks
multitype_data <- data.frame(
  x = c(restaurants_coords[,1], retail_coords[,1], realestate_coords[,1]),
  y = c(restaurants_coords[,2], retail_coords[,2], realestate_coords[,2]),
  marks = c(rep("Restaurants", nrow(restaurants_coords)),
           rep("Retail", nrow(retail_coords)), 
           rep("RealEstate", nrow(realestate_coords)))
)

# Create multitype ppp object
ppp_multitype <- ppp(x = multitype_data$x,
                     y = multitype_data$y, 
                     marks = as.factor(multitype_data$marks),
                     window = sg_owin)

# Verify it's multitype
print(is.multitype(ppp_multitype))  # Should return TRUE
```

::: panel-tabset

#### Restaurants vs Retail

```{r}
#| cache: true
# Cross L-function calculation
cross_l_rest_retail <- Lcross(ppp_multitype, i = "Restaurants", j = "Retail",
                              correction = c("border"))

# Plot cross L-function
plot(cross_l_rest_retail, main = "Cross L-function: Restaurants vs Retail\n(Above line = Attraction)")

```

```{r}
#| cache: true
# Envelope test for statistical significance
cross_l_env_rest_retail <- envelope(ppp_multitype, Lcross, nsim = 39,
                                    i = "Restaurants", j = "Retail",
                                    correction = "border", verbose = FALSE)

# Plot cross L-function envelope
plot(cross_l_env_rest_retail, main = "Cross L Envelope: Restaurants ↔ Retail\n(Above = Attraction, Below = Repulsion)")

```

#### Restaurants vs Real Estate  

```{r}
#| cache: true
# Cross L-function calculation
cross_l_rest_realestate <- Lcross(ppp_multitype, i = "Restaurants", j = "RealEstate",
                                  correction = c("border"))

# Plot cross L-function
plot(cross_l_rest_realestate, main = "Cross L-function: Restaurants vs Real Estate\n(Above line = Attraction)")

```

```{r}
#| cache: true
# Envelope test for statistical significance
cross_l_env_rest_realestate <- envelope(ppp_multitype, Lcross, nsim = 39,
                                        i = "Restaurants", j = "RealEstate", 
                                        correction = "border", verbose = FALSE)

# Plot cross L-function envelope
plot(cross_l_env_rest_realestate, main = "Cross L Envelope: Restaurants ↔ Real Estate\n(Above = Attraction, Below = Repulsion)")

```

#### Retail vs Real Estate

```{r}
#| cache: true
# Cross L-function calculation
cross_l_retail_realestate <- Lcross(ppp_multitype, i = "Retail", j = "RealEstate", 
                                    correction = c("border"))

# Plot cross L-function
plot(cross_l_retail_realestate, main = "Cross L-function: Retail vs Real Estate\n(Above line = Attraction)")

```

```{r}
#| cache: true
# Envelope test for statistical significance
cross_l_env_retail_realestate <- envelope(ppp_multitype, Lcross, nsim = 39,
                                          i = "Retail", j = "RealEstate",
                                          correction = "border", verbose = FALSE)

# Plot cross L-function envelope
plot(cross_l_env_retail_realestate, main = "Cross L Envelope: Retail ↔ Real Estate\n(Above = Attraction, Below = Repulsion)")

```

:::

### Combined F-L Function Analysis Summary

The Combined F-L Function Analysis provides comprehensive spatial pattern assessment by examining business clustering from two complementary perspectives: 
- The L-function measures inter-business distances to detect clustering intensity (above envelope = clustered, below = regular)
- The F-function analyzes empty space accessibility from random locations (below envelope = dense coverage, above = sparse distribution). 

Together, they reveal complete spatial patterns: consistent results indicate strong clustering or regularity, while conflicting results suggest complex mixed patterns requiring detailed investigation. 

#### Create summary function for clustering assessment

```{r}
#| cache: true

assess_clustering <- function(l_env, f_env, business_type) {
  # Extract test results
  l_result <- if(any(l_env$obs > l_env$hi)) "CLUSTERED" else if(any(l_env$obs < l_env$lo)) "REGULAR" else "RANDOM"
  f_result <- if(any(f_env$obs < f_env$lo)) "CLUSTERED" else if(any(f_env$obs > f_env$hi)) "REGULAR" else "RANDOM"
  
  # Consistency check
  consistent <- l_result == f_result
  
  return(data.frame(
    Business_Type = business_type,
    L_Function = l_result,
    F_Function = f_result,
    Consistent = consistent,
    Pattern = if(consistent) l_result else "MIXED_SIGNALS"
  ))
}
```

#### Assess clustering for all business types

```{r}
#| cache: true

clustering_assessment <- bind_rows(
  assess_clustering(l_env_restaurants.csr, F_restaurants_all.csr, "Restaurants_561"),
  assess_clustering(l_env_retail.csr, F_retail_all.csr, "Retail_471"),
  assess_clustering(l_env_realestate.csr, F_realestate_all.csr, "RealEstate_681")
)

print("=== F-FUNCTION AND L-FUNCTION CLUSTERING ASSESSMENT ===")
print(clustering_assessment)
```

#### Cross-interaction assessment

```{r}
#| cache: true

assess_cross_interaction <- function(cross_l_env, pair_name) {
  interaction <- if(any(cross_l_env$obs > cross_l_env$hi)) "ATTRACTION" else 
                 if(any(cross_l_env$obs < cross_l_env$lo)) "REPULSION" else "INDEPENDENT"
  
  return(data.frame(
    Business_Pair = pair_name,
    Interaction_Type = interaction
  ))
}
```

#### Assess cross-interactions

```{r}
#| cache: true


interaction_assessment <- bind_rows(
  assess_cross_interaction(cross_l_env_rest_retail, "Restaurants ↔ Retail"),
  assess_cross_interaction(cross_l_env_rest_realestate, "Restaurants ↔ Real Estate"),
  assess_cross_interaction(cross_l_env_retail_realestate, "Retail ↔ Real Estate")
)

print("\n=== CROSS-INTERACTION ASSESSMENT ===")
print(interaction_assessment)

```

## Conclusions and Future Improvements

### Conclusion

This comprehensive dual-layer spatial analysis of Singapore's business registration patterns from 2024-2025 provides crucial evidence-based insights for strategic urban land use planning and management. The consistent clustering patterns across restaurants (561), retail (471), and real estate (681) business types, particularly concentrated in Marina Bay and CBD areas, demonstrate the powerful influence of agglomeration economics on business location decisions. The F-function and L-function analyses reveal minimal empty space and strong clustering tendencies in these prime districts, indicating market-driven spatial optimization that urban planners should acknowledge and support rather than resist.

The findings advocate for a **multi-tiered commercial development** strategy that balances intensification of successful business ecosystems with strategic decentralization to underserved areas. Marina Bay and CBD's dominance necessitates enhanced mixed-use zoning policies with higher plot ratios and vertical integration requirements to maximize land efficiency while maintaining sustainable density levels. Simultaneously, the identification of excessive empty space in peripheral regions like Yishun, Sembawang, and Pasir Ris presents opportunities for targeted satellite commercial hub development through strategic incentives and infrastructure investments.

The cross L-function results demonstrating attraction relationships between business types support integrated commercial corridor development, enabling Singapore to create seamless business ecosystems that leverage synergistic relationships. However, the intense clustering also raises legitimate concerns about market oversaturation, commercial rent inflation, and equitable access to services across the island. Urban planners must implement competition management policies including business diversity quotas and affordable commercial space preservation to maintain market health.

The registration intent versus business success comparison reveals that entrepreneur preferences align with viable business locations, suggesting market mechanisms effectively identify optimal commercial areas. This validates a planning approach that works with natural clustering tendencies while proactively developing alternative business districts to ensure sustainable, equitable urban growth. These evidence-based recommendations provide a robust framework for managing Singapore's commercial landscape evolution while maintaining its position as a leading global business hub.

### Acknowledged Limitations for Future Improvements

This study presents several methodological limitations that may affect the comprehensiveness of findings. 

- SSIC aggregation at 3-digit level potentially masks important intra-category spatial variations - for example, fine-dining restaurants (56111) versus fast-food outlets (56112) likely exhibit distinct clustering behaviors that remain undetected in the current 561-level analysis. 

- Temporal scope limitation to 18 months (Jan 2024-Jun 2025) may not capture long-term cyclical patterns or seasonal business formation trends.

- Scale dependency represents another constraint - the analysis focuses on citywide patterns without examining neighborhood-level clustering variations or micro-location factors influencing business success within identified hotspots.

