---
title: "Take-home Exercise 1 | Geospatial Analytics for Public Good"
subtitle: "In this take-home exercise, we are interested to investigate the spatial and spatio-temporal patterns of new businesses and firms established in Singapore for the first six months of 2025."
date: "Sep 13, 2025"
date-modified: "last-modified"
author: Linh Nguyen
execute:
  echo: true
  eval: true
  freeze: auto
  message: false
  warning: false
format:
  html:
    code-link: true
    toc: true
number-sections: true
number-offset: 1
editor: visual
---

## Overview

### Setting The Scene

Where businesses choose to locate plays a huge role in how Singapore continues to grow and develop as a city-state. Smart locations give companies advantages like easier access to skilled workers and excellent infrastructure, while also shaping different districts like the Central Business District, Jurong Innovation District, or Punggol Digital District, and making it easier for goods to move through our efficient transport networks and for employees to commute via MRT and buses.

Studying how businesses spread out across Singapore and change over time is crucial for urban planning by agencies like URA and JTC. This research helps planners understand how economic activity and our urban landscape develop together, enabling them to design better integrated developments, plan for future growth in areas like the Greater Southern Waterfront, use our limited land resources efficiently, and build infrastructure that supports both business competitiveness and residents' quality of life in our compact island nation.

### Objectives

-   To identify where and when new businesses are established in Singapore, revealing clusters, patterns, and trends that help inform urban planning, resource allocation, and economic development decisions.

-   To uncover complex relationships in how new businesses are distributed across the island - such as how related industries cluster together, how past business locations influence future ones, and how commercial districts develop structurally over time. These insights are essential for evidence-based urban planning and policy-making.

By understanding business locations and how their distribution evolves, Singapore's urban planners and agencies can gain valuable spatial insights, optimize city operations, and develop effective strategies to attract investment and promote sustainable growth across our city-state.

### The Data

| Dataset | Description | Source | Format |
|------------------|------------------|------------------|------------------|
| **ACRA Information on Corporate Entities** | Information of registered entities in Singapore, including location data | [data.gov.sg](https://data.gov.sg/datasets?query=ACRA+Information+on+Corporate+Entities&page=1&resultId=2) | CSV |
| **SSIC: Singapore Standard Industrial Classification 2020** | List of codes represented economic activity of business establishments in Singapore, 2020 version. | [SSIC: Singapore Standard Industrial Classification](https://classification.codes/classifications/industry/ssic-singapore) | Excel |

## Set up R environment

### Import R Packages

```{r}
pacman::p_load(tidyverse, sf, tmap, httr, readxl, dplyr, stringr, rvest, knitr, spatstat, terra, raster, sparr)
```

### Set Seed

```{r}
set.seed(1234)
```

## Data Import & Preparation

### ACRA Data

```{r}
#| eval: false
folder_path <- "data/aspatial/ACRA"
file_list <- list.files(path = folder_path, 
                        pattern = "^ACRA*.*\\.csv$", 
                        full.names = TRUE)

acra_data <- file_list %>%
  map_dfr(read_csv)
```

```{r}
#| echo: false
#| eval: false
write_rds(acra_data, "data/rds/acra_data.rds")
```

```{r}
#| echo: false
##| eval: false
acra_data <- read_rds("data/rds/acra_data.rds")
```

#### Extracting data of chosen business types

The chosen business types are:
- SSIC 561**: Restaurants, Cafes, Food Courts, Food Kiosks And Bars
- SSIC 471**: Retail Sale in Non-Specialised Stores
- SSIC 681**: Real Estate Activities With Own Or Leased Property

```{r}
#| label: Extracting data of chosen business types
#| echo: false
##| eval: false
biz_selected <- acra_data %>%
  select(1:24) %>%
  filter(str_detect(primary_ssic_code, "^(561|471|681)")) %>%
  rename(date = registration_incorporation_date) %>%
  mutate(date = as.Date(date),
         YEAR = year(date),
         MONTH_NUM = month(date),
         MONTH_ABBR = month(date, 
                            label = TRUE, 
                            abbr = TRUE)) %>% 
  mutate(
    postal_code = str_pad(postal_code, 
    width = 6, side = "left", pad = "0")) %>%
  filter(YEAR >= 2024 & YEAR <= 2025) %>%  
  filter(date >= as.Date("2024-01-01") & date <= as.Date("2025-06-30")) 
```

```{r}
#| echo: false
#| eval: false
write_rds(biz_selected, "data/rds/biz_selected.rds")
```

```{r}
# Read the files you just wrote
#| echo: false
biz_selected <- read_rds("data/rds/biz_selected.rds")
```

### SSIC Data Import

```{r}
# Read the Excel file
ssic_data <- read_excel("data/ssic2020-classification-structure.xlsx", 
                        sheet = "SSIC 2020 Structure",
                        skip = 4)  # Skip the header rows

# Clean the data - remove any rows where SSIC 2020 is NA
ssic_data <- ssic_data %>%
  filter(!is.na(`SSIC 2020`))

# Filter to get only level 5 codes (5-character codes)
level5_data <- ssic_data %>%
  filter(str_length(`SSIC 2020`) == 5)

# Create the hierarchy columns
ssic_hierarchy <- level5_data %>%
  mutate(
    # Level 2: First 2 characters (2-digit number)
    level_2 = str_sub(`SSIC 2020`, 1, 2),
    
    # Level 3: First 3 characters (3-digit number)
    level_3 = str_sub(`SSIC 2020`, 1, 3),
    
    # Level 4: First 4 characters (4-digit number)
    level_4 = str_sub(`SSIC 2020`, 1, 4),
    
    # Level 5: Full 5-character code
    level_5 = `SSIC 2020`,
    
    # Keep the title for reference
    level_5_description = `SSIC 2020 Title`
  ) %>%
  select(level_2, level_3, level_4, level_5, level_5_description)

# Add the descriptions for each level by joining with the original data
# Create lookup tables for each level
level_2_lookup <- ssic_data %>%
  filter(str_length(`SSIC 2020`) == 2) %>%
  select(level_2 = `SSIC 2020`, level_2_description = `SSIC 2020 Title`)

level_3_lookup <- ssic_data %>%
  filter(str_length(`SSIC 2020`) == 3) %>%
  select(level_3 = `SSIC 2020`, level_3_description = `SSIC 2020 Title`)

level_4_lookup <- ssic_data %>%
  filter(str_length(`SSIC 2020`) == 4) %>%
  select(level_4 = `SSIC 2020`, level_4_description = `SSIC 2020 Title`)

# Join all the descriptions
ssic_final <- ssic_hierarchy %>%
  left_join(level_2_lookup, by = "level_2") %>%
  left_join(level_3_lookup, by = "level_3") %>%
  left_join(level_4_lookup, by = "level_4") %>%
  select(level_2, level_2_description, 
         level_3, level_3_description,
         level_4, level_4_description,
         level_5, level_5_description)

# Display the structure
cat("Final dataset structure:\n")
str(ssic_final)

cat("\nFirst few rows:\n")
head(ssic_final, 10)

cat("\nSummary:\n")
cat("Total level 5 business types:", nrow(ssic_final), "\n")
cat("Number of level 2 categories:", length(unique(ssic_final$level_2)), "\n")
cat("Number of level 3 categories:", length(unique(ssic_final$level_3)), "\n")
cat("Number of level 4 categories:", length(unique(ssic_final$level_4)), "\n")
```

#### Join ACRA and SSIC Data

```{r}
# Left join ssic_final to biz_selected
# Convert primary_ssic_code to character and pad with leading zeros if needed
biz_with_hierarchy <- biz_selected %>%
  mutate(primary_ssic_code = sprintf("%05d", primary_ssic_code)) %>%  # Ensures 5-digit format
  left_join(ssic_final, by = c("primary_ssic_code" = "level_5"))

# Check the result
str(biz_with_hierarchy)
head(biz_with_hierarchy)
```

```{r}
# Check for any unmatched records
cat("Records in biz_selected:", nrow(biz_selected), "\n")
cat("Records after join:", nrow(biz_with_hierarchy), "\n")
cat("Records with no SSIC match:", sum(is.na(biz_with_hierarchy$level_5_description)), "\n")

# View some examples of the joined data
biz_with_hierarchy %>%
  select(primary_ssic_code, level_2, level_2_description, 
         level_3, level_3_description, level_4, level_4_description, 
         level_5_description) %>%
  head(10)
```
### Import Geospatial data into R

The Master Plan 2019 Boundary (No Sea) (KML) was downloaded from data.gov.sg that consists of the geographical boundary of Singapore at the planning subzone level. The data is based on URA Master Plan 2019.
We will use this to plot the boundaries of the subzones.

:::panel-tabset

#### Import data
```{r}
mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
```

#### Tidy data

```{r}
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}
```

```{r}
mpsz <- mpsz %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  dplyr::select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())

mpsz_cl <- mpsz %>%
  filter(SUBZONE_N != "SOUTHERN GROUP",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS")
```

```{r}
mpsz
```

:::

### Geocoding the data

```{r}
# Initiate 2 empty data frames
found <- data.frame() # to store Postal codes that were successfully geocoded with their coordinates
not_found <- data.frame(postcode = character()) # to store postal codes that couldn't be matched
```

```{r}
##| eval: false

postcodes <- unique(biz_selected$postal_code)

url <- "https://onemap.gov.sg/api/common/elastic/search"

for (pc in postcodes) {
  query <- list(
    searchVal = pc,
    returnGeom = "Y",
    getAddrDetails = "Y",
    pageNum = "1"
  )
  
  res <- GET(url, query = query)
  json <- content(res)
  
  if (json$found != 0) {
    df <- as.data.frame(json$results, stringsAsFactors = FALSE)
    df$input_postcode <- pc
    found <- bind_rows(found, df)
  } else {
    not_found <- bind_rows(not_found, data.frame(postcode = pc))
  }
}
```

```{r}
##| eval: false

# Tidying the geocoded data
found <- found %>%
  select(1:10)
```

### Appending the location information

```{r}
##| eval: false
biz_selected_geocoded = biz_with_hierarchy %>%
  left_join(found, 
            by = c('postal_code' = 'POSTAL'))
```

```{r}
##| eval: false
write_rds(biz_selected_geocoded, "data/rds/biz_selected_geocoded.rds")
```

```{r}
#| echo: false
biz_selected_geocoded <- read_rds("data/rds/biz_selected_geocoded.rds")
```

### Create separate datasets for each business type

```{r}
#| eval: false
# Create separate datasets for each business type to count each type
biz_restaurant_cafe <- biz_with_hierarchy %>% filter(str_detect(primary_ssic_code, "^561"))
biz_retail_nonspec <- biz_with_hierarchy %>% filter(str_detect(primary_ssic_code, "^471"))    
biz_real_estate <- biz_with_hierarchy %>% filter(str_detect(primary_ssic_code, "^681"))
```

```{r}
#| eval: false
#| echo: false
write_rds(biz_restaurant_cafe, "data/rds/biz_restaurant_cafe.rds")
write_rds(biz_retail_nonspec, "data/rds/biz_retail_nonspec.rds")
write_rds(biz_real_estate, "data/rds/biz_real_estate.rds")
```

```{r}
#| echo: false
biz_restaurant_cafe <- read_rds("data/rds/biz_restaurant_cafe.rds")
biz_retail_nonspec <- read_rds("data/rds/biz_retail_nonspec.rds")
biz_real_estate <- read_rds("data/rds/biz_real_estate.rds")
```

```{r}
# Check the counts for each category
cat("Total selected businesses:", nrow(biz_selected), "\n")
cat("Restaurants, Cafes, Food Courts, Food Kiosks And Bars (561xx):", nrow(biz_restaurant_cafe), "\n")
cat("Retail Sale in Non-Specialised Stores (471xx):", nrow(biz_retail_nonspec), "\n") 
cat("Retail Sale of Food, Beverage, and Tobacco in Specialised Store (681xx):", nrow(biz_real_estate), "\n")
```

### Converting into simple point features - SF data frame
Before applying the `st_as_sf` function to convert the data into simple point features, we check for missing values first and remove such records.

```{r}
biz_with_coords <- biz_selected_geocoded %>%
  # Remove records without coordinates
  filter(!is.na(X) & !is.na(Y)) %>%
  # Convert LATITUDE and LONGITUDE to numeric if they're character
  mutate(
    LATITUDE = as.numeric(X),
    LONGITUDE = as.numeric(Y)
  )
```

```{r}
biz_sf <- st_as_sf(biz_with_coords, 
                   coords = c("X","Y"), 
                   crs=3414)
```

```{r}
#| eval: false
# Create separate datasets for each business type for later individual analysis
biz_restaurant_cafe_sf <- biz_sf %>% filter(str_detect(primary_ssic_code, "^561"))
biz_retail_nonspec_sf <- biz_sf %>% filter(str_detect(primary_ssic_code, "^471"))    
biz_real_estate_sf <- biz_sf %>% filter(str_detect(primary_ssic_code, "^681"))
```

```{r}
# Save point features as GeoPackage for future use
st_write(biz_sf, "data/sf/biz_sf.gpkg", delete_dsn = TRUE)
st_write(biz_restaurant_cafe_sf, "data/sf/biz_restaurant_cafe_sf.gpkg", delete_dsn = TRUE)
st_write(biz_retail_nonspec_sf, "data/sf/biz_retail_nonspec_sf.gpkg", delete_dsn = TRUE)
st_write(biz_real_estate_sf, "data/sf/biz_real_estate_sf.gpkg", delete_dsn = TRUE)
```

### Verify the point features

::: panel-tabset

#### All selected types

```{r}
summary(biz_sf)
plot(st_geometry(biz_sf), main = "Business Point Features", pch = 16, cex = 0.5)
```

#### 561xx

```{r}
summary(biz_sf)
plot(st_geometry(biz_restaurant_cafe_sf), main = "Business Point Features", pch = 16, cex = 0.5)
```

#### 471xx

```{r}
summary(biz_sf)
plot(st_geometry(biz_retail_nonspec_sf), main = "Business Point Features", pch = 16, cex = 0.5)
```

#### 681xx

```{r}
summary(biz_sf)
plot(st_geometry(biz_real_estate_sf), main = "Business Point Features", pch = 16, cex = 0.5)
```

:::


## Exploratory Data Analysis

### Plotting business points on Singapore boundary

::: panel-tabset

#### All selected types
```{r}
plot(st_geometry(mpsz))
plot(st_geometry(biz_sf), add = TRUE, col = "purple", pch = 16, cex = 0.3)
title("Selected Business Locations in Singapore")
```

#### 561xx
```{r}
plot(st_geometry(mpsz))
plot(st_geometry(biz_restaurant_cafe_sf), add = TRUE, col = "blue", pch = 16, cex = 0.3)
title("Selected Business Locations in Singapore")
```

#### 471xx
```{r}
plot(st_geometry(mpsz))
plot(st_geometry(biz_retail_nonspec_sf), add = TRUE, col = "green", pch = 16, cex = 0.3)
title("Selected Business Locations in Singapore")
```

#### 681xx
```{r}
plot(st_geometry(mpsz))
plot(st_geometry(biz_real_estate_sf), add = TRUE, col = "orange", pch = 16, cex = 0.3)
title("Selected Business Locations in Singapore")
```
:::

### Visualising the distribution

::: panel-tabset

#### 561xx

```{r}
ggplot(data = biz_restaurant_cafe_sf,
       aes(x = reorder(str_wrap(level_5_description, width = 30), 
                      level_5_description, length))) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), 
            hjust = -0.1, size = 2.4) +  # Add this line
  coord_flip() +
  theme_minimal() +
  labs(title = "Count of Businesses by Business Type Description",
       x = "Business Type Description", 
       y = "Number of Businesses")
```

#### 471xx

```{r}
ggplot(data = biz_retail_nonspec,
       aes(x = reorder(str_wrap(level_5_description, width = 30), 
                      level_5_description, length))) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), 
            hjust = -0.1, size = 3) +  # Add this line
  coord_flip() +
  theme_minimal() +
  labs(title = "Count of Businesses by Business Type Description",
       x = "Business Type Description", 
       y = "Number of Businesses")
```

#### 681xx

```{r}
ggplot(data = biz_real_estate,
       aes(x = reorder(str_wrap(level_5_description, width = 30), 
                      level_5_description, length))) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), 
            hjust = -0.1, size = 3) +  # Add this line
  coord_flip() +
  theme_minimal() +
  labs(title = "Count of Businesses by Business Type Description",
       x = "Business Type Description", 
       y = "Number of Businesses")
```
:::

### Year-over-Year Comparison

::: panel-tabset

#### 561xx

```{r}
# Compare business formation rates
biz_2024_561 <- biz_restaurant_cafe %>% filter(YEAR == 2024)
biz_2025_561 <- biz_restaurant_cafe %>% filter(YEAR == 2025)

# Monthly comparison between years
monthly_comparison <- biz_restaurant_cafe %>%
  filter(MONTH_NUM <= 6) %>%  # Jan-Jun only
  group_by(YEAR, MONTH_NUM, primary_ssic_code) %>%
  summarise(count = n()) %>%
  pivot_wider(names_from = YEAR, values_from = count, names_prefix = "year_")

# Calculate growth rates
monthly_comparison$growth_rate <- (monthly_comparison$year_2025 - monthly_comparison$year_2024) / monthly_comparison$year_2024 * 100

monthly_comparison
```

#### 471xx

#### 681xx

:::

## First-order Analysis

### Spatial KDE

**Create Owin object**

```{r}
sg_owin <- as.owin(mpsz_cl)
```

::: panel-tabset

#### 561xx

**Convert sf data frames to ppp class**

```{r}
biz_restaurant_cafe_ppp <- as.ppp(biz_restaurant_cafe_sf)
summary(biz_restaurant_cafe_ppp)
```

**Combine point events object and owin object**

```{r}
biz_restaurant_cafe_ppp = biz_restaurant_cafe_ppp[sg_owin]
biz_restaurant_cafe_ppp
```

#### 471xx

**Convert sf data frames to ppp class**

```{r}
biz_retail_nonspec_ppp <- as.ppp(biz_retail_nonspec_sf)
summary(biz_retail_nonspec_ppp)
```

**Combine point events object and owin object**

```{r}
biz_retail_nonspec_ppp = biz_retail_nonspec_ppp[sg_owin]
biz_retail_nonspec_ppp
```

#### 681xx

```{r}
biz_real_estate_ppp <- as.ppp(biz_real_estate_sf)
summary(biz_real_estate_ppp)
```

**Combine point events object and owin object**

```{r}
biz_real_estate_ppp = biz_real_estate_ppp[sg_owin]
biz_real_estate_ppp
```

:::

### Clark-Evans test
In this section, we will perform Clark-Evans tests for the 3 business entities separately, to quantify whether a point pattern is clustered, random, or uniformly spaced. The result will help us understand the underlying spatial patterns of each business type. This will inform our interpretation of the subsequent KDE analysis.

:::panel-tabset

#### 561xx

The test hypotheses are:
Ho = The distribution of 561xx are randomly distributed.
H1 = The distribution of 561xx are not randomly distributed.
The 95% confident interval will be used.

```{r}
clarkevans.test(biz_restaurant_cafe_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                method="MonteCarlo",
                nsim=999)
```
The Clark-Evans test reveals that restaurants and cafes (SSIC 561xx) in the study area exhibit a significantly clustered spatial distribution (R = 0.35908, p < 0.05). This strong clustering pattern suggests that food service establishments tend to co-locate rather than distribute randomly across the landscape. The clustering likely reflects agglomeration benefits where restaurants benefit from proximity to competitors through shared customer traffic, as well as urban planning policies that concentrate commercial activities in designated zones. This spatial concentration creates identifiable dining districts and food hubs within the study area, which has important implications for business location strategy and urban commercial planning.

#### 471xx

The test hypotheses are:
Ho = The distribution of 471xx are randomly distributed.
H1 = The distribution of 471xx are not randomly distributed.
The 95% confident interval will be used.

```{r}
clarkevans.test(biz_retail_nonspec_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                method="MonteCarlo",
                nsim=999)
```
The Clark-Evans test indicates that non-specialised retail stores (SSIC 471xx) exhibit a significantly clustered spatial distribution (R = 0.46, p < 0.001). While still showing clustering behavior, these retail establishments are somewhat more dispersed than restaurants and cafes, reflecting their need to serve broader geographic markets while still benefiting from commercial agglomeration in shopping districts and retail centers.

#### 681xx

The test hypotheses are:
Ho = The distribution of 681xx are randomly distributed.
H1 = The distribution of 681xx are not randomly distributed.
The 95% confident interval will be used.

```{r}
clarkevans.test(biz_real_estate_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                method="MonteCarlo",
                nsim=999)
```
The Clark-Evans test shows that real estate businesses (SSIC 681xx) exhibit a significantly clustered spatial distribution (R = 0.45, p < 0.001). This clustering pattern suggests that real estate agencies tend to co-locate in commercial districts and business centers, likely benefiting from proximity to financial services, legal offices, and high foot traffic areas where property transactions commonly occur.

:::

### Compute KDE with adaptive bandwidth

Given the significant clustering patterns identified through Clark-Evans testing (R-values of 0.36-0.46, all p < 0.001), adaptive bandwidth was selected for the kernel density estimation analysis. The clustered nature of business locations creates highly variable point densities across the study area, with dense concentrations in commercial districts and sparse distributions in other zones. Adaptive bandwidth addresses this heterogeneity by applying narrower bandwidths in high-density clusters to preserve important spatial details, while using wider bandwidths in sparse areas to provide appropriate smoothing. This approach ensures more accurate representation of business density patterns compared to fixed bandwidth, which would over-smooth dense business districts and potentially create artifacts in sparse areas.

:::panel-tabset

#### 561xx

```{r}
par(bg = "#EEEEEE")
biz_restaurant_cafe_ppp_km <- rescale.ppp(biz_restaurant_cafe_ppp, 1000, "km")
kde_biz_restaurant_cafe_ab <- adaptive.density(
  biz_restaurant_cafe_ppp_km, method = "kernel"
)
plot(kde_biz_restaurant_cafe_ab)
```

#### 471xx

```{r}
par(bg = "#EEEEEE")
biz_retail_nonspec_ppp_km <- rescale.ppp(biz_retail_nonspec_ppp, 1000, "km")
kde_biz_retail_nonspec_ab <- adaptive.density(
  biz_retail_nonspec_ppp_km, method = "kernel"
)
plot(kde_biz_retail_nonspec_ab)
```

#### 681xx 

```{r}
par(bg = "#EEEEEE")
biz_real_estate_ppp_km <- rescale.ppp(biz_real_estate_ppp, 1000, "km")
kde_biz_real_estate_ab <- adaptive.density(
  biz_real_estate_ppp_km, method = "kernel"
)
plot(kde_biz_real_estate_ab)
```

:::

### Spatio-Temporal KDE
Jaundry, Beauty, and Dental business types have 150, 1762 and 81 points respectively during Jan 2024 - June 2025. In this study, we will slice the data compute quarterly spatio-temporal KDE for analysis to see seasonal patterns. 

#### Visualise business entity points by month/quarter

:::panel-tabset

##### 561xx

```{r}
#| fig-width: 12
#| fig-height: 12

biz_restaurant_cafe_quarter <- biz_restaurant_cafe_sf %>%
  mutate(
    quarter = case_when(
      MONTH_NUM %in% c(1, 2, 3) ~ "Q1",
      MONTH_NUM %in% c(4, 5, 6) ~ "Q2",
      MONTH_NUM %in% c(7, 8, 9) ~ "Q3",
      MONTH_NUM %in% c(10, 11, 12) ~ "Q4"
    ),
    year_quarter = paste(YEAR, quarter, sep = "-")
  )

tmap_options(bg.color = "#EEEEEE")
tm_shape(mpsz_cl) + 
  tm_polygons(fill = "#FFF2F2", col = "grey80") +
tm_shape(biz_restaurant_cafe_quarter) +
  tm_dots(fill = "#476EAE", size = 0.4) +
  tm_facets(by = "year_quarter", free.coords = FALSE, drop.units = TRUE,
            ncol = 2) +
tm_layout(
    main.title = "561xx- Quarterly Density",
    bg.color = "#EED3D9")
```

##### 471xx 

```{r}
#| fig-width: 12
#| fig-height: 32

biz_retail_nonspec_my <- biz_retail_nonspec_sf %>%
  mutate(
    date = as.Date(paste(YEAR, MONTH_NUM, "01", sep = "-")),
    month_year = format(date, "%Y-%m"),
    month_year_label = format(date, "%b %Y")
  )

tmap_options(bg.color = "#EEEEEE")
tm_shape(mpsz_cl) + 
  tm_polygons(fill = "#FFF2F2", col = "grey80") +
  tm_shape(biz_retail_nonspec_my) +
  tm_dots(fill = "#BA487F", size = 0.2) +
  tm_facets(by = "month_year", free.coords = FALSE, drop.units = TRUE,
            ncol = 2) +
tm_layout(
    main.title = "471xx - Monthly Density",
    bg.color = "#CADABF")
```

##### 681xx 

```{r}
#| fig-width: 12
#| fig-height: 12

biz_real_estate_quarter <- biz_real_estate_sf %>%
  mutate(
    quarter = case_when(
      MONTH_NUM %in% c(1, 2, 3) ~ "Q1",
      MONTH_NUM %in% c(4, 5, 6) ~ "Q2",
      MONTH_NUM %in% c(7, 8, 9) ~ "Q3",
      MONTH_NUM %in% c(10, 11, 12) ~ "Q4"
    ),
    year_quarter = paste(YEAR, quarter, sep = "-")
  )

tmap_options(bg.color = "#EEEEEE")
tm_shape(mpsz_cl) + 
  tm_polygons(fill = "#FFF2F2", col = "grey80") +
  tm_shape(biz_real_estate_quarter) +
  tm_dots(fill = "#EF7722", size = 0.4) +
  tm_facets(by = "year_quarter", free.coords = FALSE, drop.units = TRUE,
            ncol = 2) +
tm_layout(
    main.title = "681xx - Quarterly Density",
    bg.color = "#B5C0D0")
```

:::

#### Compute Spatio-Temporal STKDE by month/quarter

:::panel-tabset

##### 561xx

```{r}
# extract laundry data by quarter
kde_biz_restaurant_cafe_quarter <- biz_restaurant_cafe_quarter %>%
  dplyr::select(year_quarter)

# create ppp
biz_restaurant_cafe_quarter_ppp <- as.ppp(kde_biz_restaurant_cafe_quarter)
  

# include owin object
biz_restaurant_cafe_quarter_owin <- biz_restaurant_cafe_quarter_ppp[sg_owin]

```

```{r}
par(bg = "#EEEEEE")
plot(biz_restaurant_cafe_quarter_owin)
```

**Compute STKDE**

```{r}
#convert year_quarter to numeric
marks(biz_restaurant_cafe_quarter_owin) <- case_when(
  marks(biz_restaurant_cafe_quarter_owin) == "2024-Q1" ~ 1,
  marks(biz_restaurant_cafe_quarter_owin) == "2024-Q2" ~ 2,
  marks(biz_restaurant_cafe_quarter_owin) == "2024-Q3" ~ 3,
  marks(biz_restaurant_cafe_quarter_owin) == "2024-Q4" ~ 4,
  marks(biz_restaurant_cafe_quarter_owin) == "2025-Q1" ~ 5,
  marks(biz_restaurant_cafe_quarter_owin) == "2025-Q2" ~ 6
)

st_kde_9601 <- spattemp.density(biz_restaurant_cafe_quarter_owin)
summary(st_kde_9601)
```

**Plot STKDE object**

```{r}
#| fig-width: 12
#| fig-height: 4
tims <- c(1,2,3,4,5,6)
par(mfcol = c(1,3), bg = "#EEEEEE")
for(i in tims){
  plot(st_kde_9601, i,
       override.par = FALSE,
       fix.range = TRUE, main = paste("561xxKDE at Quarter", i))
}

```

##### 471xx 

```{r}
# extract laundry data by quarter
kde_biz_retail_nonspec_month <- biz_retail_nonspec_my %>%
  dplyr::select(month_year)

# create ppp
biz_retail_nonspec_month_ppp <- as.ppp(kde_biz_retail_nonspec_month)
  

# include owin object
biz_retail_nonspec_month_owin <- biz_retail_nonspec_month_ppp[sg_owin]

```

```{r}
par(bg = "#EEEEEE")
plot(biz_retail_nonspec_month_owin)
```

**Compute STKDE**

```{r}
#convert year_quarter to numeric
marks(biz_retail_nonspec_month_owin) <- case_when(
  marks(biz_retail_nonspec_month_owin) == "2024-01" ~ 1,
  marks(biz_retail_nonspec_month_owin) == "2024-02" ~ 2,
  marks(biz_retail_nonspec_month_owin) == "2024-03" ~ 3,
  marks(biz_retail_nonspec_month_owin) == "2024-04" ~ 4,
  marks(biz_retail_nonspec_month_owin) == "2024-05" ~ 5,
  marks(biz_retail_nonspec_month_owin) == "2024-06" ~ 6,
  marks(biz_retail_nonspec_month_owin) == "2024-07" ~ 7,
  marks(biz_retail_nonspec_month_owin) == "2024-08" ~ 8,
  marks(biz_retail_nonspec_month_owin) == "2024-09" ~ 9,
  marks(biz_retail_nonspec_month_owin) == "2024-10" ~ 10,
  marks(biz_retail_nonspec_month_owin) == "2024-11" ~ 11,
  marks(biz_retail_nonspec_month_owin) == "2024-12" ~ 12,  
  marks(biz_retail_nonspec_month_owin) == "2025-01" ~ 13,
  marks(biz_retail_nonspec_month_owin) == "2025-02" ~ 14,
  marks(biz_retail_nonspec_month_owin) == "2025-03" ~ 15,
  marks(biz_retail_nonspec_month_owin) == "2025-04" ~ 16,
  marks(biz_retail_nonspec_month_owin) == "2025-05" ~ 17,
  marks(biz_retail_nonspec_month_owin) == "2025-06" ~ 18
  )

st_kde_9602 <- spattemp.density(biz_retail_nonspec_month_owin)
summary(st_kde_9602)
```

**Plot STKDE object**

```{r}
#| fig-width: 12
#| fig-height: 4

tims <- c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18)
par(mfcol = c(1,2), bg = "#EEEEEE")
for(i in tims){
  plot(st_kde_9602, i,
       override.par = FALSE,
       fix.range = TRUE, main = paste("471xx KDE at Month", i))
}

```


##### 681xx 

```{r}
# extract laundry data by quarter
kde_biz_real_estate_quarter <- biz_real_estate_quarter %>%
  dplyr::select(year_quarter)

# create ppp
biz_real_estate_quarter_ppp <- as.ppp(kde_biz_real_estate_quarter)
  

# include owin object
biz_real_estate_quarter_owin <- biz_real_estate_quarter_ppp[sg_owin]

```

```{r}
par(bg = "#EEEEEE")
plot(biz_real_estate_quarter_owin)
```

**Compute STKDE**

```{r}
#convert year_quarter to numeric
marks(biz_real_estate_quarter_owin) <- case_when(
  marks(biz_real_estate_quarter_owin) == "2024-Q1" ~ 1,
  marks(biz_real_estate_quarter_owin) == "2024-Q2" ~ 2,
  marks(biz_real_estate_quarter_owin) == "2024-Q3" ~ 3,
  marks(biz_real_estate_quarter_owin) == "2024-Q4" ~ 4,
  marks(biz_real_estate_quarter_owin) == "2025-Q1" ~ 5,
  marks(biz_real_estate_quarter_owin) == "2025-Q2" ~ 6
)

st_kde_86204 <- spattemp.density(biz_real_estate_quarter_owin)
summary(st_kde_86204)
```

**Plot STKDE object**

```{r}
#| fig-width: 12
#| fig-height: 4
tims <- c(1,2,3,4,5,6)
par(mfcol = c(1,3), bg = "#EEEEEE")
for(i in tims){
  plot(st_kde_86204, i,
       override.par = FALSE,
       fix.range = TRUE, main = paste("681xx KDE at Quarter", i))
}

```

:::

## Second order spatial and spatio-temporal analysis on the selected business types

## Conclusions and Future Improvements
