---
title: "Take-home Exercise 1 | Geospatial Analytics for Public Good"
subtitle: "In this take-home exercise, we are interested to investigate the spatial and spatio-temporal patterns of new businesses and firms established in Singapore for the first six months of 2025."
date: "Sep 13, 2025"
date-modified: "last-modified"
author: Linh Nguyen
execute:
  echo: true
  eval: true
  freeze: auto
  message: false
  warning: false
format:
  html:
    code-link: true
    toc: true
number-sections: true
number-offset: 1
editor: visual
---

# Overview

## Setting The Scene

Where businesses choose to locate plays a huge role in how Singapore continues to grow and develop as a city-state. Smart locations give companies advantages like easier access to skilled workers and excellent infrastructure, while also shaping different districts like the Central Business District, Jurong Innovation District, or Punggol Digital District, and making it easier for goods to move through our efficient transport networks and for employees to commute via MRT and buses.

Studying how businesses spread out across Singapore and change over time is crucial for urban planning by agencies like URA and JTC. This research helps planners understand how economic activity and our urban landscape develop together, enabling them to design better integrated developments, plan for future growth in areas like the Greater Southern Waterfront, use our limited land resources efficiently, and build infrastructure that supports both business competitiveness and residents' quality of life in our compact island nation.

## Objectives

-   To identify where and when new businesses are established in Singapore, revealing clusters, patterns, and trends that help inform urban planning, resource allocation, and economic development decisions.

-   To uncover complex relationships in how new businesses are distributed across the island - such as how related industries cluster together, how past business locations influence future ones, and how commercial districts develop structurally over time. These insights are essential for evidence-based urban planning and policy-making.

By understanding business locations and how their distribution evolves, Singapore's urban planners and agencies can gain valuable spatial insights, optimize city operations, and develop effective strategies to attract investment and promote sustainable growth across our city-state.

## The Data

| Dataset | Description | Source | Format |
|------------------|------------------|------------------|------------------|
| **ACRA Information on Corporate Entities** | Information of registered entities in Singapore, including location data | [data.gov.sg](https://data.gov.sg/datasets?query=ACRA+Information+on+Corporate+Entities&page=1&resultId=2) | CSV |
| **SSIC: Singapore Standard Industrial Classification 2020** | List of codes represented economic activity of business establishments in Singapore, 2020 version. | [SSIC: Singapore Standard Industrial Classification](https://classification.codes/classifications/industry/ssic-singapore) | Excel |

# Set up R environment

## Import R Packages

```{r}
pacman::p_load(tidyverse, sf, tmap, httr)
```

## Set Seed

```{r}
set.seed(1234)
```

# Spatial Data Wrangling

## Data Import & Preparation

Data Import 

```{r}
#| eval: false
folder_path <- "data/aspatial/ACRA"
file_list <- list.files(path = folder_path, 
                        pattern = "^ACRA*.*\\.csv$", 
                        full.names = TRUE)

acra_data <- file_list %>%
  map_dfr(read_csv)
```

```{r}
#| echo: false
#| eval: false
write_rds(acra_data, "data/rds/acra_data.rds")
```

```{r}
#| echo: false
acra_data <- read_rds("data/rds/acra_data.rds")
```

## Data Wrangling

### Extracting data of chosen business types

The chosen business types are:
- SSIC 56: Food and Beverage Service Activities
- SSIC 47: Retail Trade  
- SSIC 68: Real Estate Activities

```{r}
#| echo: false
#| eval: false
biz_selected <- acra_data %>%
  select(1:24) %>%
  filter(str_detect(primary_ssic_code, "^(56|47|68)")) %>%
  rename(date = registration_incorporation_date) %>%
  mutate(date = as.Date(date),
         YEAR = year(date),
         MONTH_NUM = month(date),
         MONTH_ABBR = month(date, 
                            label = TRUE, 
                            abbr = TRUE)) %>% 
  mutate(
    postal_code = str_pad(postal_code, 
    width = 6, side = "left", pad = "0")) %>%
  filter(YEAR >= 2024 & YEAR <= 2025) %>%  # Updated date range as per assignment
  filter(date >= as.Date("2024-01-01") & date <= as.Date("2025-06-30"))  # Exact date range
```


```{r}
#| eval: false
# Create separate datasets for each business type for individual analysis
biz_restaurants <- biz_selected %>% filter(str_detect(primary_ssic_code, "^56"))
biz_retail_nonspec <- biz_selected %>% filter(str_detect(primary_ssic_code, "^47"))    
biz_real_estate_own <- biz_selected %>% filter(str_detect(primary_ssic_code, "^68"))
```

```{r}
#| eval: false
#| echo: false
# Write and save so that we don't have to run the above part next time, and reduce the render time
# write_rds(biz_selected, "data/rds/biz_selected.rds")
write_rds(biz_food_beverage, "data/rds/biz_food_beverage.rds")
write_rds(biz_retail, "data/rds/biz_retail.rds")
write_rds(biz_real_estate, "data/rds/biz_real_estate.rds")
```

```{r}
# Read the files you just wrote
#| echo: false
# biz_selected <- read_rds("data/rds/biz_selected.rds")
biz_food_beverage <- read_rds("data/rds/biz_food_beverage.rds")
biz_retail <- read_rds("data/rds/biz_retail.rds")
biz_real_estate <- read_rds("data/rds/biz_real_estate.rds")
```

```{r}
# Check the counts for each category
cat("Food & Beverage (56xxx):", nrow(biz_food_beverage), "\n")
cat("Retail Trade (47xxx):", nrow(biz_retail), "\n") 
cat("Real Estate (68xxx):", nrow(biz_real_estate), "\n")
```

### Geocoding

```{r}
#| eval: false
postcodes <- unique(biz_real_estate$postal_code)

url <- "https://onemap.gov.sg/api/common/elastic/search"

found <- data.frame()
not_found <- data.frame(postcode = character())

for (pc in postcodes) {
  query <- list(
    searchVal = pc,
    returnGeom = "Y",
    getAddrDetails = "Y",
    pageNum = "1"
  )
  
  res <- GET(url, query = query)
  json <- content(res)
  
  if (json$found != 0) {
    df <- as.data.frame(json$results, stringsAsFactors = FALSE)
    df$input_postcode <- pc
    found <- bind_rows(found, df)
  } else {
    not_found <- bind_rows(not_found, data.frame(postcode = pc))
  }
}
```

### Tidying the geocoded data

```{r}
#| eval: false
found <- found %>%
  select(1:10)
```

### Appending the location information

```{r}
#| eval: false
biz_real_estate = biz_real_estate %>%
  left_join(found, 
            by = c('postal_code' = 'POSTAL'))
```

```{r}
#| eval: false
write_rds(biz_real_estate, "data/rds/biz_real_estate.rds")
```

```{r}
biz_real_estate <- read_rds("data/rds/biz_real_estate.rds")
```

### Converting into SF data frame

```{r}
biz_real_estate_sf <- st_as_sf(biz_real_estate, 
                         coords = c("X","Y"), 
                         crs=3414)
```

# Exploratory Data Analysis

## Visualising the distribution

```{r}
ggplot(data = biz_real_estate,
       aes(x = MONTH_ABBR)) +
  geom_bar()
```

## Visualising the business

```{r}
tmap_mode('view')
tm_shape(biz_real_estate_sf)+
  tm_dots() +
  tm_view(set_zoom_limits = c(10, 14))
```

# Spatial and spatio-temporal KDE of business types

# Second order spatial and spatio-temporal analysis on the selected business types

# Conclusions and Future Improvements
